Identification and characterization of early human photoreceptor states and cell-state-specific retinoblastoma-related features

Version 1

Abstract

Abstract

Human cone photoreceptors differ from rods and serve as the retinoblastoma cell-of-origin. Here, we used deep full-length single-cell RNA-sequencing to distinguish post-mitotic cone and rod developmental states and cone-specific features that contribute to retinoblastomagenesis. The analyses revealed early post-mitotic cone- and rod-directed populations characterized by higher THRB or NRL regulon activities, an immature photoreceptor precursor population with concurrent cone and rod gene and regulon expression, and distinct early and late cone and rod maturation states distinguished by maturation-associated declines in RAX regulon activity. Unexpectedly, both L/M cone and rod precursors co-expressed NRL and THRB RNAs, yet they differentially expressed functionally antagonistic NRL isoforms and prematurely terminated THRB transcripts. Early L/M cone precursors exhibited successive expression of lncRNAs along with MYCN, which composed the seventh most L/M-cone-specific regulon, and SYK, which contributed to the early cone precursors’ proliferative response to RB1 loss. These findings reveal previously unrecognized photoreceptor precursor states and a role for early cone-precursor-intrinsic SYK expression in retinoblastoma initiation.

Introduction

Vertebrate photoreceptors develop from optic vesicle retinal progenitor cells (RPCs) through progressive RPC lineage restriction, fate determination, and post-mitotic maturation (1, 2). While several transcription factors that govern these events have been identified, important aspects of photoreceptor fate determination and maturation remain unclear. For example, it is unclear if fate is determined in RPCs, where OTX2 and ONECUT1 are thought to control the post-mitotic expression of long- or medium-wavelength (L/M) cone determinant TRβ2 and rod determinant NRL (3), or is determined in post-mitotic photoreceptor precursors with concurrent TRβ2 and NRL expression (4). Following fate commitment, post-mitotic developmental stages have been defined based on morphologic features and phototransduction-related gene or protein expression (5, 6), but it is unclear if progression through such stages is subdivided into distinct cell states governed by unique transcription factor combinations or represents a developmental continuum.

An improved understanding of photoreceptor development may provide insight into the pathogenesis of retinal dystrophies, retinal degenerations, and the retinal cone precursor cancer, retinoblastoma (7, 8). In the latter case, L/M cone precursors lacking the retinoblastoma protein (pRB) were shown to proliferate in a manner dependent on the L/M-cone lineage factors RXRγ and TRβ2 and the intrinsically highly expressed MDM2 and MYCN oncoproteins, likely representing the first step of retinoblastoma tumorigenesis (8–11). Similarly, retinoblastoma cell proliferation depends on RXRγ, TRβ2, MDM2, and MYCN (9), implying that intrinsic L/M-cone factors contribute to the oncogenic state. However, retinoblastoma cells also express rod lineage factor NRL RNAs, which – along with other evidence – suggested a heretofore unexplained connection between rod gene expression and retinoblastoma development (12, 13). Improved discrimination of early photoreceptor states is needed to determine if co-expression of rod- and cone-related genes is adopted during tumorigenesis or reflects the co-expression of such genes in the retinoblastoma cell of origin.

The cone precursors’ propensity to form retinoblastoma is a human-specific feature whose study requires analysis of developing human retina (8). Single-cell RNA-sequencing (scRNA-seq) is well suited to such analyses as it enables discrimination of cell type-specific fate determination and maturation features. scRNA-seq studies employing 3’ end-counting have defined age-related post-mitotic transition populations, fate-determining features of post-mitotic photoreceptor precursors (14–16), and gene expression changes associated with the cone fate decision and early development (14–19). However, 3’ end-counting cannot be used to interrogate transcript isoforms, and the relatively low number of genes detected per cell limits the ability to distinguish closely related states in individual cells.

In this study, we sought to further define the transcriptomic underpinnings of human photoreceptor development and their relationship to retinoblastoma development. We generated deep, full-length single-cell transcriptomes of human retinal progenitor cells (RPCs) and developing photoreceptors from fetal week 13-19 retinae, with enrichment of rare cone precursor populations, and applied long-read cDNA sequencing, RNA velocity, pseudotemporal trajectory reconstruction, and single-cell regulatory network inference and clustering (SCENIC) to interrogate individual cell states. These analyses discriminated previously unrecognized photoreceptor developmental states, identified photoreceptor precursor states with cone and rod-related RNA co-expression, uncovered cell type-specific expression of RNA isoforms of photoreceptor fate-determining genes, elucidated post-mitotic photoreceptor developmental trajectories, and defined retinoblastoma cell-of-origin features that contribute to retinoblastoma genesis.

Results

Regulon-defined RPC and photoreceptor precursor states

To interrogate transcriptomic changes during human photoreceptor development, dissociated RPCs and photoreceptor precursors were FACS-enriched from 18 fetal week 13–19 retinae and isolated using microfluidics or direct sorting into microliter droplets, followed by full-length cDNA synthesis, paired-end sequencing, and alignment to Ensembl transcript isoforms (Figure 1A).

Photoreceptor-enriched full-length scRNA-seq of developing human retina.

A. Overview of sample collection and sequencing. B. UMAP plot and low-resolution cell type clusters. C. Expression of marker genes for RPC/MGs (LHX2), rods (NR2E3), S cones (OPN1SW), L/M cones (THRB). Insets: Gene expression violin plots (from left to right): RPC/MG (red), iPRP (brown), LM cone (green), S cone (teal), early rod (blue), late rod (pink). D. Expression of markers of rod maturation (PDE6G, RHO) and cone maturation (PDE6H, OPN1LW). Arrowheads: Late maturing RHO+ rods (top), late maturing OPN1LW+ cones (bottom). See Figure S1 for additional examples. UMAP and violin plots for any gene or Ensembl transcript isoform can be produced at https://docker.saban.chla.usc.edu/cobrinik/app/seuratApp/.

The FACS enrichment was based on a prior cone isolation method (10) but with wider gating to include rods and RPCs. Exclusion of all cells with <100,000 reads and 18 cells expressing markers of retinal ganglion, amacrine, and/or horizontal cells (POU4F1, POU4F2, POU4F3, TFAP2A, TFAP2B, ISL1) and lacking photoreceptor lineage marker OTX2 yielded 794 single cells with averages of 3,750,417 uniquely aligned reads, 8,278 genes detected, and 20,343 Ensembl transcripts inferred (Figure S1A-C). Sequencing batches were normalized and transcriptomes clustered and visualized in uniform manifold approximation and projection (UMAP) plots that integrated cells across different retinae, isolation methods, and sequencing runs (Figure 1B, S1D-F).

Low resolution Louvain clustering generated six clusters that segregated into mostly distinct UMAP domains (Figure 1B). One cluster was comprised of RPCs and Mϋller glia (MG), with specific expression of LHX2, VSX2, SOX2, and SLC1A3, while five clusters were comprised of cells with photoreceptor features, with wide expression of OTX2 and CRX and cluster-specific rod- and cone gene expression (Figure 1C, S2). In UMAP space, the cluster designated immature photoreceptor precursors (iPRPs) intermixed with the RPC/MG population, extended towards early LM cone precursors, and was predominantly comprised of cells expressing the L/M cone determinant THRB (Figures 1B,C).

Two clusters expressing the rod determinant NR2E3 were designated early maturing rod (ER) and late maturing rod (LR) (Figure 1B), based on the latter’s increased expression of rod phototransduction genes GNAT1, CNGB1, PDE6G, GNGT1 and RHO (Figure 1D, S2C).

Differential expression analysis revealed other genes upregulated in late rods (GNB1, SAMD7, NT5E) (20, 21) as well as the downregulated CRABP2, DCT, and FABP7 (Figure S3A, Table S1). Genes upregulated in the LR cluster were enriched for photoreceptor and light sensing gene ontologies (Figure S3B) including spectrin binding, likely relating to proteins that control photoreceptor polarity and synapse formation (22, 23), and purine biosynthesis and ribonucleotide metabolic processes, potentially related to the developing photoreceptors’ high NAD+ requirements (24).

Cones segregated into distinct S- and L/M-cone clusters, with differential expression of cone subtype markers (OPN1SW, THRB), previously identified S-cone enriched genes (CCDC136, UPB1) (25–27), novel S cone enriched genes (MEGF10, NRXN3, ACKR3), and the L/M cone transcription factor ISL2 (15), among others (Figure 1B,C, S3C, Table S2). Gene ontology analysis did not reveal relevant terms enriched in S cones, whereas L/M cones were enriched for protein translation related ontologies (Figure S3D) due to increased expression of ribosomal protein genes RPL23A, RPLP0, RPS19, RPS27, RPS27A, RPS29, RPS3A (Table S2).

In UMAP space, the THRB+ L/M-cone cluster segregated into a large 363-cell proximal population and a 5-cell distal population inferred to represent early maturing and late maturing stages, respectively, based on the latter’s increased expression of cone phototransduction genes OPN1LW (encoding L-opsin), PDE6H, and GUCA1C (Figure 1D, S3E,F), analogous to RHO, PDE6G, and GNGT1 upregulation in late rods. Compared to the early L/M population, late L/M cones had upregulation of three M-opsin genes, TTR, encoding the retinol-binding protein transthyretin, PCP4, a small protein that binds calmodulin previously noted in foveal cones (28), and MYL4, a myosin light chain gene upregulated in retinal organoid L/M cones (27), among others (Figure S3E,F, Table S3). The low proportions of OPN1MW+ and OPN1LW+ late maturing L/M cones is consistent with a prior analysis of similar-age retinae and with the further upregulation of these proteins in later maturation (15).

To further interrogate cell identities, we used SCENIC to identify cluster-specific transcription factor regulons, which represent the overall expression of single transcription factors and their likely coregulated target genes (29). The highest specificity regulons defined major cell populations including RPC/MG-specific E2F2, E2F3, VSX2, and PAX6; pan-photoreceptor NEUROD1, OTX2, and CRX; rod-specific NRL; and L/M cone-specific THRB and ISL2 (Figure 2, Table S4). SCENIC also distinguished ER and LR states via the latter’s increased NRL, CRX, ATF4, and LHX3 and decreased HMX1 and RAX activities (p <0.0005 for each, Dunn test). RAX activity also decreased in the 5-cell late maturing L/M cone group (Figure 2B), supporting its distinct transcriptomic identity and suggesting a similar mode of late cone and late rod maturation. Additionally, iPRPs expressed their most specific regulons, LHX9 and OLIG2, at levels similar to RPC/MGs along with photoreceptor-related regulons, consistent with the transitional nature of this population (Figure 2A, Table S4). However, SCENIC did not identify S cone-specific regulons, in keeping with the notion that S cones represent a default photoreceptor state induced by pan-photoreceptor factors such as OTX2, CRX, and NEUROD1 in the absence of NRL and THRB (30). Thus, deep, full-length scRNA-seq enabled identification of regulons underlying RPC and developing photoreceptor states at the single cell level.

Regulon-defined RPC and photoreceptor precursor states

A. Ward-clustered heatmap of the highest scoring SCENIC regulons in each cluster, displaying Z-score normalized regulon activities. Late = late maturing L/M cones. B. Box plot of RAX regulon area under the curve (AUC) values for early and late L/M cones and rods. *, p<0.005; ***, p<0.0005, Dunn test. C,D. UMAP plots of regulon AUC values for (C) PAX6 (RPC/MG) and E2F2 (RPC), and (D) OTX2 (photoreceptors and photoreceptor-committed RPCs), NRL (rod) and THRB and ISL2 (L/M cone).

Differential expression of NRL and THRB isoforms in rod and cone precursors

Although cone and rod precursors segregated into distinct clusters, mRNAs encoding rod-determining factor NRL, L/M cone-determining factor TRβ, and cone marker RXRγ were co-expressed in both rod and cone precursor populations, with mean NRL expression only 4.3-fold higher in the ER vs LM cluster, mean THRB expression 5.0-fold higher in LM vs LR, and mean RXRG expression 4.3-fold higher in LM vs ER (Figure S2C,D and below). Cone NRL expression was unexpected given NRL’s role in rod fate determination (27) and rod-specific NRL regulon activity (Figure 2A,D). Similarly, rod RXRG and THRB expression were unexpected given their roles in cone gene expression and fate determination (31, 32) and L/M cone-specific THRB regulon activity (Figure 2A,D). Accordingly, we used full-length scRNA-seq data to determine if cone and rod precursors differentially express NRL, THRB, and RXRG transcript isoforms.

For NRL, the most highly assigned transcript isoform (ENST00000397002) encoded the canonical full-length NRL (FL-NRL) (RefSeq NP_001341697.1), while the second-most assigned isoform (ENST00000560550) encoded a previously uncharacterized transcript that used an alternative “P2” promoter and first exon, here termed exon 1T (Figure 3A-C). The novel transcript was predicted to encode an N-terminally truncated NRL (Tr-NRL) retaining the leucine zipper DNA binding domain but lacking the minimal transactivation domain (33) (Figure 3C).

Differential expression of NRL isoforms in rod and cone precursor populations.

A. Expression of NRL gene and the most highly assigned Ensembl isoforms ENST00000397002 (FL-NRL) and ENST00000560550 (Tr-NRL). B. Mean NRL isoform assignments for clusters defined in Figure 1B, presented as total counts (top) and percentage of total counts (bottom). Significance for LM vs. ER fold change, colored by isoform. ****, p <0.0002; *****, <0.000001 (bootstrapped Welch’s t-test). C. Top: Mean read counts across Ensembl NRL exons for each cluster. Bottom: Transcript structures numbered according to amino acid positions. Minimal transactivation domain (MTD) in green. Arrowheads: Red/black: First exons where red is higher of two peaks. D. Relative difference box plot of raw reads mapping to truncated (Tr) and full length (F) transcript first exons in each cell, according to cluster. Relative difference is the difference in reads mapping to truncated and full-length NRL first exons (Tr-F) divided by the sum of both (Tr+F). Values >0 indicate more reads assigned to truncated isoform, values < 0 indicate more reads assigned to full-length isoform. ***, p <0.0001 (post-hoc Dunn test). E. NRL and RXRγ immunostaining and RNA FISH with probes specific to truncated Tr-NRL exon 1T (green puncta) and FL-NRL exons 1 and 2 (orange puncta) in FW16 retina. Boxed regions enlarged at right show a RXRlo, NRL+ rod with one Tr-NRL and six FL-NRL puncta (top) and an RXRγhi, NRL− cone with one Tr-NRL and no FL-NRL puncta (bottom), indicated with same-color arrows. Scale bar: 10 µm. F. Ratio of fluorescent puncta observed in experiment depicted in (E) for NRL+ or RXRγhi cells where Tr puncta >0. ***; p <0.0005 (Welch’s t-test). G. Top: PDE6B-luciferase reporter activity in NIH-3T3 cells transfected with indicated amounts (ng) of pcDNA4C-EF1α and derived FL-NRL and Tr-NRL. Bottom: Expression constructs. Error bars = standard deviation of triplicate measurements. *, p <0.05; **, <0.005 (Student’s t-test). Data representative of two experiments in NIH-3T3 and one in 293T. H. Long-read nanopore sequencing of pooled 5’ RACE reactions initiated with NRL exon 3 primers and performed on cDNA libraries from 23 ER cells (top) and 21 LM cells (bottom). Each schematic shows total exon coverage (above) and individual transcript structures (below), where expressed sequences are grey and introns light blue. Full-length (FL), alternatively spliced or internally initiated exon 2 (Δex2), and truncated (Tr) transcripts are indicated by brackets. Red arrow: Transcripts resembling DD10, with internal exon 2 transcription initiation and premature splicing to exon 3. Ensembl FL-NRL and Tr-NRL transcript isoforms and RACE primer positions are shown below.

While both isoforms were inferred to be more highly expressed in the ER vs LM cluster, the average FL-NRL (ENST00000397002): Tr-NRL (ENST00000560550) ratio was 2.8:1 in early rods and 0.72:1 in L/M cones (Figure 3B). Consistent with the assigned isoform ratios, mean Tr-NRL exon 1T coverage was higher in S and L/M cones while FL-NRL exon 1 coverage was higher in rods (Figure 3C, red vs. black arrowheads). Comparing the reads mapped to each first exon relative to total reads further confirmed that the Tr-NRL exon 1T predominated in individual cones whereas the FL-NRL exon 1 predominated in rods (Figure 3D). The cone cells’ higher proportional expression of Tr-NRL first exon sequences was validated by RNA fluorescence in situ hybridization (FISH) of FW16 fetal retina (Figure 3E,F).

While the Tr-NRL isoform was not to our knowledge previously described, another NRL isoform that initiated within exon 2 and lacked the NRL transactivation domain due to alternative exon 2 splicing, termed DD10 (Figure 3C), was previously identified in adult retina (34). Concordantly, we detected reads spanning the unique DD10 splice junction, yet at lower levels than the unique Tr-NRL junction (5,942 versus 57,048). As both Tr-NRL and DD10 lack the NRL transactivation domain and as DD10 interferes with FL-NRL transactivation (35), we examined if Tr-NRL similarly opposes FL-NRL transcriptional activity. In luciferase assays, Tr-NRL suppressed FL-NRL activation of a PDE6B promoter (Figure 3G), demonstrating that the cone-enriched Tr-NRL antagonized FL-NRL-mediated transactivation.

To further define NRL isoforms, we performed NRL 5’ rapid amplification of cDNA ends (RACE) on the already generated single cell cDNA libraries from 23 early rod (ER) and 21 early L/M cone (LM) cells, followed by nanopore long-read sequencing of the pooled RACE products. The long-read sequencing revealed isoforms consistent with FL-NRL, Tr-NRL, DD10, and several other NRL isoforms with alternative transcription initiation and alternative splicing within exon 2 as well as within the Tr-NRL exon 1T (Figure 3H, Figure S4). In keeping with the computationally assigned isoforms and differential exon usage, ER libraries had a higher proportion of FL-NRL exon 1 and exon 2 reads and LM libraries had a higher proportion of Tr-NRL exon 1T reads (Figure 3H, Figure S4). Moreover, alternative splicing within NRL exon 2 was more prevalent in LM libraries, affecting 6,788 of 29,177 (23 %) of exon 2 reads compared to 5,378 of 85,860 (6 %) in ER cells (p<0.0001; Chi-square with Yates correction), resulting in rarer full-length (exon 1-2-3) transcripts than inferred from short-read sequencing. Thus, long-read sequencing revealed cell type-biased NRL isoform expression with a preponderance of FL-NRL in early rods and disrupted FL-NRL and Tr-NRL isoforms in LM cones.

Despite our detection of L/M cone RNAs encoding Tr-NRL and FL-NRL, cone expression of NRL protein has not been reported. To assess endogenous Tr-NRL expression, we performed immunoblot analysis of CHLA-VC-RB-31 retinoblastoma cells (36), which express FL-NRL- and Tr-NRL transcripts in a cone-like 0.57:1 ratio, with an antibody that recognizes both FL- and Tr-NRL proteins (Figure S5A,B). As in other retinoblastoma cells (13), FL-NRL increased in response to retinoic acid and proteasome inhibition, whereas Tr-NRL was not detected (Figure 5C). Similarly, Tr-NRL was not detected in EGFP+, RXRγ+ cones following lentiviral transduction of an explanted fetal retina with an EGFP-P2A-Tr-NRL cassette (Figure S5D-G). These findings suggest that Tr-NRL is poorly translated or unstable in most cone precursors.

For THRB, the most highly assigned transcript isoforms included those encoding the L/M cone-specific TRβ2 (ENST00000280696) and the more widely expressed TRβ1 (ENST00000396671 and others) (Figure 4A-C, S6). While both TRβ1 and TRβ2 promote L/M cone fate determination (37), isoforms encoding TRβ2 predominated in L/M cones while those encoding TRβ1 predominated in early rods (Figure 4B). Moreover, late rods preferentially expressed THRB exons 1-6 and the THRB 3’ untranslated region (UTR), implying that RNAs encoding full-length TRβ proteins were rare (Figure 4B,C). Notably, a higher percentage of reads extended from the exon 4 and exon 6 splice donor sequences into the subsequent introns in LR versus LM cells (Figure 4D,E) (p ≤ 0.001 for both, two-tailed Chi square test), suggesting that premature transcription termination (PTT) in introns 4 and 6 preferentially limits TRβ expression in the LR population. The inferred PTT events are consistent with structures of the assigned Ensembl isoforms (Figure 4B, S6).

Differential expression of THRB isoforms in rod and cone precursors.

A. Expression of THRB and highly assigned isoforms ENST00000280696 (encoding TRβ2) and ENST00000396671 (TRβ1). B. Mean THRB isoform assignments for each cluster presented as counts (top) and percentage of counts (bottom); β1 PTT isoforms indicated. C. Top: Mean read counts across Ensembl THRB exons. Bottom: Transcript structures for TRβ1, TRβ2, and two TRβ1 truncations. Green arrowhead: First TRβ2 exon. ENST00000396671 exon numbers are indicated above and protein domains (AF1, DNA-binding (DBD), and ligand binding (LBD)) below. D. Read coverage for LR cells across THRB exons 4 and 6 splice donor sites. E. Percentage of exon splice donor reads that are spliced or readthrough to the subsequent intron. F. Long-read sequencing of pooled 3’ RACE reactions initiated with exon 4 (left) or TRβ2 exon 1 (right) performed on cDNA libraries from 21 LM cells (top) and 5 LR cells (bottom). Schematics show total coverage (above) and individual transcripts (below). TRβ1 and TRβ2 first exons (green boxes) are enlarged at right. Red arrowheads: intronic premature transcription termination (PTT).

Two post-mitotic photoreceptor precursor populations rapidly expressing rod or cone markers.

A. UMAP plot colored by high resolution clusters. B. RNA Velocity plots with cell clusters as in A and cell velocity vectors. C. Enlarged view highlighting RPC and MG clusters (left), RPC-localized iPRP and TR clusters (middle), and RNA velocity (right). Black line: limit of the RPC-localized region. Arrows depict inferred trajectories. D. Violin plots depicting expression of selected genes in RPC, MG, and RPC-localized iPRP and TR cells. Colored asterisks compare cluster of same color to cluster at right of line. E. SCENIC regulon violin and box plots for RPC-localized cells in each cluster, selected from most specific regulons for MG, RPC, ER and LM clusters. ^, p <0.1; *, <0.05; **, <0.005; ***, <0.0005 (post-hoc Dunn test).

To further evaluate PTT events, we used THRB 3’ RACE and long-read sequencing on single cell cDNA libraries from 23 L/M cone cells and five LR cells selected for high THRB expression. RACE reactions were performed separately with primers complementary to the TRβ1-specific exon 4 and to the TRβ2-specific first exon (Figure 4F). Sequencing of LM and LR RACE products corroborated pronounced PTT in introns 4 and 6, with greater intron 6 PTT in LR versus LM cells (Figure 4F, Figure S7). However, we did not corroborate the late rods’ proportionately higher intron 4 PTT, likely due to the small number of LR cells examined and heterogeneity in PTT frequency in individual cells. Long read sequencing also revealed PTT following the TRβ2-specific exon and a novel transcription-terminating exon following the canonical exon 5 solely in TRβ2 transcripts (Figure 4F, S7). 3’ RACE transcripts rarely extended into the 3’ UTR in LM or LR cells, suggesting that reads mapping to this region do not reflect protein-coding RNAs and confound the assessment of THRB expression. These analyses demonstrate that THRB is regulated by multiple PTT events in rod and cone precursors as well as by cell type-specific promoter utilization and independent 3’ UTR expression.

For RXRG, short read sequencing reads were assigned to several isoforms that differed in their 5’ promoter position and exon utilization (Figure S8A-C). However, 5’ RACE and long-read sequencing did not support differential isoform expression (Figure S8D), and quantitative imaging revealed an average ∼3.5-fold higher RXRγ protein in cones compared to rods (Figure S8E). Thus, long-read sequencing clarified that RXRG expression is moderately higher in human cone versus rod precursors without evidence of cell-type specific isoforms.

Two post-mitotic photoreceptor precursor populations

To further define photoreceptor developmental states, we subdivided the initial six cell populations with higher resolution clustering, identified high resolution cluster-associated genes and regulons, and inferred each cell’s rate and direction of transcriptomic change using RNA velocity (38) (Figure 5A,B). Increased clustering resolution divided the RPC/MG cluster into separate RPC and Müller glia (MG) groups, divided L/M cones into four subgroups (LM1 – LM4), which partially overlapped in UMAP space, and yielded a new cluster, here designated transitional rods (TR), derived from the low resolution iPRP and rod clusters (Figure 5C). Similar clusters were observed at reduced k.param values that defines nearest neighbors.

The distinction between RPCs and MGs was corroborated by the expression of known marker genes, with the RPC cluster having increased expression of cell cycle markers (CCNE2, CCNA2, CCNB2) and the cell cycle-associated PBK and E2F7 (Figure 5D, S9A,B, Table S5). The MG cluster was spatially segregated, lacked expression of most cell cycle genes, and expressed genes known to be expressed in both populations (CCND1, SLC1A3, PAX6, LHX2, VSX2) as well as the MG marker RLBP1 (39–41) (Figure 5C,D, S9A). Differential gene expression analyses revealed upregulation of cell cycle-related genes and G2/M checkpoint and E2F target ontologies in the RPC cluster but no significantly upregulated ontologies in the MG population (Figure S9B,C), consistent with early MGs resembling quiescent RPCs (42).

High resolution clustering also resolved iPRP and TR cells positioned near the RPCs in UMAP space (Figure 5C). RNA velocity suggested that these ‘RPC localized’ cell groups flowed towards the larger iPRP population and towards early maturing rods, respectively (Figure 5B,C). Compared to RPCs and MG, RPC-localized iPRP and TR cells had minimal expression of cyclin RNAs or RPC markers PAX6, LHX2, and VSX2, and had increased expression of photoreceptor determinants OTX2 and CRX (Figure 5D), consistent with their being immediately post-mitotic photoreceptor precursors. The iPRP cells in this region also upregulated the early cell fate determinant ONECUT1 (3, 43), the L/M cone determinant THRB, and the cone differentiation marker GNAT2, but lacked the rod determinant NR2E3, indicative of a cone-bias (Figure 5D). In contrast, TR cells in the same region trended towards higher expression of the rod-related NRL and the rod-specific NR2E3 (p= 0.19, p=0.054, respectively). Prior droplet scRNA-seq analyses did not resolve early post-mitotic cone versus rod precursors (15, 16), nor could we discriminate early post-mitotic THRB+ iPRP and NR2E3+ TR populations through a similar analysis of single cell transcriptomes generated by 3’ end-counting. Thus, deep sequencing enabled detection of THRB, NR2E3, and perhaps other low-expressed RNAs that are important to discrimination of early iPRP and TR populations.

SCENIC regulons further clarified the identities of the four RPC-localized clusters (Figure 5E, S10). The MG cluster was best specified (i.e., had highest regulon specificity scores) by PAX6 and SOX9, both previously described in RPCs and MGs (39, 44, 45), whereas RPCs were best specified by E2F2 and E2F3. These RPC and MG regulons were low or absent in the early post-mitotic iPRP and TR cells, suggesting they undergo an abrupt cell state change. Moreover, immediately post-mitotic iPRPs and TRs that colocalized with RPCs showed differential THRB and NRL regulon activities. iPRP cells had greater THRB regulon signal than TR cells and TR cells had higher NRL regulon signal than iPRPs (p<0.05 for both), consistent with the bifurcated direction of gene expression change implied by RNA velocity (Figure 5C,E). TR cells also had lower activity of the pan-photoreceptor regulons LHX3, OTX2, CRX, and NEUROD1 (Figure S10), consistent with their immature nature.

Overall, high resolution clustering resolved two post-mitotic PR precursor populations with iPRPs having more cone-related and TRs having more rod-related gene expression and regulon activities, consistent with RPC-localized TR cells proceeding towards the ER cluster and iPRP cells continuing towards the larger iPRP population. To our knowledge, this is the first indication of two immediately post-mitotic photoreceptor precursor populations with cone versus rod-biased gene expression (15, 16, 19).

Early cone precursors with cone- and rod-related RNA expression

In addition to the immediately post-mitotic iPRP cells, the iPRP cluster included cells bridging the UMAP region between early maturing cones and early maturing rods (Figure 6A). Cells in this region expressed cone markers (GNAT2, THRB), rod markers (GNAT1, NR2E3), or, in many cases, both (Figure 6B), resembling a proposed hybrid cone/rod precursor (4). To determine whether the intermixed cone and rod gene expression reflect a hybrid cone-rod precursor state, we examined bridge region NRL and THRB regulon activities, which embody the overall transcriptomic effects of these factors. Indeed, in the iPRP bridge and adjacent ER regions, cells with NRL and THRB regulon signals intermixed, and some cells showed strong signals for both (Figure 6C, arrows). However, bridge region iPRP cells had low NRL regulon z-scores similar to L/M cones and had THRB z-scores above that of adjacent ER cells (Figure 6D). The difference between regulon signals increased in rod and cone clusters outside the bridge as NRL or THRB regulons became more dominant (Figure 6D).

An iPRP population with cone-rod properties.

A-C. UMAP ‘bridge’ region cells colored by ER and iPRP cluster and RNA velocity (A), rod and cone marker gene expression (B), and NRL and THRB regulon activity (C, arrows indicate cells with both regulon signals). D. Box plot of Z-score-normalized NRL and THRB regulon AUCs for each cluster. E,F, Tiled composite fluorescence image of FW14 retinal section after immunofluorescence staining of RXRγ and NR2E3 and RNA FISH of GNAT2 and NR2E3 (E) and diagram of same retina with the section’s most peripheral (i.e., youngest and least mature) and most central (i.e., oldest and most mature) rods and cones labeled (F). Boxes indicate regions further evaluated as in Supplementary Figure S11 and quantitated in G and H. G,H. Quantitation of (G) outer and (H) inner layer photoreceptor precursors expressing combinations of RXRγ and NR2E3 proteins and GNAT2 and NR2E3 RNAs.

To determine if cells with early cone and rod RNA co-expression can be detected in the developing retina, we performed multiplex RNA FISH for early cone marker GNAT2 and rod marker NR2E3. To identify cell types with such co-expression, FISH analyses were combined with immunostaining for RXRγ, which has high expression in outer neuroblastic layer (NBL) cone precursors and low expression in inner NBL rod precursors (Figure S8E), and for NR2E3, which is detected solely in rods. To infer the spatiotemporal pattern of such expression across human retinal development, we examined GNAT2 and NR2E3 RNA co-expression in outer NBL RXRγ+ cones and in inner NBL RXRγlo rods across 13 peripheral fetal retina regions (Figure 6E,F, S11). The analyses revealed that most of the far peripheral (hence, nascent) outer NBL RXRγ+ cones (Figure 6E, regions 2 and 13) lacked detectable GNAT2 and NR2E3 RNA, whereas those in the more mature central retina (regions 3-12) were uniformly GNAT2+ (Figure 6G). However, starting with regions 3 and 11, some outer NBL GNAT2+ cones were also NR2E3+ (Figure 6G). The proportion of GNAT2+, NR2E3+ outer NBL cells increased in the more central retina, yet all of these cells had strong RXRγ staining and lacked NR2E3 protein, consistent with their having a cone identity. In contrast, most of the inner NBL RXRγlo cells had either low NR2E3 RNA without NR2E3 protein, in nascent rods, or prominent NR2E3 RNA with NR2E3 protein, in maturing rods (Figure 6H). However, we did not detect GNAT2 puncta in inner NBL RXRγlo, NR2E3+ cells, including the most peripheral RXRγlo cells with low-level NR2E3 RNA. In summary, most, if not all, photoreceptor precursors with GNAT2 and NR2E3 RNA co-expression had high RXRγ, no detectable NR2E3 protein, and outer NBL positions expected of cone precursors. This supports the notion that early bridge region iPRP cells with combined cone- and rod-related RNA expression are likely cone-directed.

To further assess whether iPRP bridge region cells comprise a distinct subpopulation, we sought to identify RNA markers of this region. Among four iPRP marker genes (Figure S9A), CHRNA1 was mainly expressed in the bridge region and adjacent rods and cones, whereas ONECUT1 was biased to cone-directed iPRPs, lncRNA CTC-378H22.2 was expressed in a narrow zone of THRB+, GNAT2+ cone precursors, and S100A6 was more widely expressed (Figure 7A, S12A). Bridge region cells also expressed ATOH7 and DLL3, which were previously proposed to define transitional photoreceptor precursors and promote cone fate determination (14–16). ATOH7 was largely restricted to cone-directed cells whereas DLL3 was broadly expressed similar to S100A6 (Figure 7A, S12A). As CHRNA1 was the most specific marker of bridge region cone and rod precursors, we evaluated whether CHRNA1 RNA marked specific cone and rod populations using FISH combined with RXRγ and NRL immunofluorescent staining. In a FW12 retina, we detected highest CHRNA1 in the earliest, most peripheral NRL+ rods and RXRγ+ cones and fewer CHRNA1+ cells in more mature, central regions (p <0.005 in cones) (Figure 7B-D), consistent with its expression in bridge region photoreceptor precursors.

CHRNA1 marks early-born cone and rod precursors.

A. UMAP plots of iPRP bridge region marker genes. B-D Combined RXRγ/NRL immunohistochemical staining and CHRNA1 FISH of FW12 retina. B. Tiled images of retina section with nuclei stained with DAPI. White boxes: fields used for quantitative fluorescent imaging. Distances along apical edge of tissue marked in mm from midpoint of central image (0 mm, C). *: Imaged region shown in C. Scale bar = 500 µm. C. Top: Retinal nuclear and cellular segmentation and identification of cells as RXRγ+ (green outline) or NRL+ (red outline). Yellow box: Field shown below. Bottom: RXRγ or NRL immunofluorescence staining with CHRNA1 FISH. Arrows: RXRγ+CHRNA1+ (green), NRL+CHRNA1+ (yellow). Scale bars = 15 µm. H. Quantitation of fluorescent puncta in RXRγ+ and NRL+ cells by image field. X-axis: Distance from the midpoint of each image to retina center (0 mm, C). **, p <0.005 (Wald test, images from 0-6 mm).

To identify factors that regulate transcriptomic states during iPRP cell fate determination, we examined the distribution of the most iPRP-specific regulons identified after high-resolution clustering, OLIG2 and LHX9 (Figure S12B, Table S6). As Olig2 was previously detected in mouse RPCs in which Onecut1 enabled Thrb expression and cone fate determination (3, 46), the OLIG2 regulon was expected to be most active in RPCs and in early, immediately post-mitotic iPRPs. While the OLIG2 and LHX9 regulons were indeed active in RPCs, both were also active in a narrow zone of ONECUT1+ cone-directed iPRP cells distal to the bridge region and immediately preceding the upregulation of the THRB regulon (Figure S12B). These data indicate that the OLIG2 and LHX9 regulons are active in and potentially relevant to human cone fate determination in post-mitotic photoreceptor precursors.

An early L/M cone trajectory marked by successive lncRNA expression

After fate commitment, L/M cones initiate a maturation process with upregulation of RNAs and proteins related to phototransduction, axonogenesis, synaptogenesis, and outer segment morphogenesis (5, 6). To evaluate whether early L/M cone maturation is comprised of distinct transcriptomic cell states, we assessed marker gene and regulon differences between high resolution clusters LM1, LM2, LM3, and LM4. While clusters LM1-4 were distinguished by sequential increasing expression of ACOT7, RTN1, PDE6H, OLAH, and NPFF, they showed only subtle differences among the three regulons with highest LM1-4 specificity scores, THRB, ISL2, and LHX3 (Figure 8A,B). The lack of cluster-specific marker genes and regulons suggest that LM1-LM4 represent different stages of a graded maturation process.

L/M cone subcluster marker genes, regulons, and pseudotemporal trajectory with successive lncRNA gene expression.

A. Violin plots of high-resolution cone cluster marker genes with increasing maturation-associated expression. All significant differences between adjacent clusters are indicated. B. Violin plots of regulons with highest LM1-4 cone cluster specificity scores. *, p <0.05; **, <0.005; ***, <0.0005 (post-hoc Dunn test). C. Pseudotemporal trajectory through the L/M cone population derived with Monocle 3. *: Root cell used to define endpoint. D. Trendlines of relative count expression (left) and UMAP plots for lncRNAs correlating with early or late-upregulating modules. Line color matched to label boxes. E, F. Combined RXRγ immunohistochemical staining and FISH of lncRNAs on FW16 retina. E. Tiled images of retina with nuclei stained with DAPI. White boxes: fields used for quantitative fluorescent imaging. Distances along apical edge of tissue marked in mm from fovea to ciliary margins. Scale bar = 500 µm. Asterisks identify fields shown in F. F. Combined RXRγ immunostaining and multiplex FISH for four lncRNAs, of which two are shown in peripheral and central retina regions. Arrows: White: RXRγ/CTC-378H22.2+. Magenta: RXRγ/HOTAIRM1+. Blue: RXRγ/CTD-2034I21.2+. Green: RXRγ/RP13-143G15.4+. Scale bar = 15 µm. G. Quantitation of lncRNA fluorescent puncta assigned to RXRγ+ cells after segmentation. Colored bars mark lncRNA expression regions as described in the text. Arrows indicate significant count differences between images (p <0.05, Wald test).

We further evaluated gene expression changes in maturing cones by pseudotemporally ordering iPRP and L/M cones (Figure 8C). This identified 967 pseudotime-correlated genes (q-value < 0.05, expression > 0.05 in >5% of cells; Table S7) in seven gene modules (Figure S13). Among the top 20 pseudotime-correlated genes in each module, we identified four lncRNAs that were sequentially expressed (Figure 8D). To determine if these lncRNAs distinguish developmentally distinct maturing cones in vivo, we probed their expression via multiplex RNA FISH in co-stained RXRγ+ cone precursors across a FW16 retina (Figure 8E,F). Quantitation of FISH puncta defined four cone maturation zones based on expression peaks and significant count differences for each lncRNA: the most peripheral cones with high CTC-378H22.2 and HOTAIRM1, peripheral cones with high HOTAIRM1 only, cones with low expression of all four lncRNAs, and parafoveal/foveal cones with upregulated RP13143G15.4, CTD-2034I21.2, and CTC-378H22.2 (Figure 8G). The detection of foveal CTC-378H22 by ISH but not in late LM transcriptomes may relate to a lack of the most mature cones in our scRNA-seq analyses. These data support the concept that maturing L/M cones sequentially express specific lncRNAs as they develop.

Cone intrinsic SYK contributions to the proliferative response to pRB loss

We next assessed whether early maturing cone transcriptomic features are conducive to the proliferative response to pRB loss (10, 11). Given the high transcriptomic similarity across the L/M cone population, we compared gene expression between all early maturing L/M cones and early maturing rods. This identified 422 genes upregulated and 119 downregulated in cones (p<0.05, log2FC>|0.4|) (Figure S14A, Table S8). Among cone-enriched genes, the top three enriched ontologies related to translation initiation, protein localization to membrane, and MYC targets (Figure S14B). Although MYC RNA was not detected, the upregulation of MYC target genes was of interest given the cone precursors’ high MYCN protein expression (Figure 9A) and the role of MYCN in the cone precursor response to pRB loss (8–10). Concordantly, the LM cluster had increased MYCN RNA (log2FC = 0.54) and MYCN regulon activity, representing the seventh highest LM cluster regulon specificity score (Figure 9B,C, Table S4).

Cone intrinsic MYCN and SYK contribute to the proliferative response to pRB loss.

A. Immunofluorescent staining of MYCN in ARR3+ cones but not in NRL+ rods in FW18 retina. B-D. UMAP plots of MYCN expression (B), MYCN regulon activity (C), and SYK expression (D). E. SYK and MYCN gene expression violin plots by cluster. *, p <0.05; ns = not significant (t-test). F. Immunohistochemical staining of SYK and cone arrestin (ARR3, LUMIF-hCAR antibody) in FW18 and FW16 retinae. Green arrow: ARR3+, SYK+. White arrow: ARR3+, SYK-. Scale bar = 25 µm. G. Effect of SYK inhibitor GS-9876 on KI67+ expression in shRB1- and YFP transduced RXRγ+ cone cells from dissociated FW16.5 fetal retina. Values represent means of three immunocytochemical analyses from two treatment replicates. Error bars: standard deviation. *, p <0.05; **, <0.005 (Student’s T-test with equal variance, 2-tailed). RXRγ+ cells: Experiment 1, n=1,340. Experiment 2, n=804. Range 107 – 366 cells per condition. H. Model of SYK expression and function in cone maturation and retinoblastoma development.

Among other differentially expressed genes, we noted cone-specific upregulation of SYK (Figure 9D, Table S8), which encodes a non-receptor tyrosine kinase. Whereas SYK was previously implicated in retinoblastoma genesis and proposed to be induced in response to pRB loss (47), its expression was not previously reported in developing fetal retina. Indeed, SYK RNA expression increased at the iPRP stage and into clusters LM3 and 4 and was abolished during the transition from early-maturing to late-maturing L/M cones, in contrast to its minimal expression in the corresponding rod TR, ER, and LR stages (Figure 9E). Similarly, immunohistochemical staining revealed high SYK protein expression in immature (ARR3-) and early maturing (ARR3+) cones from the retinal periphery to the maturing foveal cones at FW16, while SYK was not detected in the most mature foveal cones at FW18 (Figure 9F). The loss of SYK protein and RNA expression with cone maturation is consistent with the lack of SYK in cones of normal retina adjacent to a retinoblastoma tumors (47) and implies that SYK is a defining feature of the human early cone precursor state. While MYCN was also preferentially expressed in early cones, it did not increase as much relative to RPCs nor decline as much in late maturing cone precursors when compared to SYK RNA dynamics (Figure 9E).

To determine if SYK contributes to retinoblastoma initiation, dissociated fetal retinal cells were transduced with an RB1-directed shRNA (shRB1-733) known to induce cone precursor proliferation (10), treated with the highly specific SYK inhibitor GS-9876 (48) for 12 days, and examined for cone precursor cell cycle entry by co-staining for RXRγ and Ki67. GS-9876 treatment suppressed the proliferative response to pRB knockdown at all concentrations from 1.0 – 5.0 μM (Figure 9G), indicating that cone precursor intrinsic SYK activity is critical to the proliferative response to pRB loss and is retained in retinoblastoma cells (Figure 9H).

Discussion

This study evaluated cell state changes associated with human cone and rod photoreceptor development using deep full-length scRNA-seq. Whereas prior scRNA-seq studies provided insights into RPC fate determination and trajectories using 3’ end counting (14–18, 25), deep full-length sequencing enabled more precise resolution of cell states and identification of cell type-specific regulon activities and transcript isoforms. Additionally, our cell enrichment strategy provided insight into the transcriptomic profiles and cancer-predisposing features of developing cones, a rare population with unique cancer cell-of-origin properties (8, 10, 11).

One advantage of full-length scRNA-seq is that it enables the detection of differential transcript isoform expression. Here, we show that the rod determinant NRL, the L/M-cone determinant THRB, and the cone maker RXRG are all co-expressed in cone and rod precursors at the RNA level yet use distinct cell type-specific mechanisms to enable appropriate differential protein expression. Long-read cDNA sequencing revealed that L/M cones preferentially express novel NRL transcripts predicted to encode a truncated NRL (Tr-NRL) lacking a transactivation domain as well as NRL isoforms with intra-exon-2 transcription initiation and alternative splicing. As with DD10 (35), Tr-NRL opposed transactivation by full-length NRL and thus may suppress rod-related transcription. However, our inability to detect intrinsic Tr-NRL or to overexpress Tr-NRL in cone precursors suggests there are additional layers of regulation, possible effects of untranslated Tr-NRL RNA, and alternative contexts in which these isoforms act. A similar assortment of NRL isoforms were expressed in rods, yet a far higher proportion of rod NRL transcripts encoded full-length NRL protein. Thus, L/M cone precursors’ greater use of the Tr-NRL first exon and NRL exon 2 disruptions reveal a multipronged approach to suppress FL-NRL function while downregulating overall NRL RNA by only ∼ 4-fold.

Similarly, we uncovered cell type-specific differences in THRB transcript isoforms, albeit generated through premature transcription termination (PTT) and 3’ UTR expression in late rod precursors. PTT is widely used to govern expression of transcription regulators (49), and while PTT-shortened THRB transcripts had been identified (50), their retinal cell specificity was not previously recognized. In contrast, the expression of THRB 3’ UTR sequences independent of THRB protein-coding sequences was not previously reported and may enable additional regulation (51). As a general matter, independent 3’ UTR transcripts may be inferred from full-length scRNA-seq but may confound the interpretation of scRNA-seq performed with 3’ end-counting. To address this issue, our publicly available Shiny app displays exon usage for all genes expressed in the RPC and photoreceptor clusters in this study.

One limitation of these studies is that individual cells may express a small and varying spectrum of transcript isoforms. While this variability was mitigated by combining cDNAs of cells deemed to be in similar states based on short-read sequencing, analysis of more cells might better reveal the spectrum of isoforms expressed by specific cell populations.

Understanding human photoreceptor development requires the identification of photoreceptor lineage developmental states and elucidation of mechanisms underlying their transitions. While cell states may be defined in various ways, at the transcriptome level they are perhaps best defined by unique combinations of transcription factor regulon activities (29). Our deep sequencing approach uncovered discrete cell states distinguished by gene expression as well as by regulon activities. For example, early and late maturing rod populations were distinguished by the latter’s increased expression of the NRL, ATF4, CRX, and LHX3 regulons and decreased HMX1 and RAX regulons. Similarly, L/M cones formed an early maturing cone cluster characterized by high THRB and ISL2 regulons, consistent with THRB and ISL2 driving L/M cone identity (4, 15, 52), and a late maturing cone group with decreased RAX activity. Downregulation of the RAX regulon in late maturing rod and cone precursors is consistent with decreasing RAX protein during photoreceptor maturation and with RAX-mediated suppression of the cone opsin and rhodopsin promoters (53, 54).

Increased clustering resolution further divided the L/M cone cluster into four subgroups with subtle maturation-associated changes in marker gene expression and regulon activity, consistent with LM1-4 comprising different stages of a graded maturation process. However, trajectory analyses revealed successive expression of lncRNAs that was validated in developing retinal tissue. Three of these lncRNAs (HOTAIRM1, CTD-2034I21.1 (neighbor gene to CTD-2034I21.2), and the mouse ortholog of CTC-378H22.2 (D930028M14Rik)) were previously observed in cone scRNA-seq (15, 17, 55). Their sequential expression in the absence of discrete regulon changes suggests that they demarcate early L/M cone precursor substates.

High resolution clustering also segregated the immature photoreceptor precursor (iPRP) and transitional rod (TR) populations. iPRP and TR cells positioned adjacent to RPC/MG cells in UMAP plots showed higher cone-like or rod-like gene expression and regulon activities, respectively, and had RNA velocities directed toward distinct cone and rod populations. Prior droplet-based scRNA-seq analyses have not to our knowledge distinguished immediately post-mitotic cone- and rod-directed photoreceptor precursors. However, scRNA-seq and scATAC-seq analyses showed distinct early neurogenic precursors directed in part towards cones and late neurogenic precursors directed in part towards rods (14, 15, 19). Thus, as one possibility, early iPRPs may comprise a cone-directed subset of early neurogenic cells while early TRs comprise a rod-directed subset of late neurogenic cells.

Although not identified as a distinct cluster, our analyses also revealed iPRP ‘bridge region’ cells that co-expressed rod and cone genes and regulons. This population resembled the T3 transition and early neurogenic precursor states identified via 3’ end-counting and marked by expression of ATOH7 and DLL3 (15, 16), although the bridge region was more precisely delineated by the new marker CHRNA1. Bridge region cells with velocity directed towards early cones expressed THRB and ONECUT1 RNAs and had OLIG2 regulon activity; as these elements were proposed to promote cone fate in lineage-restricted RPCs (3), their expression in the iPRP population suggests a similar role in post-mitotic photoreceptor precursors. Conversely, the status of bridge region cells with velocity directed towards early rods is less clear; while the velocity vectors and concurrent cone- and rod-related gene expression are consistent with iPRP cells that follow an alternative route to rods, our finding that only outer NBL, RXRγhi cells co-express cone marker GNAT2 and rod marker NR2E3 RNAs suggests that most cells with combined cone- and rod-related gene expression are cone-committed. Moreover, the early iPRP and LM cone precursors’ expression of NR2E3 and NRL RNAs suggest that their presence in retinoblastomas (12, 13) reflects their expression in the L/M cone precursor cells of origin.

We also mined gene expression differences in early cone versus rod photoreceptors to identify factors that impact human cone sensitivity to RB1 loss (10, 11). We detected upregulation of genes targeted by MYC along with upregulated MYCN RNA, MYCN protein, and MYCN regulon activity. As MYCN is required for the proliferative response to pRB loss (10) and triggers retinoblastoma when amplified and overexpressed (8), these findings suggest that increased MYCN RNA expression and regulon activity contribute to pRB-deficient cone precursor proliferation and retinoblastoma genesis. We further found that cone precursors highly express SYK, an oncoprotein previously detected in human retinoblastomas and RB1-null retinal organoids but not in healthy tumor-associated retina (47, 56). The high SYK expression preceding early cone precursor maturation and its loss during late maturation implies that it is a retinoblastoma cell-of-origin-specific feature. SYK inhibition impaired pRB-depleted cone precursor cell cycle entry, implying that native SYK expression rather than de novo induction contributes to the cone precursors’ initial proliferation (Figure 9H).

In summary, through deep, full-length RNA sequencing, we identified photoreceptor cell type-specific differences in transcript isoform expression and post-mitotic photoreceptor precursor states with distinctive gene expression and regulon activities. The discrimination of these states enables the identification of developmental stage-specific cone precursor features that underlie retinoblastoma predisposition.

Materials and methods

Primary human tissue

Human fetal samples were provided by Advanced Bioscience Resources, Inc., Alameda, CA, or collected under Institutional Review Board approval at the University of Southern California and Children’s Hospital Los Angeles. Following the patient decision for pregnancy termination, patients were offered the option of donation of the products of conception for research purposes, and those that agreed signed an informed consent. This did not alter the choice of termination procedure, and the products of conception from those who declined participation were disposed of in a standard fashion. Fetal age was determined according to the American College of Obstetrics and Gynecology guidelines (57). Each retina subjected to sequencing was from a different donor.

Retinal dissociation and RNA-sequencing

Single cell isolation

Retinae were dissected and dissociated in ice-cold phosphate-buffered saline (PBS) as described (10). Briefly, retina was removed and placed in 200 µl Earle’s balanced salt solution (EBSS) in a 6-well plate on ice. 2 ml of 10 u/ml papain (Worthington LK003176) solution was added and then incubated at 37°C for 10 min, followed by pipetting with a 1000 µl pipet tip to break up large pieces and additional 5 min incubations until cells were dissociated to single cell level. An additional 1-2 ml papain was added after 20 min if dissociation was insufficient (∼40 min total). Retina culture media [Iscove’s Modified Dulbecco’s Medium (IMDM; Corning), 10% FBS, 0.28 U/mL insulin, 55 µM β-mercaptoethanol (Sigma-Aldrich), glutamine, penicillin, and streptomycin] was used to stop enzyme activity and cells were centrifuged in 14 ml round bottom tubes at 300 x g, 4°C for 10 min. Supernatants were centrifuged at 1100 x g for an additional 3 min. After resuspension in HBSS, tubes were recombined and cells counted.

FACS isolation of single cells and cDNA synthesis

Single cells were FACS-isolated as described (10) with differences as follows. Dissociated cells were centrifuged and resuspended in a 5% fetal bovine serum in PBS (FBS/PBS) with CD133-PE (Miltenyi Biotec 130-080-801), CD44-FITC (BD Pharmingen 555478) and CD49b-FITC antibodies (BD Pharmingen 555498) to a final concentration of 10,000 cells/µl. After incubation at room temperature (RT) for 1 h, samples were diluted in 5% FBS/PBS, centrifuged as above, and resuspended in 300-400 µl 1% FBS/PBS containing 10 µg/ml 4′,6-diamidino-2-phenylindole (DAPI). Single cells were sorted on a BD FACSAria I at 4°C using 100 µm nozzle in single-cell mode into 1.2 µl lysis buffer droplets on parafilm-covered glass slides (S. Lee et al., in preparation). Gating removed low forward- and side-scatter cells before collecting a CD133-high, CD44/49b-low population. Upon collection of eight cells, droplets were transferred to individual low-retention PCR tubes (Bioplastics K69901, B57801) on ice. cDNA was prepared and amplified using the SMART-Seq V4 Ultra Low Input RNA Kit (Takara Bio 634891) in 10x reduced volume reactions using a Mantis liquid handling system (S. Lee et al., in preparation). Samples were stored at −20°C until quantitation and library preparation. All RNA/cDNA volumes during FACS and processing were handled using low retention siliconized pipette tips (VWR 53503-800, 535093-794).

C1 isolation of single cells and cDNA synthesis

Dissociated cells were FACS isolated as above and collected into a single 1.5 ml tube. Cells were centrifuged and resuspended at 400-800 cells/µl, combined with C1 suspension reagent, and loaded onto the C1 chip and imaged at each site to define cell number in each well. cDNA was synthesized using SMARTer chemistry (SMARTer Ultra Low RNA Kit for the Fluidigm C1 System, Clontech 634835/634935) (three retinae, Seq 1, FW17 and FW13-1), or SMART-Seq V4 (all others). Final cDNA was harvested into low-retention tube strips and stored at −20°C.

Quality control, library preparation and sequencing

DNA was quantitated using Quant-iT PicoGreen dsDNA assay and quality checked by Bioanalyzer. Libraries > 0.05 ng/µl were prepared using the Nextera XT DNA Library Preparation Kit (Illumina, FC-131-1096) and sequenced on the Illumina NextSeq 500 (2×75) (Seq 1) or on Illumina HiSeq 4000 (2×75) (all others).]

5’ and 3’ RACE and nanopore sequencing

1 μl of amplified single cell cDNAs libraries produced as above (23 ER cells, 21 LM cells, and 5 LR cells) were re-amplified using the 5’ PCR Primer II A with CloneAmp HiFi PCR Premix (Clontech Laboratories, Inc. Takara Bio #639298, Table S9) by heating to 98°C x 2 min followed by 20 cycles of 98°C x 10 s, 65 °C x 15 s, and 72 °C x 35 s, followed by 72 °C x 5 min. Re-amplified cDNAs were quantified using QubitTMFlex Fluorometer (InVitrogen) and 50 pg of each re-amplified library used for separate RACE PCR reactions with gene-specific NRL, RXRγ, TRβ1 and TRβ2 primers and universal 5’ RACE and 3’RACE primers (Table S9) using CloneAmp HiFi PCR Premix by heating to 98°C x 2 min followed by 30 cycles of 98°C x 10 s, 60 °C x 15 s, and 72 °C x 35 s, followed by 72 °C x 5 min. Equal volumes of RACE PCR products were pooled according to cell type, and DNA concentrations of LM (32.8 ng/μl), ER (22.6 ng/μl), and LR (16.5ng/μl) samples determined using Qubit fluorometer. Long read sequencing libraries were prepared and samples barcoded using Oxford Nanopore Technology (ONT) Native Barcoding Kit 24 V14 as specified by the manufacturer for average DNA fragment lengths of ∼ 2kb. 200-300ng of each DNA pool was subjected to repair/dA-tailing (End-Prep) with NEB Ultra II End-Prep Enzyme Mix (New England Biolabs, Ipswich, MA) followed by Native Barcode Ligation as specified by ONT with NB01 assigned to the ER pool, NB02 assigned to the LM pool, and NB03 assigned to the LR pool. Ligation reactions were quenched by addition of 4 μl EDTA and each reaction mixture (24 μl) pooled for a 72 μl volume. The final pool was incubated with Native Adapter (NA) and NEB Quick T4 ligase. The adapter-ligated pool was recovered by AMPure XP magnetic beads and then loaded onto a PromethION flowcell (R10.4.1) as specified by the manufacturer. Sequencing was performed on a PromethION 24 system using MinKNOW software for 72 hours. Raw FAST5 data files were converted to FASTQ files using ONT MinKNOW base calling software.

Computational Methods

Software packages used in this study are described in Table S10.

Read processing and dimensionality reduction

Adapter sequences were removed using the trimgalore wrapper for Cutadapt (58). Trimmed FASTQ files were used as input for HISAT2 (59) with a non-canonical splice penalty of 20, maximum and minimum penalties for mismatch of 0 and 1, and maximum and minimum penalties for soft-clipping set to 3 and 1. Aligned bam files were quantified with StringTie (60) and Tximport (61), yielding cell-by-transcript and cell-by-gene count matrices annotated according to Ensembl build 87. All analyses can be reproduced using a custom snakemake (62) workflow accessible at https://github.com/whtns/ARMOR.

For short-read sequencing, dimensionality reduction and data visualization were performed using the Seurat (V3) toolset (63, 64). Expression counts from all sequencing datasets were integrated using Seurat’s standard integration workflow. Seven sample sequencing batches were integrated after default normalization and scaling using the top 2000 most variable features in each set to identify anchor features. Normalized read counts for gene expression are reported as raw feature read counts divided by total cell read counts, then multiplied by a scaling factor of 10,000 and natural log transformed. These features of the integrated dataset were used to calculate principal component analysis (PCA) and UMAP embeddings (65, 66). A nearest-neighbors graph was constructed from the PCA embedding and clusters were identified using a Louvain algorithm at low and high resolutions (0.4 and 1.6) (67).

Read coverage plots for genes of interest were generated using wiggleplotr (68) to visualize BigWig files with ENSEMBL isoforms. Individual exon counts were identified using DEXseq (69), which takes exons from available isoforms and bins them into intervals before assigning any read counts that overlap a bin to that same bin. These counts were normalized for length for calculation of fold change or relative difference of exon use. Reads crossing target splice donor sites were evaluated as spliced or unspliced from BAM files to determine intron readthrough.

Analysis of long-read sequencing data was performed with the nf-core nanoseq version 3.1.0 pipeline (70) (https://nf-co.re/nanoseq/3.1.0). Quality control on raw reads was conducted using FastQC. Reads were aligned using minimap2 (71). SAM files were converted to coordinate-sorted BAM files and mapping metrics were obtained using samtools. bigWig coverage tracks were created for visualization using BEDTools and bedGraphToBigWig. bigBed coverage tracks were created using BEDTools and bedToBigBed. Transcripts were reconstructed and quantified using bambu. Quality control results for raw reads and alignment results were presented using MultiQC. All reads whose ends overlapped gene-specific RACE primers were displayed and quantitated using the Integrated Genome Browser (IGV). For THRB 3’ RACE reactions, reads inferred to initiate through exonic internal oligo(dT) priming were removed based on the following features adjacent to the 3′ end of the alignments: six continuous adenines (As), more than seven As in a 10 nucleotide window, AG-runs of six or more nucleotides, eight or more As or Gs in a 10 nucleotide window, eight or more As or high A/T content (27 out of 30 bases), or 12 or more adenines present in an 18 nt window (72). The proportion of alternative NRL exon 2 splicing was calculated based on the maximum read coverage between the exon 2 splice acceptor and the first splice donor and the minimum coverage after the internal splice donor as reported in IGV.

Differential expression and overrepresentation analyses

Marker features for each cluster were identified using the Wilcoxon-rank sum tests and specificity scores were computed using the genesorteR (73) R package. Only genes with an adjusted p-value (pAdj) <0.05 were considered for further analyses. Differential expression was performed using the Seurat FindMarkers function with the Wilcoxon-rank sum test on log-normalized counts. Results were displayed as volcano plots made with EnhancedVolcano (74).

Overrepresentation analyses were performed with WebGestalt (75). Analyses were run with the default settings (5-2000 genes per category, Benjamini-Hochberg multiple-testing correction), gene list enrichment was compared to the genome reference set and ontologies were displayed with a false discovery rate (FDR) <0.05 and ‘weighted set cover’ redundancy reduction. All gene sets evaluated were provided within WebGestalt and the same three were used unless otherwise indicated: GO – Biological Process, KEGG, and Hallmark50.

Transcription factor regulons

Transcription factor regulons were identified using pySCENIC, the python-based implementation of Single-Cell Regulatory Network Inference and Clustering (SCENIC) (29, 76). The tool was run using the basic settings as shown (https://pyscenic.readthedocs.io/en/latest/tutorial.html). Initial filtering required a gene to have a minimum of 3 raw counts in 1% of cells to be considered for inclusion in a regulon. AUC scores were z-score normalized for comparison between regulons.

Trajectory analysis

RNA velocity analysis was performed with R package velocyto (38). Spliced and unspliced count matrices were assembled using the run_smartseq2 subcommand with repeat-masked ENSEMBL build 87. RNA velocity was calculated using cell kNN pooling with a gamma fit based on a 95% quantile and velocities were overlaid on integrated UMAP visualizations using velocyto. Pseudotemporal trajectories were calculated using Monocle 3 (77). The SeuratWrappers package provided integration between Seurat’s final integrated dimensional reduction and Monocle. The data was subset to include only cone-directed iPRP and L/M cones, then a principal graph was fit across the UMAP plot and a ‘root cell’ chosen to represent the latest maturation point in the OPN1LW+ late maturing cone group. After assigning pseudotime values to each cell, genes that significantly changed as a function of pseudotime were identified and those with a correlation q-value of <0.05 with expression > 0.5 in at least 5% of the cells present in the pseudotime were grouped into modules of co-regulation by performing UMAP and Louvain community analysis.

Histology

Fixation and cryosectioning of fetal retina

Retinae were procured and dissected in cold 1x PBS. The cornea and lens were removed with a cut around the limbus of the iris leaving the front of the eye open. The tissue was submerged in ∼25 ml of 4% paraformaldehyde (PFA) in 1x TBS and placed at 4°C on a rocker at low speed for ∼16-18 hrs. The tissue was washed 3 times with 1x PBS before incubation in 30% sucrose [sucrose + 1x PBS] overnight at 4°C. A mold was made by cutting off the top 5 ml of a 50 ml conical tube and placed on dry ice. A solution of 1:2 OCT Compound:30% sucrose was mixed and centrifuged to remove bubbles. The mold was partially filled before adding the dissected eye and then covering fully. 10 µm frozen sections were cut at ∼-24°C with the front of the eye facing side on to the blade. Explanted retinae were fixed for 15 min in 4% PFA, washed with 1x PBS, incubated in 30% sucrose for 15 min, after which the membrane was cut from its frame and transferred to a mold containing 1:2 OCT:30% sucrose and frozen.

Immunohistochemical staining

For SYK and ARR3 co-staining, tissue sections were warmed at RT until dry and then washed in Coplin jars twice with 1x PBS for 5 min at RT. Samples were permeabilized in 1x PBS-T for 5 min, washed again in 1x PBS twice for 5 min, blocked for 1 hr in blocking solution [1x PBS, 0.1% Triton X-100, 5% normal donkey serum, 5% normal horse serum]. Primary mouse anti-SYK and rabbit anti-ARR3 (LUMIF-hCAR) antibodies were mixed in blocking solution and applied to samples overnight (16-18 h) at 4°C. Slides were washed three times with 1x PBS for 10 min each, probed with secondary antibodies (Key Resources Table) in blocking solution for 1 hr at RT, washed twice 1x PBS for 15 min each, incubated in1x PBS containing 10 µg/ml DAPI for 10 min, and mounted with Mowiol solution. Additional immunohistological staining of tissue sections and cells on coverslips was performed as described (11) but without initial EDTA wash.

Combined in situ hybridization and immunohistological staining

Human retinal sections were prepared as above but using RNase-free reagents for all stages (PBS, TBS) and ultrapure sucrose (JT Baker, 4097-04) and cryostats cleaned with ELIMINase (Decon Labs, Inc) and 70% ethanol prior to use. RNA FISH hybridization chain reaction probes were designed by and obtained from Molecular Instruments, Inc, with 20 20-bp single strand DNA probes per target RNA sequence except Tr-NRL exon 1 (4 probes), FL-NRL exons 1 and (8 probes), and HOTAIRM1 (14 probes). The accession numbers for target sequences are shown in Table S11, except for RP13-143G15.4 which included regions of several isoforms for maximum gene coverage. The in situ protocol was performed with fluorescent hairpins and buffers from Molecular Instruments using the manufacturer’s instructions for mammalian cells on a slide (https://files.molecularinstruments.com/MI-Protocol-RNAFISH-MammalianCellSlide-Rev7.pdf), beginning with a 4% PFA treatment after thawing and drying slides at RT for 10 min, with the following changes: samples were incubated in 70% ethanol overnight in a Coplin jar, probes were added to the final hybridization volume at 4x the recommended concentration for tissue sections (2.4 pmol in 150 µl of hybridization buffer), tissues were pre-hybridized at 42°C and then lowered to 37°C once probes were added, and hairpins were used at a volume of 1 µl (6 pmol) in 50 µl amplification buffer per section. After removing amplification solution and washing, samples were directly used in the immunofluorescence protocol. Briefly, samples were blocked in a basic RNase-free solution (1% BSA, 0.1% Triton X-100, 0.05% Tween-20 in PBS) for 1 hr and primary antibodies mixed in blocking solution and applied to samples overnight at 4°C. Samples were washed three times for 10 min with 1x PBS, then incubated with secondary antibodies in blocking solution for 40 min at RT. Samples were washed as above with DAPI in the third and final wash, and mounted with Mowiol before imaging using a Leica STELLARIS 5 inverted confocal microscope. For GNAT2 and NRL FISH, samples were washed with DAPI in the third and final wash, mounted and then imaged as above. After this first imaging, coverslip was removed, slides were washed in PBS, and then were used in the immunofluorescence protocol with an alternate blocking solution) (2.5% Horse, donkey, human sera, 1% BSA, 0.1% Triton X-100, 0.05% Tween-20 in TBS).

Quantitation of FISH

For NRL, CHRNA1, and lncRNAs, QuPath (78) was used to identify nuclei, predict cell expansion, classify cells, and count FISH puncta. Apical retina regions containing RXRγ+ and/or NRL+ cells were visually selected for evaluation. The StarDist extension (79) was used to outline nuclei and cell expansion zones with deep learning model ‘dsb2018_heavy_augment.pb’ on the DAPI image channel (threshold = 0.6, pixelSize = 0.15). Model results were reviewed and cells with partial, multiple, or overlapping nuclei removed from consideration. FISH puncta were thresholded then output, using the subcellular detection function (expected spot size = 0.5 µm2, min spot size = 0.2 µm2, max spot size = 0.7 µm2 (NRL isoforms) or 1 µm2 (CHRNA1 and lncRNAs) to capture the most puncta above background. Image positions were determined by tracing the retina apical edge with line segments in FIJI, identifying the closest point on the line to the midpoint of each image, then measuring the distance from the line starting position. For GNAT2 and NR2E3 FISH, images were analyzed in FIJI. The outer layer of cells was delimited manually based on the RXRγ+ cells and the number of RXRγ+ and/or GNAT2+ and/or NR2E3+ cells were manually counted. For the inner layer, the number of NR2E3 RNA positive cells and NR2E3 protein positive cells were counted.

NRL isoform analyses

Proteasome inhibition of CHLA-VC-RB-31

106 CHLA-VC-RB-31 cells (36) were cultured in 2 ml retina culture medium in a 24-well plate with or without 10 µM retinoic acid (Cayman Chemical 11017) and incubated at 37°C in 5% CO2. MG132 (Sigma-Aldrich 474790) was added at 23 h to a final concentration of 10 µM and cells incubated for 4 h. Cells were collected and centrifuged at 300 x g for 4 min, resuspended in 60 µl cold 1x RIPA buffer (Millipore 20-188), 10% sodium dodecyl sulfate (SDS), protease and phosphatase inhibitors diluted per manufacturer instructions (Sigma-Aldrich 5892970001, 4906837001)], incubated on ice for 30 min, then centrifuged at 4°C for 20 min at 14,000 RPM. 15 µl of supernatant were mixed with 3 µl 6X sample buffer (300 mM Tris, 60% glycerol,12% SDS, 86 mM B-mercaptoethanol, 0.6 mg/ml bromophenol blue), heated to 95°C for 5 min, and separated on a 4-12% Bis-Tris gel (Invitrogen, NP0321) using 1xMOPS running buffer (Life Technologies NP0001). Protein was wet transferred to a PVDF membrane in Towbin transfer buffer (25 mM Tris, 192 mM glycine pH 8.3,10% methanol) at 20 V ON at RT. The membrane was blocked in 1xTBS-T containing 5% w/v nonfat dry milk for 1 hr, washed with TBS-T three times for 5 min, incubated overnight in 0.1% dry milk TBS-T solution containing primary antibody at 4°C with slow rocking, washed three times for 15 min in TBS-T, incubated with HRP-conjugated secondary antibody for 1 hr at RT in the 0.1% milk solution, washed six times for 15 min, incubated in chemiluminescent substrate (Thermo Fisher, 34094) and imaged.

NRL isoform and reporter constructs

The pcDNA4-His-Max-C-Nrl (80) (gift of A. Swaroop) CMV promoter was replaced with EF1α promoter to produce pcDNA4-His-Max-C-EF1a-FL-NRL. Using In-Fusion (Takara Bio) the His-Max tag was removed to produce full-length NRL expression plasmid pCDNA4-C-EF1a-FL-NRL (primers #1,2). The His-Max tag plus first 417 bp of NRL N-terminal coding sequence were removed to produce truncated NRL expression plasmid pCDNA4-C-EF1a-Tr-NRL (primers #1,2,3,4). pcDNA-C-EF1a-empty transfection carrier plasmid was made by removing NRL coding sequence from pCDNA4-C-EF1a-Fl-NRL (primers #5,6). pGL3-PDE6B-146-luc was produced by amplifying a 146 bp promoter region of PDE6B (81) from H9 hESC genomic DNA and inserting into a pGL3-SV40 (Promega, primers #7,8), then removing the SV40 promoter (primers #9,10) to generate pGL3-PDE6B-146. The promoter-less pGL3-empty was produced by removing SV40 sequences from pGL3-SV40 (primers #9,11). pUltra-EGFP-FL-NRL was produced by amplification of the FL-NRL coding sequence from pcDNA4-C-EF1α-FL-NRL and inserting into BamHI digested pUltra-EGFP (Addgene 24129) (primers #12, 13). All primers are listed in the Key Resources Table.

NRL expression and analysis

pcDNA4-C-EF1α-NRL constructs were transfected into HEK293T cells. 4 µg plasmid in 1 ml DMEM was combined with 1 ml DMEM + 12 µg PEI and combined with 106 trypsinized HEK293T cells, plated into two wells of a 6-well plate with an additional 1 ml of DMEM + 10% FBS, then cultured overnight in a 37°C incubator. Media was replaced with fresh 293T media the next day and after two additional days cells were collected, lysed, and protein quantitated via BCA assay (Thermo Scientific PI23227). 30 µg of each well was used for western blot as above, with membranes probed with antibody raised against the NRL N-terminus (Santa Cruz SC-374277) or against full-length NRL (R&D AF2945).

Luciferase assays

575 ng of total DNA (50 ng reporter, 100ng FL-NRL and/or 100-300 ng Tr-NRL expression vector, and the remainder pcDNA4C-EF1α empty vector) in 5 µl DMEM was mixed with 2.3 µg PEI in another 5 µl DMEM (per condition, in triplicate) and then plated into one well of a white bottom 96-well cell culture plate with 50,000 NIH3T3 cells in 100 µl 293T media. The luciferase assay was carried out 48 h after transfection with the Nano-Glo Dual-Luciferase Reporter Assay System (Promega N1610) per manufacturer instructions. Luminescence was recorded using the Promega Glomax Multi+ Detection System.

Truncated NRL overexpression in fetal retina

Concentrated pUltra-EGFP-NRL-205 and control pUltra-EGFP lentivirus were produced as described (82). At 16-24 h after 293T transfection in 15 cm dishes, media was exchanged for 20 mL UltraCULTURE (Lonza Inc.) containing 1% HEPES, 1% GlutaMax and 1% Penicillin-Streptomycin. After 60-64 h, media was harvested, centrifuged at 3000 x g for 10 min at 4°C, filtered through 0.45 µm PVDF flask filter, concentrated via tangential flow filtration using a Midikros 20 cm 500 KD 0.5 mm column (D02-5500-05-S), with both input and final supernatant containers kept in ice. The concentrate was re-filtered using a 0.45 µm PVDF syringe filter (Millipore), stored in aliquots at −80°C, and titered using a p24 ELISA kit (ZeptoMetrix 801002).

As described (11), fetal eyes were washed in 70% ethanol, washed three times in 1x sterile PBS, and dissected in cold 1x PBS. The cornea was removed and tissue cut ∼one-third of the distance towards the posterior pole along 4 lines approximately equidistant and between attached tendons. The retina was removed in this flattened state after cutting the optic nerve, transferred to a polytetrafluoroethylene culture insert (Millipore, PICM0RG50) with photoreceptor side down, and placed in a 6-well plate with 1,200 µl retinal culture medium. Retinae were infected as described, cultured for 7 days with half media changes every other day, then fixed in 4% PFA for 15 min, washed three times in 1x PBS, equilibrated in sucrose (30% in PBS) for 15 min, and embedded and sectioned as above. Immunostaining and imaging were performed as above.

RB1 knockdown and SYK inhibitor treatment

FW18.5 retina was partially dissociated with papain, cultured overnight in retinal culture media at 37°C in a 6-well plate, then frozen in 10% DMSO solution and stored in liquid nitrogen. Samples were thawed and revived in retinal culture media overnight and infected with lentivirus carrying shRNAs targeting RB1 (shRB1-733) (10) or a scrambled control (shSCR) and diluted β-mercaptoethanol, insulin, glutamine, penicillin, and streptomycin to the final retina culture medium concentrations along with 5 µg/ml Polybrene, for 24 hrs. After replacing 2/3 of media in each well with fresh media, cells were treated with GS-9876 (MedChemExpress, CAT# HY-109091) in DMSO at or with the same volume of DMSO vehicle control. After 12 d, cells were attached to poly-L-Lysine-coated coverslips for 3 h, fixed in 4% PFA for 10 min, washed in 1x PBS three times, and stored at −20°C until immunocytological staining.

Statistical Analysis

Statistical methods and packages are cited in the Methods text and figure legends. Statistical comparison between more than two groups of cells for gene expression, regulon activity, or exon count proportions used the non-parametric Kruskal-Wallis test followed by pairwise post-hoc Dunn tests with Benjamini-Hochberg correction. Mean expression levels of NRL and RXRG isoforms across cell clusters were evaluated by Welch’s t-tests for groups with unequal variance; p-values for these tests were estimated using bootstrapping with 1,000,000 replications per comparison. Changes in FISH puncta counts were examined within each channel individually using zero-inflated negative binomial regression models with robust standard error terms. Distance across the sample tissue was used to model degenerate zero inflation. P values reported in the manuscript text and legends are Wald Chi-Squared tests that evaluate differences in count between inflection points. Analyses were performed using Microsoft Excel, the R statistical language in RStudio, or Stata SE v14.2.

eLife Assessment

In this paper, the authors use single-cell RNA sequencing to understand post-mitotic cone and rod developmental states and identify cone-specific features that contribute to retinoblastoma genesis. The work is important and the evidence is generally convincing. The findings of rod/cone fate determination at a very early stage are intriguing.

Reviewer #1 (Public review):

Summary:

The authors have used full-length single-cell sequencing on a sorted population of human fetal retina to delineate expression patterns associated with the progression of progenitors to rod and cone photoreceptors. They find that rod and cone precursors contain a mix of rod/cone determinants, with a bias in both amounts and isoform balance likely deciding the ultimate cell fate. Markers of early rod/cone hybrids are clarified, and a gradient of lncRNAs is uncovered in maturing cones. Comparison of early rods and cones exposes an enriched MYCN regulon, as well as expression of SYK, which may contribute to tumor initiation in RB1 deficient cone precursors.

Strengths:

(1) The insight into how cone and rod transcripts are mixed together at first is important and clarifies a long-standing notion in the field.

(2) The discovery of distinct active vs inactive mRNA isoforms for rod and cone determinants is crucial to understanding how cells make the decision to form one or the other cell type. This is only really possible with full-length scRNAseq analysis.

(3) New markers of subpopulations are also uncovered, such as CHRNA1 in rod/cone hybrids that seem to give rise to either rods or cones.

(4) Regulon analyses provide insight into key transcription factor programs linked to rod or cone fates.

(5) The gradient of lncRNAs in maturing cones is novel, and while the functional significance is unclear, it opens up a new line of questioning around photoreceptor maturation.

(6) The finding that SYK mRNA is naturally expressed in cone precursors is novel, as previously it was assumed that SYK expression required epigenetic rewiring in tumors.

Weaknesses:

(1) The writing is very difficult to follow. The nomenclature is confusing and there are contradictory statements that need to be clarified.

(2) The drug data is not enough to conclude that SYK inhibition is sufficient to prevent the division of RB1 null cone precursors. Drugs are never completely specific so validation is critical to make the conclusion drawn in the paper.

Reviewer #2 (Public review):

Summary:

The authors used deep full-length single-cell sequencing to study human photoreceptor development, with a particular emphasis on the characteristics of photoreceptors that may contribute to retinoblastoma.

Strengths:

This single-cell study captures gene regulation in photoreceptors across different developmental stages, defining post-mitotic cone and rod populations by highlighting their unique gene expression profiles through analyses such as RNA velocity and SCENIC. By leveraging full-length sequencing data, the study identifies differentially expressed isoforms of NRL and THRB in L/M cone and rod precursors, illustrating the dynamic gene regulation involved in photoreceptor fate commitment. Additionally, the authors performed high-resolution clustering to explore markers defining developing photoreceptors across the fovea and peripheral retina, particularly characterizing SYK's role in the proliferative response of cones in the RB loss background. The study provides an in-depth analysis of developing human photoreceptors, with the authors conducting thorough analyses using full-length single-cell RNA sequencing. The strength of the study lies in its design, which integrates single-cell full-length RNA-seq, long-read RNA-seq, and follow-up histological and functional experiments to provide compelling evidence supporting their conclusions. The model of cell type-dependent splicing for NRL and THRB is particularly intriguing. Moreover, the potential involvement of the SYK and MYC pathways with RB in cone progenitor cells aligns with previous literature, offering additional insights into RB development.

Weaknesses:

The manuscript feels somewhat unfocused, with a lack of a strong connection between the analysis of developing photoreceptors, which constitutes the bulk of the manuscript, and the discussion on retinoblastoma. Additionally, given the recent publication of several single-cell studies on the developing human retina, it is important for the authors to cross-validate their findings and adjust their statements where appropriate.

Reviewer #3 (Public review):

Summary:

The authors use high-depth, full-length scRNA-Seq analysis of fetal human retina to identify novel regulators of photoreceptor specification and retinoblastoma progression.

Strengths:

The use of high-depth, full-length scRNA-Seq to identify functionally important alternatively spliced variants of transcription factors controlling photoreceptor subtype specification, and identification of SYK as a potential mediator of RB1-dependent cell cycle reentry in immature cone photoreceptors.

Human developing fetal retinal tissue samples were collected between 13-19 gestational weeks and this provides a substantially higher depth of sequencing coverage, thereby identifying both rare transcripts and alternative splice forms, and thereby representing an important advance over previous droplet-based scRNA-Seq studies of human retinal development.

Weaknesses:

The weaknesses identified are relatively minor. This is a technically strong and thorough study, that is broadly useful to investigators studying retinal development and retinoblastoma.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

Summary:

The authors have used full-length single-cell sequencing on a sorted population of human fetal retina to delineate expression patterns associated with the progression of progenitors to rod and cone photoreceptors. They find that rod and cone precursors contain a mix of rod/cone determinants, with a bias in both amounts and isoform balance likely deciding the ultimate cell fate. Markers of early rod/cone hybrids are clarified, and a gradient of lncRNAs is uncovered in maturing cones. Comparison of early rods and cones exposes an enriched MYCN regulon, as well as expression of SYK, which may contribute to tumor initiation in RB1 deficient cone precursors.

Strengths:

(1) The insight into how cone and rod transcripts are mixed together at first is important and clarifies a long-standing notion in the field.

(2) The discovery of distinct active vs inactive mRNA isoforms for rod and cone determinants is crucial to understanding how cells make the decision to form one or the other cell type. This is only really possible with full-length scRNAseq analysis.

(3) New markers of subpopulations are also uncovered, such as CHRNA1 in rod/cone hybrids that seem to give rise to either rods or cones.

(4) Regulon analyses provide insight into key transcription factor programs linked to rod or cone fates.

(5) The gradient of lncRNAs in maturing cones is novel, and while the functional significance is unclear, it opens up a new line of questioning around photoreceptor maturation.

(6) The finding that SYK mRNA is naturally expressed in cone precursors is novel, as previously it was assumed that SYK expression required epigenetic rewiring in tumors.

We thank the reviewer for describing the study’s strengths, reflecting the major conclusions of the initially submitted manuscript. However, based on new analyses – including the requested analyses of other scRNA-seq datasets, our revision clarifies that:

- related to point (1), cone and rod transcripts do not appear to be mixed together at first (i.e., in immediately post-mitotic immature cone and rod precursors) but appear to be coexpressed in subsequent cone and rod precursor stages; and

- related to point (3), CHRNA1 appears to mark immature cone precursors that are distinct from the maturing cone and rod precursors that co-express cone- and rod-related RNAs (despite the similar UMAP positions of the two populations in our dataset).

Weaknesses:

(1) The writing is very difficult to follow. The nomenclature is confusing and there are contradictory statements that need to be clarified.

(2) The drug data is not enough to conclude that SYK inhibition is sufficient to prevent the division of RB1 null cone precursors. Drugs are never completely specific so validation is critical to make the conclusion drawn in the paper.

We thank the reviewer for noting these important issues. Accordingly, in the revised manuscript:

(1) We improve the writing and clarify the nomenclature and contradictory statements, particularly those noted in the Reviewer’s Recommendations for Authors.

(2) We scale back claims related to the role of SYK in the cone precursor response to RB1 loss, with wording changes in the Abstract, Results, and Discussion, which now recognize that the inhibitor studies only support the possibility that cone-intrinsic SYK expression contributes to retinoblastoma initiation, as detailed in our responses to Reviewer’s Recommendations for Authors. We agree and now mention that genetic perturbation of SYK is required to prove its role.

Reviewer #2 (Public review):

Summary:

The authors used deep full-length single-cell sequencing to study human photoreceptor development, with a particular emphasis on the characteristics of photoreceptors that may contribute to retinoblastoma.

Strengths:

This single-cell study captures gene regulation in photoreceptors across different developmental stages, defining post-mitotic cone and rod populations by highlighting their unique gene expression profiles through analyses such as RNA velocity and SCENIC. By leveraging fulllength sequencing data, the study identifies differentially expressed isoforms of NRL and THRB in L/M cone and rod precursors, illustrating the dynamic gene regulation involved in photoreceptor fate commitment. Additionally, the authors performed high-resolution clustering to explore markers defining developing photoreceptors across the fovea and peripheral retina, particularly characterizing SYK's role in the proliferative response of cones in the RB loss background. The study provides an in-depth analysis of developing human photoreceptors, with the authors conducting thorough analyses using full-length single-cell RNA sequencing. The strength of the study lies in its design, which integrates single-cell full-length RNA-seq, longread RNA-seq, and follow-up histological and functional experiments to provide compelling evidence supporting their conclusions. The model of cell type-dependent splicing for NRL and THRB is particularly intriguing. Moreover, the potential involvement of the SYK and MYC pathways with RB in cone progenitor cells aligns with previous literature, offering additional insights into RB development.

We thank the reviewer for summarizing the main findings and noting the compelling support for the conclusions, the intriguing cell type-dependent splicing of rod and cone lineage factors, and the insights into retinoblastoma development.

Weaknesses:

The manuscript feels somewhat unfocused, with a lack of a strong connection between the analysis of developing photoreceptors, which constitutes the bulk of the manuscript, and the discussion on retinoblastoma. Additionally, given the recent publication of several single-cell studies on the developing human retina, it is important for the authors to cross-validate their findings and adjust their statements where appropriate.

We agree that the manuscript covers a range of topics resulting from the full-length scRNAseq analyses and concur that some studies of developing photoreceptors were not well connected to retinoblastoma. However, we also note that the connection to retinoblastoma is emphasized in several places in the Introduction and throughout the manuscript and was a significant motivation for pursuing the analyses. We suggest that it was valuable to highlight how deep, fulllength scRNA-seq of developing retina provides insights into retinoblastoma, including i) the similar biased expression of NRL transcript isoforms in cone precursors and RB tumors, ii) the cone precursors’ co-expression of rod- and cone-related genes such as NR2E3 and GNAT2, which may explain similar co-expression in RB cells, and iii) the expression of SYK in early cones and RB cells. While the earlier version had mainly highlighted point (iii), the revised Discussion further refers to points (i) and (ii) as described further in the response to the Reviewer’s Recommendations for Authors.

We address the Reviewer’s request to cross-validate our findings with those of other single-cell studies of developing human retina by relating the different photoreceptor-related cell populations identified in our study to those characterized by Zuo et al (PMID 39117640), which was specifically highlighted by the reviewer and is especially useful for such cross-validation given the extraordinarily large ~ 220,000 cell dataset covering a wide range of retinal ages (pcw 8–23) and spatiotemporally stratified by macular or peripheral retina location. Relevant analyses of the Zuo et al dataset are shown in Supplementary Figures S3G-H, S10B, S11A-F, and S13A,B.

Reviewer #3 (Public review):

Summary:

The authors use high-depth, full-length scRNA-Seq analysis of fetal human retina to identify novel regulators of photoreceptor specification and retinoblastoma progression.

Strengths:

The use of high-depth, full-length scRNA-Seq to identify functionally important alternatively spliced variants of transcription factors controlling photoreceptor subtype specification, and identification of SYK as a potential mediator of RB1-dependent cell cycle reentry in immature cone photoreceptors.

Human developing fetal retinal tissue samples were collected between 13-19 gestational weeks and this provides a substantially higher depth of sequencing coverage, thereby identifying both rare transcripts and alternative splice forms, and thereby representing an important advance over previous droplet-based scRNA-Seq studies of human retinal development.

Weaknesses:

The weaknesses identified are relatively minor. This is a technically strong and thorough study, that is broadly useful to investigators studying retinal development and retinoblastoma.

We thank the reviewer for describing the strengths of the study. Our revision addresses the concerns raised separately in the Reviewer’s Recommendations for Authors, as detailed in the responses below.

Recommendations for the authors:

Reviewing Editor Comments:

The reviewers have completed their reviews. Generally, they note that your work is important and that the evidence is generally convincing. The reviewers are in general agreement that the paper adds to the field. The findings of rod/cone fate determination at a very early stage are intriguing. Generally, the paper would benefit from clarifications in the writing and figures. Experimentally, the paper would benefit from validation of the drug data, for example using RNAi or another assay. Alternatively, the authors could note the caveats of the drug experiments and describe how they could be improved. In terms of analysis, the paper would be improved by additional comparisons of the authors' data to previously published datasets.

We thank the reviewing editor for this summary. As described in the individual reviewer responses, we clarify the writing and figures and provide comparisons to previously published datasets (in particular, the large snRNA-seq dataset of Zuo et al., 2024 (PMID 39117640). With regard to the drug (i.e., SYK inhibitor) studies, we opted to provide caveats and describe the need for genetic approaches to validate the role of SYK, owing to the infeasibility of completing genetic perturbation experiments in the appropriate timeframe. We are grateful for the opportunity to present our findings with appropriate caveats.

Reviewer #1 (Recommendations for the authors):

Shayler cell sort human progenitor/rod/cone populations then full-length single cell RNAseq to expose features that distinguish paths towards rods or cones. They initially distinguish progenitors (RPCs), immature photoreceptor precursors (iPRPs), long/medium wavelength (LM) cones, late-LM cones, short wavelength (S) cones, early rods (ER) and late rods (LR), which exhibit distinct transcription factor regulons (Figures 1, 2). These data expose expected and novel enriched genes, and support the notion that S cones are a default state lacking expression of rod (NRL) or cone (THRB) determinants but retaining expression of generic photoreceptor drivers (CRX/OTX2/NEUROD1 regulons). They identify changes in regulon activity, such as increasing NRL activity from iPRP to ER to LR, but decreasing from iPRP to cones, or increasing RAX/ISL2/THRB regulon activity from iPRP to LM cones, but decreasing from iPRP to S cones or rods.

They report co-expression of rod/cone determinants in LM and ER clusters, and the ratios are in the expected directions (NRLTHRB or RXRG in ER). A novel insight from the FL seq is that there are differing variants generated in each cell population. Full-length NRL (FL-NRL) predominates in the rod path, whereas truncated NRL (Tr-NRL) does so in the cone path, then similar (but opposite) findings are presented for THRB (Fig 3, 4), whereas isoforms are not a feature of RXRG expression, just the higher expression in cones.

The authors then further subcluster and perform RNA velocity to uncover decision points in the tree (Figure 5). They identify two photoreceptor precursor streams, the Transitional Rods (TRs) that provide one source for rod maturation and (reusing the name from the initial clustering) iPRPs that form cones, but also provide a second route to rods. TR cells closest to RPCs (immediately post-mitotic) have higher levels of the rod determinant NR2E3 and NRL, whereas the higher resolution iPRPs near RPCs lack NR2E3 and have higher levels of ONECUT1, THRB, and GNAT2, a cone bias. These distinct rod-biased TR and cone-biased high-resolution iPRPs were not evident in published scRNAseq with 3′ end-counting (i.e. not FL seq). Regulon analysis confirmed higher NRL activity in TR cells, with higher THRB activity in highresolution iPRP cells.

Many of the more mature high-resolution iPRPs show combinations of rod (GNAT1, NR2E3) and cone (GNAT2, THRB) paths as well as both NRL and THRB regulons, but with a bias towards cone-ness (Figure 6). Combined FISH/immunofluorescence in fetal retina uncovers cone-biased RXRG-protein-high/NR2E3-protein-absent cone-fated cells that nevertheless expressed NR2E3 mRNA. Thus early cone-biased iPRP cells express rod gene mRNA, implying a rod-cone hybrid in early photoreceptor development. The authors refer to these as "bridge region iPRP cells".

In Figure 7, they identify CHRNA1 as the most specific marker of these bridge cells (overlapping with ATOH7 and DLL3, previously linked to cone-biased precursors), and FISH shows it is expressed in rod-biased NRL protein-positive and cone-biased RXRG proteinpositive cones at fetal week 12.

Figure 8 outlines the graded expression of various lncRNAs during cone maturation, a novel pattern.

Finally (Figure 9), the authors identify differential genes expressed in early rods (ER cluster from Figure 1) vs early cones (LM cluster, excluding the most mature opsin+ cells), revealing high levels of MYCN targets in cones. They also find SYK expression in cones. SYK was previously linked to retinoblastoma, so intrinsic expression may predispose cone precursors to transformation upon RB loss. They finish by showing that a SYK inhibitor blocks the proliferation of dividing RB1 knockdown cone precursors in the human fetal retina.

Overall, the authors have uncovered interesting patterns of biased expression in cone/rod developmental paths, especially relating to the isoform differences for NRL and THRB which add a new layer to our understanding of this fate choice. The analyses also imply that very soon after RPCs exit the cell cycle, they generate post-mitotic precursors biased towards a rod or cone fate, that carry varying proportions of mixed rod/cone determinants and other rod/cone marker genes. They also introduce new markers that may tag key populations of cells that precede the final rod/cone choice (e.g. CHRNA1), catalogue a new lncRNA gradient in cone maturation, and provide insight into potential genes that may contribute to retinoblastoma initiation, like SYK, due to intrinsic expression in cone precursors. However, as detailed below, the text needs to be improved considerably, and overinterpretations need to be moderated, removed, or tested more rigorously with extra data.

Major Comments

The manuscript is very difficult to follow. The nomenclature is at times torturous, and the description of hybrid rod/cone hybrid cells is confusing in many aspects.

(1) A single term, iPRP, is used to refer to an initial low-resolution cluster, and then to a subset of that cluster later in the paper.

We agree that using immature photoreceptor precursor (iPRP) for both high-resolution and lowresolution clusters was confusing. We kept this name for the low-resolution cluster (which includes both immature cone and immature rod precursors), renamed the high-resolution iPRP cluster immature cone precursors (iCPs). and renamed their transitional rod (TR) counterparts immature rod precursors (iRPs). These designations are based on

- the biased expression of THRB, ONECUT1, and the THRB regulon in iCPs (Fig. 5D,E);

- the biased expression of NRL, NR2E3, and NRL regulon iRPs (Fig. 5D,E);

- the partially distinct iCP and iRP UMAP positions (Figure 5C); and

- the evidence of similar immature cone versus rod precursor populations in the Zuo et al 3’ snRNA-seq dataset, as noted below and described in two new paragraphs starting at the bottom of p. 12.

(2) To complicate matters further, the reader needs to understand the subset within the iPRP referred to as bridge cells, and we are told at one point that the earliest iPRPs lack NR2E3, then that they later co-express NR2E3, and while the authors may be referring to protein and RNA, it serves to further confuse an already difficult to follow distinction. I had to read and re-read the iPRP data many times, but it never really became totally clear.

We agree that the description of the high-resolution iPRP (now “iCP”) subsets was unclear, although our further analyses of a large 3’ snRNA-seq dataset in Figure S11 support the impression given in the original manuscript that the earliest iCPs lack NR2E3 and then later coexpress NR2E3 while the earliest iRPs lack THRB and then later express THRB. As described in new text in the Two post-mitotic immature photoreceptor precursor populations section (starting on line 7 of p. 13):

When considering only the main cone and rod precursor UMAP regions, early (pcw 8 – 13) cone precursors expressed THRB and lacked NR2E3 (Figure S11D,E, blue arrows), while early (pcw 10 – 15) rod precursors expressed NR2E3 and lacked THRB (Figure S11D,E, red arrows), similar to RPC-localized iCPs and iRPs in our study (Figure 5D).

Next, as summarized in new text in the Early cone and rod precursors with rod- and conerelated RNA co-expression section (new paragraph at top of p. 16):

Thus, a 3’ snRNA-seq analysis confirmed the initial production of immature photoreceptor precursors with either L/M cone-precursor-specific THRB or rod-precursor-specific NR2E3 expression, followed by lower-level co-expression of their counterparts, NR2E3 in cone precursors and THRB in rod precursors. However, in the Zuo et al. analyses, the co-expression was first observed in well-separated UMAP regions, as opposed to a region that bridges the early cone and early rod populations in our UMAP plots. These findings are consistent with the notion that cone- and rod-related RNA co-expression begins in already fate-determined cone and rod precursors, and that such precursors aberrantly intermixed in our UMAP bridge region due to their insufficient representation in our dataset.

Importantly, and as noted in our ‘Public response’ to Reviewer 1, “CHRNA1 appears to mark immature cone precursors that are distinct from the maturing cone and rod precursors that coexpress cone- and rod-related RNAs (despite the similar UMAP positions of the two populations in our dataset).” In support of this notion, the immature cone precursors expressing CHRNA1 and other populations did not overlap in UMAP space in the Zuo et al dataset. We hope the new text cited above along with other changes will significantly clarify the observations.

(3) The term "cone/rod precursor" shows up late in the paper (page 12), but it was clear (was it not?) much earlier in this manuscript that cone and rod genes are co-expressed because of the coexpressed NRL and THRB isoforms in Figures 3/4.

We thank the reviewer for noting that the differential NRL and THRB isoform expression already implies that cone and rod genes are co-expressed. However, as we now state, the co-expression of RNAs encoding an additional cone marker (GNAT2) and rod markers (GNAT1, NR2E3) was

“suggestive of a proposed hybrid cone/rod precursor state more extensive than implied by the coexpression of different THRB and NRL isoforms” (first paragraph of “Early cone and rod …” section on p. 14; new text underlined).

(4) The (incorrect) impression given later in the manuscript is that the rod/cone transcript mixture applies to just a subset of the iPRP cells, or maybe just the bridge cells (writing is not clear), but actually, neither of those is correct as the more abundant and more mature LM and ER populations analyzed earlier coexpress NRL and THRB mRNAs (Figures 2, 3). Overall, the authors need to vastly improve the writing, simplify/clarify the nomenclature, and better label figures to match the text and help the reader follow more easily and clearly. As it stands, it is, at best, obtuse, and at worst, totally confusing.

We thank the reviewer for bringing the extent of the confusing terminology and wording to our attention. We revised the terminology (as in our response to point 1) and extensively revised the text. We also performed similar analyses of the Zuo et al. data (as described in more detail in our response to Reviewer 2), which clarifies the distinct status of cells with the “rod/cone transcript mixture” and cells co-expressing early cone and rod precursor markers.

To more clearly describe data related to cells with rod- and cone-related RNA co-expression, we divided the former Figure 6 into two figures, with Figure 6 now showing the cone- and rodrelated RNA co-expression inferred from scRNA-seq and Figure 7 showing GNAT2 and NR2E3 co-expression in FISH analyses of human retina plus a new schematic in the new panel 7E.

To separate the conceptually distinct analyses of cone and rod related RNA co-expression and the expression of early photoreceptor precursor markers (which were both found in the so-called bridge region – yet now recognized to be different subpopulations), we separated the analyses of the early photoreceptor precursor markers to form a new section, “Developmental expression of photoreceptor precursor markers and fate determinants,” starting on p. 16.

Additionally, we further review the findings and their implications in four revised Discussion paragraphs starting at the bottom of p. 23).

(5) The data showing that overexpressing Tr-NRL in murine NIH3T3 fibroblasts blocks FL-NRL function is presented at the end of page 7 and in Figure 3G. Subsequent analysis two paragraphs and two figures later (end page 8, Figure 5C + supp figs) reveal that Tr-NRL protein is not detectable in retinoblastoma cells which derive from cone precursors cells and express Tr-NRL mRNA, and the protein is also not detected upon lentiviral expression of Tr-NRL in human fetal retinal explants, suggesting it is unstable or not translated. It would be preferable to have the 3T3 data and retinoblastoma/explant data juxtaposed. E.g. they could present the latter, then show the 3T3 that even if it were expressed (e.g. briefly) it would interfere with FL-NRL. The current order and spacing are somewhat confusing.

We thank the reviewer for this suggestion and moved the description of the luciferase assays to follow the retinoblastoma and explant data and switched the order of Figure panels 3G and 3H.

(6) On page 15, regarding early rod vs early cone gene expression, the authors state: "although MYCN mRNA was not detected....", yet on the volcano plot in Figure S14A MYCN is one of the marked genes that is higher in cones than rods, meaning it was detected, and a couple of sentences later: "Concordantly, the LM cluster had increased MYCN RNA". The text is thus confusing.

With respect, we note that the original text read, “although MYC RNA was not detected,” which related to a statement in the previous sentence that the gene ontology analysis identified “MYC targets.” However, given that this distinction is subtle and may be difficult for readers to recognize, we revised the text (now on p. 19) to more clearly describe expression of MYCN (but not MYC) as follows:

“The upregulation of MYC target genes was of interest given that many MYC target genes are also targets of MYCN, that MYCN protein is highly expressed in maturing (ARR3+) cone precursors but not in NRL+ rods (Figure 10A), and that MYCN is critical to the cone precursor proliferative response to pRB loss8–10. Indeed, whereas MYC RNA was not detected, the LM cone cluster had increased MYCN RNA …”

(7) The authors state that the SYK drug is "highly specific". They provide no evidence, but no drug is 100% specific, and it is possible that off-target hits are important for the drug phenotype. This data should be removed or validated by co-targeting the SYK gene along with RB1.

We agree that our data only show the potential for SYK to contribute to the cone proliferative response; however, we believe the inhibitor study retains value in that a negative result (no effect of the SYK inhibitor) would disprove its potential involvement. To reflect this, we changed wording related to this experiment as follows:

In the Abstract, we changed:

(1) “SYK, which contributed to the early cone precursors’ proliferative response to RB1 loss” To: “SYK, which was implicated in the early cone precursors’ proliferative response to RB1 loss.”

(2) “These findings reveal … and a role for early cone-precursor-intrinsic SYK expression.” To: “These findings reveal … and suggest a role for early cone-precursor-intrinsic SYK expression.”

In the last paragraph of the Results, we changed:

(1) “To determine if SYK contributes…” To: “To determine if SYK might contribute…”

(2) “the highly specific SYK inhibitor” To: “the selective SYK inhibitor”

(3) “indicating that cone precursor intrinsic SYK activity is critical to the proliferative response” To: “consistent with the notion that cone precursor intrinsic SYK activity contributes to the proliferative response.”

In the Results, we added a final sentence:

“However, given potential SYK inhibitor off-target effects, validation of the role of SYK in retinoblastoma initiation will require genetic ablation studies.”

In the Discussion (2nd-to-last paragraph), we changed:

“SYK inhibition impaired pRB-depleted cone precursor cell cycle entry, implying that native SYK expression rather than de novo induction contributes to the cone precursors’ initial proliferation.” To: “…the pRB-depleted cone precursors’ sensitivity to a SYK inhibitor suggests that native SYK expression rather than de novo induction contributes to the cone precursors’ initial proliferation, although genetic ablation of SYK is needed to confirm this notion.” In the Discussion last sentence, we changed:

“enabled the identification of developmental stage-specific cone precursor features that underlie retinoblastoma predisposition.” To: “enabled the identification of developmental stage-specific cone precursor features that are associated with the cone precursors’ predisposition to form retinoblastoma tumors.”

Minor/Typos

Figure 7 legend, H should be D.

We corrected the figure legend (now related to Figure 8).

Reviewer #2 (Recommendations for the authors):

(1) The author should take advantage of recently published human fetal retina data, such as PMID:39117640, which includes a larger dataset of cells that could help validate the findings. Consequently, statements like "To our knowledge, this is the first indication of two immediately post-mitotic photoreceptor precursor populations with cone versus rod-biased gene expression" may need to be revised.

We thank the reviewer for noting the evidence of distinct immediately post-mitotic rod and cone populations published by others after we submitted our manuscript. In response, we omitted the sentence mentioned and extensively cross-checked our results including:

- comparison of our early versus late cone and rod maturation states to the cone and rod precursor versus cone and rod states identified by Zuo et al (new paragraph on the top half of p. 6 and new figure panels S3G,H);

- detection of distinct immediately post-mitotic versus later cone and rod precursor populations (two new paragraphs on pp. 12-13 and new Figures S10B and S11A-E);

- identification of cone and rod precursor populations that co-express cone and rod marker genes (two new paragraphs starting at the bottom of p. 15 and new Figures S11D-F);

- comparison of expression patterns of immature cone precursor (iCP) marker genes in our and the Zuo et al dataset (new paragraph on top half of p. 17 and new Figure S13).

We also compare the cell states discerned in our study and the Zuo et al. study in a new Discussion paragraph (bottom of p. 23) and new Figure S17.

(2) The data generated comes from dissociated cells, which inherently lack spatial context. Additionally, it is unclear whether the dataset represents a pool of retinas from multiple developmental stages, and if so, whether the developmental stage is known for each cell profiled. If this information is available, the authors should examine the distribution of developmental stages on the UMAP and trajectory analysis as part of the quality control process.

We thank the reviewer for highlighting the importance of spatial context and developmental stage.

Related to whether the dataset represents a pool of retinae from multiple developmental stages, the different cell numbers examined at each time point are indicated in Figure S1A. To draw the readers’ attention to this detail, Figure S1A is now cited in the first sentence of the Results.

Related to the age-related cell distributions in UMAP plots, the distribution of cells from each retina and age was (and is) shown in Fig. S1F. In addition, we now highlight the age distributions by segregating the FW13, FW15-17, and FW17-18-19 UMAP positions in the new Figure 1C. We describe the rod temporal changes in a new sentence at the top of p. 5:

“Few rods were detected at FW13, whereas both early and late rods were detected from FW15-19 (Figure 1C), corroborating prior reports [15,20].”

We describe the cone temporal changes and note the likely greater discrimination of cell state changes that would be afforded by separately analyzing macula versus peripheral retina at each age in a new sentence at the bottom of p. 5:

“L/M cone precursors from different age retinae occupied different UMAP regions, suggesting age-related differences in L/M cone precursor maturation (Figure 1C).”

Moreover, they should assess whether different developmental stages impact gene expression and isoform ratios. It is well established that cone and rod progenitors typically emerge at different developmental times and in distinct regions of the retina, with minimal physical overlap. Grouping progenitor cells based solely on their UMAP positioning may lead to an oversimplified interpretation of the data.

(2a) We agree that different developmental stages may impact gene expression and isoform ratios, and evaluated stages primarily based on established Louvain clustering rather than UMAP position. However, we also used UMAP position to segregate so-called RPC-localized and nonRPC-localized iCPs and iRPs, as well as to characterize the bridge region iCP sub-populations. In the revision, we examine whether cell groups defined by UMAP positions helped to identify transcriptomically distinct populations and further examine the spatiotemporal gene expression patterns of the same genes in the Zuo et al. 3’ snRNA-seq dataset.

(2b) Related to analyses of immediately post-mitotic iRPs and iCPs, the new Figure S10A expanded the violin plots first shown in Figure 5D to compare gene expression in RPC-localized versus non-RPC-localized iCPs and iRPs and subsequent cone and rod precursor clusters (also presented in response to Reviewer 3). The new Figure S10C, shows a similar analysis of UMAP region-specific regulon activities. These figures support the idea that there are only subtle UMAP region-related differences in the expression of the selected gene and regulons.

To further evaluate early cone and rod precursors, we compared expression patterns in our cluster- and UMAP-defined cell groups to those of the spatiotemporally defined cell groups in the Zuo et al. 3’ snRNA-seq study. The results revealed similar expression timing of the genes examined, although the cluster assignments of a subset of cells were brought into question, especially the assigned rod precursors at pcw 10 and 13, as shown in new Figures S10B (grey columns) and S11, and as described in two new paragraphs starting near the bottom of p.12.

(2c) Related to analyses of iCPs in the so-called bridge region, our analyses of the Zuo et al dataset helped distinguish early cone and rod precursor populations (expressing early markers such as ATOH7 and CHRNA1) from the later stages exhibiting rod- and cone-related gene coexpression, which had intermixed in the UMAP bridge region in our dataset. Further parsing of early cone precursor marker spatiotemporal expression revealed intriguing differences as now described in the second half of a new paragraph at the top of p. 17, as follows:

“Also, different iCP markers had different spatiotemporal expression: CHRNA1 and ATOH7 were most prominent in peripheral retina with ATOH7 strongest at pcw 10 and CHRNA1 strongest at pcw 13; CTC-378H22.2 was prominently expressed from pcw 10-13 in both the macula and the periphery; and DLL3 and ONECUT1 showed the earliest, strongest, and broadest expression (Figure S13B). The distinct patterns suggest spatiotemporally distinct roles for these factors in cone precursor differentiation.”

(3) I would commend the authors for performing a validation experiment via RNA in situ to validate some of the findings. However, drawing conclusions from analyzing a small number of cells can still be dangerous. Furthermore, it is not entirely clear how the subclustering is done. Some cells change cell type identities in the high-resolution plot. For example, some iPRP cells from the low-resolution plots in Figure 1 are assigned as TR in high-resolution plots in Figure 5.

The authors should provide justification on the identifies of RPC localized iPRP and TR.

Comparison of their data with other publicly available data should strengthen their annotation

We agree that drawing conclusions from scRNA-seq or in situ hybridization analysis of a small number of cells can be dangerous and have followed the reviewer’s suggestion to compare our data with other publicly available data, focusing on the 3’ snRNA-seq of Zuo et al. given its large size and extensive annotation. Our analysis of the Zuo et al. dataset helped clarify cell identities by segregating cone and rod precursors with similar gene expression properties in distinct UMAP regions. However, we noted that the clustering of early cone and rod precursors likely gave numerous mis-assigned cells (as noted in response 2b above and shown in the new Figure S11). It would appear that insights may be derived from the combination of relatively shallow sequencing of a high number of cells and deep sequencing of substantially fewer cells.

Related to how subclustering was done, the Methods state, “A nearest-neighbors graph was constructed from the PCA embedding and clusters were identified using a Louvain algorithm at low and high resolutions (0.4 and 1.6)[70],” citing the Blondel et al reference for the Louvain clustering algorithm used in the Seurat package. To clarify this, the results text was revised such that it now indicates the levels used to cluster at low resolution (0.4, p. 4, 2nd paragraph) and at high resolution (1.6, top of p. 11) .

Related to the assignment of some iPRP cells from the low-resolution plots in Figure 1 to the TR cluster (now called the ‘iRP’ ‘cluster) in the high-resolution plots in Figure 5, we suggest that this is consistent with Louvain clustering, which does not follow a single dendrogram hierarchy.

The justification for referring to these groups as RPC-localized iCPs and iRPs relates to their biased gene and regulon expression in Fig. 5D and 5E, as stated on p. 12:

“In the RPC-localized region, iCPs had higher ONECUT1, THRB, and GNAT2, whereas iRPs trended towards higher NRL and NR2E3 (p= 0.19, p=0.054, respectively).”

(4) Late-stage LM5 cluster Figure 9 is not defined anywhere in previous figures, in which LM clusters only range from 1 to 4. The inconsistency in cluster identification should be addressed.

We revised the text related to this as follows:

“Indeed, our scRNA-seq analyses revealed that SYK RNA expression increased from the iCP stage through cluster LM4, in contrast to its minimal expression in rods (Figure 10E). Moreover, SYK expression was abolished in the five-cell group with properties of late maturing cones (characterized in Figure 1E), here displayed separately from the other LM4 cells and designated LM5 (Figure 10E).” (p. 19-20)

(5) Syk inhibitor has been shown to be involved in RB cell survival in previous studies. The manuscript seems to abruptly make the connection between the single-cell data to RB in the last figure. The title and abstract should not distract from the bulk of the manuscript focusing on the rod and cone development, or the manuscript should make more connection to retinoblastoma.

We appreciate the reviewer’s concern that the title may seem to over-emphasize the connection to retinoblastoma based solely on the SYK inhibitor studies. However, we suggest the title also emphasizes the identification and characterization of early human photoreceptor states, per se, and that there are a number of important connections beyond the SYK studies that could warrant the mention of cell-state-specific retinoblastoma-related features in the title.

Most importantly, a prior concern with the cone cell-of-origin theory was that retinoblastoma cells express RNAs thought to mark retinal cell types other than cones, especially rods. The evidence presented here, that cone precursors also express the rod-related genes helps resolve this issue. The issue is noted numerous times in the manuscript, as follows:

In the Introduction, we write:

“However, retinoblastoma cells also express rod lineage factor NRL RNAs, which – along with other evidence – suggested a heretofore unexplained connection between rod gene expression and retinoblastoma development[12,13]. Improved discrimination of early photoreceptor states is needed to determine if co-expression of rod- and cone-related genes is adopted during tumorigenesis or reflects the co-expression of such genes in the retinoblastoma cell of origin.” (bottom, p. 2) And:

“In this study, we sought to further define the transcriptomic underpinnings of human photoreceptor development and their relationship to retinoblastoma tumorigenesis.” (last paragraph, p. 3)

The Discussion also alluded to this issue and in the revised Discussion, we aimed to make the connection clearer. We previously ended the 3rd-to-last paragraph with,

“iPRP [now iCP] and early LM cone precursors’ expression of NR2E3 and NRL RNAs suggest that their presence in retinoblastomas[12,13] reflects their normal expression in the L/M cone precursor cells of origin.”

We now separate and elaborate on this point in a new paragraph as follows:

“Our characterization of cone and rod-related RNA co-expression may help resolve questions about the retinoblastoma cell of origin. Past studies suggested that retinoblastoma cells co-express RNAs associated with rods, cones, or other retinal cells due to a loss of lineage fidelity[12]. However, the early L/M cone precursors’ expression of NR2E3 and NRL RNAs suggest that their presence in retinoblastomas[12,13] reflects their normal expression in the L/M cone precursor cells of origin. This idea is further supported by the retinoblastoma cells’ preferential expression of cone-enriched NRL transcript isoforms (Figure S5B).” (middle of p. 24) Based on the above, we elected to retain the title.

Minor comments:

(1) It is difficult to see the orange and magenta colors in the Fig 3E RNA-FISH image. The colors should be changed, or the contrast threshold needs to be adjusted to make the puncta stand out more.

We re-assigned colors, with red for FL-NRL puncta and green for Tr-NRL puncta.

(2) Figure 5C on page 8 should be corrected to Supplementary Figure 5C.

We thank the reviewer for noting this error and changed the figure citation.

Reviewer #3 (Recommendations for the authors):

(1) Minor concerns

a. Abbreviation of some words needs to be included, example: FW.

We now provide abbreviation definitions for FW and others throughout the manuscript.

b. Cat # does not matches with the 'key resource table' for many reagents/kits. Some examples are: CD133-PE mentioned on Page # 22 on # 71, SMART-Seq V4 Ultra Low Input RNA Kit and SMARTer Ultra Low RNA Kit for the Fluidigm C1 Sytem on Page # 22 on # 77, Nextera XT DNA Library preparation kit on Page # 23 on # 77.

We thank the reviewer for noting these discrepancies. We have now checked all catalog numbers and made corrections as needed.

c. Cat # and brand name of few reagents & kits is missing and not mentioned either in methods or in key resource table or both. Eg: FBS, Insulin, Glutamine, Penicillin, Streptomycin, HBSS, Quant-iT PicoGreen dsDNA assay, Nextera XT DNA LibraryPreparation Kit, 5' PCR Primer II A with CloneAmp HiFi PCR Premix.

Catalog numbers and brand names are now provided for the tissue culture and related reagents within the methods text and for kits in the Key Resources Table. Additional descriptions of the primers used for re-amplification and RACE were added to the Methods (p. 28-29).

d. Spell and grammar check is needed throughout the manuscript is needed. Example. In Page # 46 RXRγlo is misspelled as RXRlo.

Spelling and grammar checks were reviewed.

(2) Methods & Key Resource table.

a. In Page # 21, IRB# needs to be stated.

The IRB protocols have been added, now at top of p. 26.

b. In Page # 21, Did the authors dissociate retinae in ice-cold phosphate-buffered saline or papain?

The relevant sentence was corrected to “dissected while submerged in ice-cold phosphatebuffered saline (PBS) and dissociated as described10.” ( p. 26)

c. In Page # 21, How did the authors count or enumerate the cell count? Provide the details.

We now state, “… a 10 µl volume was combined with 10 µl trypan blue and counted using a hemocytometer” (top of p. 27)

d. Why did the authors choose to specifically use only 8 cells for cDNA preparation in Page # 22? State the reason and provide the details.

The reasons for using 8 cells (to prevent evaporation and to manually transfer one slide-worth of droplets to one strip of PCR tubes) and additional single cell collection details are now provided as follows (new text underlined):

“Single cells were sorted on a BD FACSAria I at 4°C using 100 µm nozzle in single-cell mode into each of eight 1.2 µl lysis buffer droplets on parafilm-covered glass slides, with droplets positioned over pre-defined marks … . Upon collection of eight cells per slide, droplets were transferred to individual low-retention PCR tubes (eight tubes per strip) (Bioplastics K69901, B57801) pre-cooled on ice to minimize evaporation. The process was repeated with a fresh piece of parafilm for up to 12 rounds to collect 96 cells). (p. 27, new text underlined)

e. Key resource table does not include several resources used in this study. Example - NR2E3 antibody.

We added the NR2E3 antibody and checked for other omissions.

(3) Results & Figures & Figure Legends

a. Regulon-defined RPC and photoreceptor precursor states

i. On page # 4, 1 paragraph - Clarify the sentence 'Exclusion of all cells with <100,000 cells read and 18 cells.........Emsembl transcripts inferred'. Did the authors use 18 cells or 18FW retinae?

The sentence was changed to:

“After sequencing, we excluded all cells with <100,000 read counts and 18 cells expressing one or more markers of retinal ganglion, amacrine, and/or horizontal cells (POU4F1, POU4F2, POU4F3, TFAP2A, TFAP2B, ISL1) and concurrently lacking photoreceptor lineage marker OTX2. This yielded 794 single cells with averages of 3,750,417 uniquely aligned reads, 8,278 genes detected, and 20,343 Ensembl transcripts inferred (Figure S1A-C).” (p. 4, new words underlined)

To clarify that 18 retinae were used, the first sentence of the Results was revised as follows:

“To interrogate transcriptomic changes during human photoreceptor development, dissociated RPCs and photoreceptor precursors were FACS-enriched from 18 retinae, ages FW13-19 …” (p. 4).

Why did the authors 'exclude cells lacking photoreceptor lineage marker OTX2' from analysis especially when the purpose here was to choose photoreceptor precursor states & further results in the next paragraph clearly state that 5 clusters were comprised of cells with OTX2 and CRX expression. This is confusing.

We apologize for the imprecise diction. We divided the evidently confusing sentence into two sentences to more clearly indicate that we removed cells that did not express OTX2, as in the first response to the previous question.

ii. In Page # 5, the authors reported the number of cell populations (363 large and 5 distal) identified in the THRB+ L/M-cone cluster. What were the # of cell populations identified in the remaining 5 clusters of the UMAP space?

We added the cell numbers in each group to Fig. 1B. We corrected the large LM group to 366 cells (p. 5) and note 371 LM cells , which includes the five distal cells, in Figure 1B.

b. Differential expression of NRL and THRB isoforms in rod and cone precursors

i. In Figure 3B, the authors compare and show the presence of 5 different NRL isoforms for all the 6 clusters that were defined in 3A. However, in the results, the ENST# of just 2 highly assigned transcript isoforms is given. What are the annotated names of the three other isoforms which are shown in 3B? Please explain in the Results.

As requested, we now annotate the remaining isoforms as encoding full-length or truncated NRL in Fig. 3B and show isoform structures in new Supplementary Figure S4B. We also refer to each transcript isoform in the Results (p. 7, last paragraph) and similarly evaluate all isoforms in RB31 cells (Fig. S5B).

ii. What does the Mean FPM in the y-axis of Fig 3C refer to?

Mean FPM represents mean read counts (fragments per million, FPM) for each position across Ensembl NRL exons for each cluster, as now stated in the 6th line of the Fig. 3 legend.

iii. A clear explanation of the results for Figures 3E-3F is missing.

We revised the text to more clearly describe the experiment as follows:

“The cone cells’ higher proportional expression of Tr-NRL first exon sequences was validated by RNA fluorescence in situ hybridization (FISH) of FW16 fetal retina in which NRL immunofluorescence was used to identify rod precursors, RXRg immunofluorescence was used to identify cone precursors, and FISH probes specific to truncated Tr-NRL exon 1T or FL-NRL exons 1 and 2 were used to assess Tr-NRL and FL-NRL expression (Figure 3E,F).” (p. 8, new text underlined).

c. Two post-mitotic photoreceptor precursor populations

i. Although deep-sequencing and SCENIC analysis clarified the identities of four RPC-localized clusters as MG, RPC, and iPRP indicative of cone-bias and TR indicative of rod-bias. It would be interesting to see the discriminating determinant between the TR and ER by SCENIC and deep-sequencing gene expression violin/box plots.

We agree it is of interest to see the discriminating determinant between the TR [now termed iRP] and ER clusters by SCENIC and deep-sequencing gene expression violin/box plots. We now provide this information for selected genes and regulons of interest in the new Supplementary Figures S10A and S10C, along with a similar comparison between the prior high-resolution iPRP (now termed iCP) cluster and the first high-resolution LM cluster, LM1, as described for gene expression on p. 12:

“Notably, THRB and GNAT2 expression did not significantly change while ONECUT1 declined in the subsequent non-RPC-localized iCP and LM1 stages, whereas NR2E3 and NRL dramatically increased on transitioning to the ER state (Figure S10A).”

And as described for regulon activities on pp. 13-14:

“Finally, activities of the cone-specific THRB and ISL2 regulons, the rod-specific NRL regulon, and the pan-photoreceptor LHX3, OTX2, CRX, and NEUROD1 regulons increased to varying extents on transitioning from the immature iCP or iRP states to the early-maturing LM1 or ER states (Figure 10C).”

We also show expression of the same genes for spatiotemporally grouped cells from the Zuo et al. dataset in the new Figure S10B, which displays a similar pattern (apart from the possibly mixed pcw 10 and pcw13 designated rod precursors).

d. Early cone precursors with cone- and rod-related RNA expression

i. On page #12, the last paragraph where the authors explain the multiplex RNA FISH results of RXRγ and NR2E3 by citing Figure S8E. However, in Fig S8E, the authors used NRL to identify the rods. Please clarify which one of the rod markers was used to perform RNA FISH?

Figure S8E (where NRL was used as a rod marker) was cited to remind readers that RXRg has low expression in rods and high expression in cones, rather than to describe the results of this multiplex FISH section. To avoid confusion on this point, Figure S8E is now cited using “(as earlier shown in Figure S8E).” With this issue clarified, we expect the markers used in the FISH + IF analysis will be clear from the revised explanation,

“… we examined GNAT2 and NR2E3 RNA co-expression in RXRg+ cone precursors in the outermost NBL and in RXRg+ rod precursors in the middle NBL … .” (p. 14-15).

To provide further clarity, we provide a diagram of the FISH probes, protein markers, and expression patterns in the new Figure 7E.

ii. The Y-axis of Fig 6G-6H needs to be labelled.

The axes have been re-labeled from “Nb of cells” to “Number of RXRg+ outermost NBL cells in each region” (original Fig. 6G, now Fig. 7C) and “Number of RXRg+ middle NBL cells in each region” (original Fig. 6H, now Fig. 7D).

iii. The legends of Figures 6G and 6H are unclear. In the Figure 6G legend, the authors indicate 'all cells are NR2E3 protein-'. Does that imply the yellow and green bars alone? Similarly, clarify the Figure 6H legend, what does the dark and light magenta refer to? What does the light magenta color referring to NR2E3+/ NR2E3- and the dark magenta color referring to NR2E3+/ NR2E3+ indicate?

We regret the insufficient clarity. We revised the Fig. 6G (now Fig. 7C) key, which now reads

“All outermost NBL cells are NR2E3 protein-negative.” We added to the figure legend for panel 7C,D “(n.b., italics are used for RNAs, non-italics for proteins).” The new scheme in Figure 7E shows the RNAs in italics proteins in non-italics. We hope these changes will clarify when RNA or protein are represented in each histogram category.

Overall, the results (on page # 13) reflecting Figures 6E-6H & Figure S11 are confusing and difficult to understand. Clear descriptions and explanations are needed.

We revised this results section described in the paragraph now spanning p. 14:

- We now refer to the bar colors in Figures 7C and 7D that support each statement.

- We provide an illustration of the findings in Figure 7E.

iv. Previously published literature has shown that cells of the inner NBL are RXRγ+ ganglion cells. So, how were these RXRγ+ ganglion cells in the inner NBL discriminated during multiplex RNA FISH (in Fig 6E-6H and in Fig S11)?

We thank the reviewer for requesting this clarification. We agree that “inner NBL” is the incorrect term for the region in which we examined RXRg+ photoreceptor precursors, as this could include RXRγ+ nascent RGCs. We now clarify that

“we examined GNAT2 and NR2E3 RNA co-expression in RXRg+ cone precursors in the outermost NBL and in RXRg+ rod precursors in the middle NBL … .” (p. 14-15) We further state,

“Limiting our analysis to the outer and middle NBL allowed us to disregard RXRγ+ retinal ganglion cells in the retinal ganglion cell layer or inner NBL (top of p. 15)”

Figure 7E is provided to further aid the reader in understanding the positions examined, and the legend states “RXRg+ retinal ganglion cells in the inner NBL and ganglion cell layer not shown.

v. In Figure 6E, what marker does each color cell correspond to?

In this figure (now panel 7A), we declined to provide the color key since the image is not sufficiently enlarged to visualize the IF and FISH signals. The figure is provided solely to document the regions analyzed and readers are now referred to “see Figure S12 for IF + FISH images” (2nd line, p. 15), where the marker colors are indicated.

vi. In Figure S11 & 6E, Protein and RNA transcript color of NR2E3, GNAT2 are hard to distinguish. Usage of other colors is recommended.

We appreciate the reviewer’s concern related to the colors (in the now redesignated Figure S12 and 7A); however, we feel this issue is largely mitigated by our use of arrows to point to the cells needed to illustrate the proposed concepts in Figure S12B. All quantitation was performed by examining each color channel separately to ensure correct attribution, which is now mentioned in the Methods (2nd-to-last line of Quantitation of FISH section, p. 35).

vii.

With due respect, we suggest that labeling each box (now in Figure 8B) makes the figure rather busy and difficult to infer the main point, which is that boxed regions were examined at various distanced from the center (denoted by the “C” and “0 mm”) with distances periodically indicated. We suggest the addition of such markers would not improve and might worsen the figure for most readers.

e. An early L/M cone trajectory marked by successive lncRNA expression

i. In Figure 8C - color-coded labelling of LM1-4 clusters is recommended.

We note Fig. 8C (now 9C) is intended to use color to display the pseudotemporal positions of each cell. We recognize that an additional plot with the pseudotime line imposed on LM subcluster colors could provide some insights, yet we are unaware of available software for this and are unable to develop such software at present. To enable readers to obtain a visual impression of the pseudotime vs subcluster positions, we now refer the reader to Figure 5A in the revised figure legend, as follows: (“The pseudotime trajectory may be related to LM1-LM4 subcluster distributions in Figure 5A.”).

ii. In Figure 8G - what does the horizontal color-coded bar below the lncRNAs name refer to? These bars are similar in all four graphs of the 8G figure.

As stated in the Fig. 8G (now 9G) legend, “Colored bars mark lncRNA expression regions as described in the text.” We revised the text to more clearly identify the color code. (p. 18-19)

f. Cone intrinsic SYK contributions to the proliferative response to pRB loss

i. In Fig 9F - The expression of ARR3+ cells (indicated by the green arrow in FW18) is poorly or rarely seen in the peripheral retina.

We thank the reviewer for finding this oversight. In panel 9F (now 10F), we removed the green arrows from the cells in the periphery, which are ARR3- due to the immaturity of cones in this region.

ii. In Figure 9F - Did the authors stain the FW16 retina with ARR3?

Unfortunately, we did not stain the FW16 retina for ARR3 in this instance.

iii. Inclusion of DAPI staining for Fig 9F is recommended to justify the ONL & INL in the images.

We regret that we are unable to merge the DAPI in this instance due to the way in which the original staining was imaged. A more detailed analysis corroborating and extending the current results is in progress.

iv. Immunostaining images for Figure 9G are missing & are required to be included. What does shSCR in Fig 9G refer to?

We now provide representative immunostaining images below the panel (now 10G). The legend was updated: “Bottom: Example of Ki67, YFP, and RXRg co-immunostaining with DAPI+ nuclei (yellow outlines). Arrows: Ki67+, YFP+, RXRg+ nuclei.” The revised legend now notes that shSCR refers to the scrambled control shRNA.

v. For Figure 9H - Is the presence and loss of SYK activity consistent with all the subpopulations (S & LM) of early maturing and matured cones?

We appreciate the reviewer’s question and interest (relating to the redesignated Figure 10H); however, we have not yet completed a comprehensive evaluation of SYK expression in all the subpopulations (S & LM) of early maturing and matured cones and will reserve such data for a subsequent study. We suggest that this information is not critical to the study’s major conclusions.

vi. Figure 9A is not explained in the results. Why were MYCN proteins assessed along with ARR3 and NRL? What does this imply?

We thank the reviewer for noting that this figure (now Figure 10A) was not clearly described.

As per the response to Reviewer 1, point 6 , the text now states,

“The upregulation of MYC target genes was of interest given that many MYC target genes are also MYCN targets, that MYCN protein is highly expressed in maturing (ARR3+) cone precursors but not in NRL+ rods (Figure 10A), and that MYCN is critical to the cone precursor proliferative response to pRB loss [8–10].” (middle, p. 19, new text underlined).

Hence, the figure demonstrates the cone cell specificity of high MYCN protein. This is further noted in the Fig. 10a legend: “A. Immunofluorescent staining shows high MYCN in ARR3+ cones but not in NRL+ rods in FW18 retina.”

Version 2

Abstract

Abstract

Human cone photoreceptors differ from rods and serve as the retinoblastoma cell-of-origin, yet the developmental basis for their distinct behaviors is poorly understood. Here, we used deep full-length single-cell RNA-sequencing (scRNA-seq) to distinguish post-mitotic cone and rod developmental states and identify cone-specific features that contribute to retinoblastomagenesis. The analyses revealed nascent, immediately post-mitotic cone- and rod precursor populations characterized by higher THRB or NRL regulon activities, immature and maturing cone and rod precursors with concurrent cone- and rod-related gene and regulon expression, and distinct early and late cone and rod maturation states distinguished by maturation-associated declines in RAX regulon activity. Cell-state-specific gene expression features inferred based on full-length scRNA-seq were consistent with past single nucleus 3’ RNA-seq analyses. Beyond the cell state characterizations, full-length scRNAseq revealed that both L/M cone and rod precursors co-expressed NRL and THRB RNAs, yet they differentially expressed functionally antagonistic NRL and THRB isoforms and prematurely terminated THRB transcripts. Moreover, early L/M cone precursors exhibited successive expression of several lncRNAs along with MYCN, which composed the seventh most L/M-cone-specific regulon, and SYK, which was implicated in the cone precursors’ proliferative response to RB1 loss. These findings reveal previously unresolved photoreceptor precursor states and suggest a role for early cone-precursor-intrinsic SYK expression in retinoblastoma initiation.

Introduction

Vertebrate photoreceptors develop from optic vesicle retinal progenitor cells (RPCs) through progressive RPC lineage restriction, fate determination, and post-mitotic maturation1,2. While several transcription factors that govern these events have been identified, important aspects of photoreceptor fate determination and maturation remain unclear. For example, it is unclear if fate is determined in RPCs, where OTX2 and ONECUT1 are thought to control the post-mitotic expression of long- or medium-wavelength (L/M) cone determinant TRβ2 and rod determinant NRL3, or is determined in post-mitotic photoreceptor precursors with concurrent TRβ2 and NRL expression 4. Following fate commitment, post-mitotic developmental stages have been defined based on morphologic features and phototransduction-related gene or protein expression5,6, but it is unclear if progression through such stages is subdivided into distinct cell states governed by unique transcription factor combinations or represents a developmental continuum.

An improved understanding of photoreceptor development may provide insight into the pathogenesis of retinal dystrophies, retinal degenerations, and the retinal cone precursor cancer, retinoblastoma7,8. In the latter case, L/M cone precursors lacking the retinoblastoma protein (pRB) were shown to proliferate in a manner dependent on the L/M-cone lineage factors RXRγ and TRβ2 and the intrinsically highly expressed MDM2 and MYCN oncoproteins, likely representing the first step of retinoblastoma tumorigenesis8–11. Similarly, retinoblastoma cell proliferation depends on RXRγ, TRβ2, MDM2, and MYCN9, implying that intrinsic L/M-cone factors contribute to the oncogenic state. However, retinoblastoma cells also express rod lineage factor NRL RNAs, which – along with other evidence – suggested a heretofore unexplained connection between rod gene expression and retinoblastoma development12,13. Improved discrimination of early photoreceptor states is needed to determine if co-expression of rod- and cone-related genes is adopted during tumorigenesis or reflects the co-expression of such genes in the retinoblastoma cell of origin.

The cone precursors’ propensity to form retinoblastoma is a human-specific feature whose study requires analysis of developing human retina8. Single-cell RNA-sequencing (scRNA-seq) is well suited to such analyses as it enables discrimination of cell type-specific fate determination and maturation features. scRNA-seq studies employing 3’ end-counting have defined age-related post-mitotic transition populations, fate-determining features of post-mitotic photoreceptor precursors, and gene expression changes associated with the cone fate decision and early development14–20. However, 3’ end-counting cannot be used to interrogate transcript isoforms, and the relatively low number of genes detected per cell limits the ability to distinguish closely related states in individual cells.

In this study, we sought to further define the transcriptomic underpinnings of human photoreceptor development and their relationship to retinoblastoma tumorigenesis. We generated deep, full-length single-cell transcriptomes of human retinal progenitor cells (RPCs) and developing photoreceptors from fetal week (FW) 13-19 retinae, with enrichment of rare cone precursor populations, and applied long-read cDNA sequencing, RNA velocity, pseudotemporal trajectory reconstruction, and single-cell regulatory network inference and clustering (SCENIC) to interrogate individual cell states. These analyses discriminated previously unresolved photoreceptor developmental states, identified photoreceptor precursor states with cone and rod-related RNA co-expression, uncovered cell type-specific expression of RNA isoforms of photoreceptor fate-determining genes, elucidated post-mitotic photoreceptor developmental trajectories, and revealed retinoblastoma cell-of-origin features that may contribute to retinoblastoma genesis.

Results

Regulon-defined RPC and photoreceptor precursor states

To interrogate transcriptomic changes during human photoreceptor development, dissociated RPCs and photoreceptor precursors were FACS-enriched from 18 retinae, ages FW13-19 (Figure S1A), and isolated using microfluidics or direct sorting into microliter droplets, followed by full-length cDNA synthesis, paired-end sequencing, and alignment to Ensembl transcript isoforms (Figure 1A). The FACS enrichment was based on a prior cone isolation method10 but with wider gating to include rods and RPCs. After sequencing, we excluded all cells with <100,000 read counts and 18 cells expressing one or more markers of retinal ganglion, amacrine, and/or horizontal cells (POU4F1, POU4F2, POU4F3, TFAP2A, TFAP2B, ISL1) and concurrently lacking photoreceptor lineage marker OTX2. This yielded 794 single cells with averages of 3,750,417 uniquely aligned reads, 8,278 genes detected, and 20,343 Ensembl transcripts inferred (Figure S1A-C). Sequencing batches were normalized and transcriptomes clustered and visualized in uniform manifold approximation and projection (UMAP) plots that integrated cells across different retinae, ages, isolation methods, and sequencing runs (Figures 1B,C, S1D-F).

Photoreceptor-enriched full-length scRNA-seq of developing human retina.

A. Overview of sample collection and sequencing. B, C. UMAP plots showing low-resolution cell type clusters (B) and ages (C). D. Expression of marker genes for RPC/MGs (LHX2), rods (NR2E3), S cones (OPN1SW), L/M cones (THRB). Insets: Gene expression violin plots (from left to right): RPC/MG (red), iPRP (brown), LM cone (green), S cone (teal), early rod (blue), late rod (pink). E. Expression of markers of rod maturation (PDE6G, RHO) and cone maturation (PDE6H, OPN1LW). Arrowheads: Late-maturing RHO+ rods (top), late-maturing OPN1LW+ cones (bottom). See Figure S2 for additional examples. UMAP and violin plots for any gene or transcript isoform can be produced at https://docker.saban.chla.usc.edu/cobrinik/app/seuratApp/.

Low resolution Louvain clustering (level 0.4) generated six clusters that segregated into mostly distinct UMAP domains (Figure 1B). One cluster was comprised of RPCs and Mϋller glia (MG), with specific expression of LHX2, VSX2, SOX2, and SLC1A3, while five clusters were comprised of cells with photoreceptor features, with wide expression of OTX2 and CRX and cluster-specific rod- and cone gene expression (Figures 1D, S2). In UMAP space, the cluster designated immature photoreceptor precursors (iPRPs) intermixed with the RPC/MG population, extended towards early L/M cone precursors, and was predominantly comprised of cells expressing the L/M cone determinant THRB (Figure 1B,D).

Two clusters highly expressing the rod determinant NR2E3 were designated early-maturing rod (ER) and late-maturing rod (LR) (Figure 1B), based on the latter’s increased expression of rod phototransduction genes GNAT1, CNGB1, PDE6G, GNGT1 and RHO (Figures 1D, S2C). Few rods were detected at FW13, whereas both early and late rods were detected from FW15-19 (Figure 1C), corroborating prior reports15,20. Differential expression analysis revealed other genes upregulated in late rods (GNB1, SAMD7, NT5E)21,22 as well as the downregulated CRABP2, DCT, and FABP7 (Figure S3A, Table S1). Genes upregulated in the LR cluster were enriched for photoreceptor and light sensing gene ontologies (Figure S3B) including spectrin binding, likely relating to proteins that control photoreceptor polarity and synapse formation23,24, and purine biosynthesis and ribonucleotide metabolic processes, potentially related to the developing photoreceptors’ high NAD+ requirements25.

Cones segregated into distinct S- and L/M-cone clusters, with differential expression of cone subtype markers (OPN1SW, THRB), previously identified S-cone enriched genes (CCDC136, UPB1)26–28, novel S cone enriched genes (MEGF10, NRXN3, ACKR3), and the L/M cone transcription factor ISL215, among others (Figure 1B,D, S3C, Table S2). Gene ontology analysis did not reveal relevant terms enriched in S cones, whereas L/M cones were enriched for protein translation related ontologies (Figure S3D) due to increased expression of ribosomal protein genes RPL23A, RPLP0, RPS19, RPS27, RPS27A, RPS29, RPS3A (Table S2).

In UMAP space, the THRB+ L/M-cone cluster segregated into a large 366-cell proximal population and a 5-cell distal population inferred to represent early-maturing and late-maturing stages, respectively, based on the latter’s increased expression of cone phototransduction genes OPN1LW (encoding L-opsin), PDE6H, and GUCA1C (Figure 1E, S3E,F), analogous to RHO, PDE6G, and GNGT1 upregulation in late rods. L/M cone precursors from different age retinae occupied different UMAP regions, suggesting age-related differences in L/M cone precursor maturation (Figure 1C). Compared to the early L/M population, late L/M cones had upregulation of three M-opsin genes, TTR, encoding the retinol-binding protein transthyretin, PCP4, a small protein that binds calmodulin previously noted in foveal cones29, and MYL4, a myosin light chain gene upregulated in retinal organoid L/M cones28, among others (Figure S3E,F, Table S3). The low proportions of OPN1MW+ and OPN1LW+ late-maturing L/M cones is consistent with a prior analysis of similar-age retinae and with the further upregulation of these proteins in later maturation15.

We next asked whether similar distinctions between early-maturing and late-maturing L/M cone and rod precursors were observed in prior studies. Indeed, a 3’ snRNA-seq analysis of ∼ 220,000 retinal cells from post-conception week (pcw) 8 – 23 distinguished the cone precursor versus ML cone and the rod precursor versus rod subclasses (Zuo et al.20). (N.b., we retain the pcw and ML cone terms of Zuo et al. when describing their data and the synonymous FW and L/M cone for our data to maintain continuity with past publications.) The Zuo et al. cone and rod precursor versus cone and rod photoreceptor comparisons were not strictly analogous to our early-maturing versus late-maturing precursor comparisons in that our early-maturing precursors excluded immature iPRP cells. Still, the comparisons revealed many of the same differentially expressed genes (Figures S3A versus S3G and S3E versus S3H; Tables S1, S3-S5).

To further interrogate cell identities, we used SCENIC to identify cluster-specific transcription factor regulons, which represent the overall expression of single transcription factors and their likely coregulated target genes30. The highest specificity regulons defined major cell populations including RPC/MG-specific E2F2, E2F3, VSX2, and PAX6; pan-photoreceptor NEUROD1, OTX2, and CRX; rod-specific NRL; and L/M cone-specific THRB and ISL2 (Figure 2, Table S6). SCENIC also distinguished ER and LR states via the latter’s increased NRL, CRX, ATF4, and LHX3 and decreased HMX1 and RAX activities (p <0.0005 for each, Dunn test). RAX activity also decreased in the 5-cell late-maturing L/M cone group (Figure 2B), supporting its distinct transcriptomic identity and suggesting a similar mode of late cone and late rod maturation. Additionally, iPRPs expressed their most specific regulons, LHX9 and OLIG2, at levels similar to RPC/MGs along with photoreceptor-related regulons, consistent with the transitional nature of this population (Figure 2A, Table S6). However, SCENIC did not identify S cone-specific regulons, in keeping with the notion that S cones represent a default photoreceptor state induced by pan-photoreceptor factors such as OTX2, CRX, and NEUROD1 in the absence of NRL and THRB31. Thus, deep, full-length scRNA-seq enabled identification of regulons underlying RPC and developing photoreceptor states at the single cell level.

Regulon-defined RPC and photoreceptor precursor states

A. Ward-clustered heatmap of the highest scoring SCENIC regulons in each cluster, displaying Z-score normalized regulon activities. Late = late-maturing L/M cones. B. Box plot of RAX regulon area under the curve (AUC) values for early and late L/M cones and rods. *, p<0.005; ***, p<0.0005, Dunn test. C,D. UMAP plots of regulon AUC values for (C) PAX6 (RPC/MG) and E2F2 (RPC), and (D) OTX2 (photoreceptors and photoreceptor-committed RPCs), NRL (rod) and THRB and ISL2 (L/M cone).

Differential expression of NRL and THRB isoforms in rod and cone precursors

Although cone and rod precursors segregated into distinct clusters, mRNAs encoding rod-determining factor NRL, L/M cone-determining factor TRβ, and cone marker RXRγ were co-expressed in both rod and cone precursor populations, with mean NRL expression only 4.3-fold higher in the ER vs LM cluster, mean THRB expression 5.0-fold higher in LM vs LR, and mean RXRG expression 4.3-fold higher in LM vs ER (Figure S2C,D and below). Cone NRL expression was unexpected given NRL’s role in rod fate determination28 and rod-specific NRL regulon activity (Figure 2A,D). Similarly, rod RXRG and THRB expression were unexpected given their roles in cone gene expression and fate determination32,33 and L/M cone-specific THRB regulon activity (Figure 2A,D). Accordingly, we used full-length scRNA-seq data to determine if cone and rod precursors differentially express NRL, THRB, and RXRG transcript isoforms.

For NRL, three assigned transcript isoforms (ENST00000397002, ENST00000561028, and ENST00000558280) are predicted to encode the canonical full-length NRL protein (FL-NRL) (RefSeq NP_001341697.1), while two others (ENST00000560550 and ENST00000396995) are previously uncharacterized transcripts predicted to use an alternative “P2” promoter and first exon, here termed exon 1T (Figure 3A-C). The novel transcripts are predicted to encode an N-terminally truncated NRL protein (Tr-NRL) retaining the leucine zipper DNA binding domain but lacking the minimal transactivation domain34 (Figure 3C, Figure S4A,B). While all transcript isoforms were inferred to be more highly expressed in the ER rod cluster versus the LM cone cluster, the ratio of all FL-NRL:Tr-NRL transcripts was 2.9:1 in early rod precursors and 2.2:1 in late rod precursors, in contrast to 0.67:1 in L/M cone precursors (Figure 3B). Consistent with the assigned isoform ratios, mean read coverage of the Tr-NRL-specific exon 1T was higher in S and LM cones while coverage of FL-NRL-specific exon 1 was higher in rods (Figure 3C, red vs. black arrowheads). Comparing the reads mapped to each first exon relative to total reads further confirmed that the Tr-NRL exon 1T predominated in individual cones whereas the FL-NRL exon 1 predominated in rods (Figure 3D). The cone cells’ higher proportional expression of Tr-NRL first exon sequences was validated by RNA fluorescence in situ hybridization (FISH) of FW16 fetal retina in which NRL immunofluorescence was used to identify rod precursors, RXRγ immunofluorescence was used to identify cone precursors, and FISH probes specific to Tr-NRL exon 1T or to FL-NRL exons 1 and 2 were used to assess Tr-NRL and FL-NRL expression (Figure 3E,F).

Differential expression of NRL isoforms in rod and cone precursors.

A. Expression of NRL gene and the most highly assigned Ensembl isoforms ENST00000397002 (FL-NRL) and ENST00000560550 (Tr-NRL). B. Mean NRL isoform assignments for clusters defined in Figure 1B, presented as total counts (top) and percentage of total counts (bottom). Significance for LM vs. ER fold change, colored by isoform. ****, p <0.0002; *****, <0.000001 (bootstrapped Welch’s t-test). Ensemble transcript IDs shown in color with structures shown in Figure S4B. C. Top: Mean read counts (fragments per million, FPM) across Ensembl NRL exons for each cluster. Bottom: Transcript structures numbered according to amino acid positions. Minimal transactivation domain (MTD) in green. Arrowheads: Red/black: First exons where red is higher of two peaks. D. Relative difference box plot of raw reads mapping to truncated (Tr) and full length (F) transcript first exons in each cell, according to cluster. Relative difference is the difference in reads mapping to truncated and full-length NRL first exons (Tr-F) divided by the sum of both (Tr+F). Values >0 indicate more reads assigned to truncated isoform, values < 0 indicate more reads assigned to full-length isoform. ***, p <0.0001 (post-hoc Dunn test). E. NRL and RXRγ immunostaining and RNA FISH with probes specific to truncated Tr-NRL exon 1T (green puncta) and FL-NRL exons 1 and 2 (red puncta) in FW16 retina. Boxed regions enlarged at right show an RXRγlo, NRL+ rod with one Tr-NRL and six FL-NRL puncta (top) and an RXRγhi, NRL− cone with one Tr-NRL and no FL-NRL puncta (bottom), indicated with same-color arrows. Scale bar: 10 µm. F. Ratio of fluorescent puncta observed in experiment depicted in (E) for NRL+ or RXRγhi cells where Tr puncta >0. ***; p <0.0005 (Welch’s t-test). G. Long-read nanopore sequencing of pooled 5’ RACE reactions initiated with NRL exon 3 primers and performed on cDNA libraries from 23 ER cells (top) and 21 LM cells (bottom). Each schematic shows total exon coverage (above) and individual transcripts (below), where expressed sequences are grey and introns light blue. Full-length (FL), alternatively spliced or internally initiated exon 2 (βex2), and truncated (Tr) transcripts are indicated by brackets. Red arrow: Transcripts resembling DD10, with internal exon 2 transcription initiation and premature splicing to exon 3. Ensembl FL-NRL and Tr-NRL transcript isoforms and RACE primer positions are shown below. H. Top: PDE6B-luciferase reporter activity in NIH-3T3 cells transfected with indicated amounts (ng) of pcDNA4-C-EF1α and derived FL-NRL and Tr-NRL constructs. Bottom: PDE6B-luc reporter and pcDNA4-C-EF1α expression constructs. Blue box = NRL response element. Error bars = standard deviation of triplicate measurements. *, p <0.05; **, <0.005 (Student’s t-test). Data representative of two experiments in NIH-3T3 and one in 293T.

While the Tr-NRL-encoding NRL isoforms were not to our knowledge previously described, another NRL isoform that initiated within exon 2 and lacked the NRL transactivation domain due to alternative exon 2 splicing, termed DD10 (Figure 3C), was previously identified in adult retina36. Concordantly, we detected reads spanning the unique DD10 splice junction, yet at lower levels than the unique Tr-NRL junction (5,942 versus 57,048).

As transcript isoforms inferred from short-read sequencing do not necessarily reflect the original transcript structures, we further defined the NRL isoforms expressed in early cone and rod precursors by performing 5’ rapid amplification of cDNA ends (RACE) on the already generated single cell cDNA libraries from 23 early rod (ER) and 21 early L/M cone (LM) cells, followed by nanopore long-read sequencing of the pooled RACE products. The long-read sequencing revealed isoforms consistent with FL-NRL, Tr-NRL, DD10, and several other NRL isoforms with alternative transcription initiation and alternative splicing within exon 2 as well as within the Tr-NRL exon 1T (Figure 3G, Figure S4C). In keeping with the computationally assigned isoforms and differential exon usage, ER libraries had a higher proportion of FL-NRL exon 1 and exon 2 reads and LM libraries had a higher proportion of Tr-NRL exon 1T reads (Figure 3G, Figure S4C). Moreover, alternative splicing within NRL exon 2 was more prevalent in LM libraries, affecting 6,788 of 29,177 (23 %) of exon 2 reads compared to 5,378 of 85,860 (6 %) in ER cells (p<0.0001; Chi-square with Yates correction), resulting in rarer full-length (exon 1-2-3) transcripts than inferred from short-read sequencing. Thus, long-read sequencing revealed cell type-biased NRL isoform expression with a preponderance of FL-NRL transcripts in early rods and disrupted FL-NRL and Tr-NRL transcript isoforms in L/M cones.

Despite our detection of L/M cone RNAs encoding Tr-NRL and FL-NRL, cone expression of NRL protein has not been reported. To assess endogenous Tr-NRL expression, we performed immunoblot analysis of CHLA-VC-RB-31 retinoblastoma cells37, which were predicted to express FL-NRL and Tr-NRL transcripts in a cone-like 0.73:1 ratio, with an antibody that recognizes both FL-NRL and Tr-NRL proteins (Figure S5A,B). As in other retinoblastoma cells13, FL-NRL increased in response to retinoic acid and proteasome inhibition, whereas Tr-NRL was not detected (Figure S5C). Similarly, Tr-NRL was not detected in EGFP+, RXRγ+ cones following lentiviral transduction of an explanted fetal retina with an EGFP-P2A-Tr-NRL cassette (Figure S5D-G). These findings suggest that Tr-NRL is poorly translated or unstable in most cone precursors.

As Tr-NRL might be expressed in contexts that were not examined in our analyses, we explored the function of the putative Tr-NRL protein. As both Tr-NRL and the previously described NRL DD10 lack the NRL transactivation domain and as DD10 interferes with FL-NRL transactivation35, we examined if Tr-NRL similarly opposes FL-NRL transcriptional activity. Indeed, in luciferase reporter assays, Tr-NRL suppressed FL-NRL activation of a PDE6B promoter (Figure 3H).

For THRB, the most highly assigned transcript isoforms encoded the L/M cone-specific TRβ2 (ENST00000280696) and the more widely expressed TRβ1 (ENST00000396671 and others) (Figure 4A-C, S6). While both TRβ1 and TRβ2 promote L/M cone fate determination38, the isoform encoding TRβ2 predominated in L/M cones while those encoding TRβ1 predominated in early rods (Figure 4B). Moreover, late rods preferentially expressed THRB exons 1-6 and the THRB 3’ untranslated region (UTR), implying that RNAs encoding full-length TRβ proteins were rare (Figure 4B,C). Notably, a higher percentage of reads extended from the exon 4 and exon 6 splice donor sequences into the subsequent introns in LR versus LM cells (Figure 4D,E) (p ≤ 0.001 for both, two-tailed Chi square test), suggesting that premature transcription termination (PTT) in introns 4 and 6 preferentially limits TRβ expression in the LR population. The inferred PTT events are consistent with structures of the assigned Ensembl isoforms (Figure 4B, S6).

Differential expression of THRB isoforms in rod and cone precursors.

A. Expression of THRB and highly assigned isoforms ENST00000280696 (encoding TRβ2) and ENST00000396671 (TRβ1). B. Mean THRB isoform assignments for each cluster presented as counts (top) and percentage of counts (bottom); Ensemble transcript IDs shown in color with β2, β1, and β1 PTT isoform structures as in Figure S6B. C. Top: Mean read counts across Ensembl THRB exons. Bottom: Transcript structures for TRβ1, TRβ2, and two TRβ1 truncations. Green arrowhead: First TRβ2 exon. ENST00000396671 exon numbers are indicated above and protein domains (AF1, DNA-binding (DBD), and ligand binding (LBD)) below. D. Read coverage for LR cells across THRB exons 4 and 6 splice donor sites. E. Percentage of exon splice donor reads that are spliced or readthrough to the subsequent intron. F. Long-read sequencing of pooled 3’ RACE reactions initiated with exon 4 (left) or TRβ2 exon 1 (right) performed on cDNA libraries from 21 LM cells (top) and 5 LR cells (bottom). Schematics show total coverage (above) and individual transcripts (below). TRβ1 and TRβ2 first exons (green boxes) are enlarged at right. Red arrowheads: intronic premature transcription termination (PTT).

To further evaluate PTT events, we used THRB 3’ RACE and long-read sequencing on single cell cDNA libraries from 23 L/M cone cells and five LR cells selected for high THRB expression. RACE reactions were performed separately with primers complementary to the TRβ1-specific exon 4 and to the TRβ2-specific first exon (Figure 4F). Sequencing of LM and LR RACE products corroborated pronounced PTT in introns 4 and 6, with greater intron 6 PTT in LR versus LM cells (Figure 4F, Figure S7). However, we did not corroborate the late rods’ proportionately higher intron 4 PTT, likely due to the small number of LR cells examined and heterogeneity in PTT frequency in individual cells. Long read sequencing also revealed PTT following the TRβ2-specific exon and a novel transcription-terminating exon following the canonical exon 5 observed solely in TRβ2 transcripts (Figure 4F, S7). 3’ RACE transcripts rarely extended into the 3’ UTR in LM or LR cells, suggesting that reads mapping to this region do not reflect protein-coding RNAs and confound the assessment of protein-coding THRB mRNA expression. These analyses demonstrate that THRB is regulated by multiple PTT events in rod and cone precursors as well as by cell type-specific promoter utilization and independent 3’ UTR RNA expression.

For RXRG, short read sequencing reads were assigned to several isoforms that differed in their 5’ promoter position and exon utilization (Figure S8A-C). However, 5’ RACE and long-read sequencing did not support differential isoform expression (Figure S8D), and quantitative imaging revealed an average ∼3.5-fold higher RXRγ protein in cones compared to rods (Figure S8E). Thus, long-read sequencing clarified that RXRG expression is moderately higher in human cone versus rod precursors without evidence of cell-type specific isoforms.

Two post-mitotic immature photoreceptor precursor populations

To further define photoreceptor developmental states, we subdivided the initial six cell populations with higher resolution clustering (level 1.6), identified high resolution cluster- associated genes and regulons, and inferred each cell’s rate and direction of transcriptomic change using RNA velocity39 (Figure 5A,B). Increased clustering resolution divided the RPC/MG cluster into separate RPC and Müller glia (MG) groups, divided L/M cones into four subgroups (LM1 – LM4), which partially overlapped in UMAP space, and divided the iPRP cluster into two clusters here designated immature cone precursors (iCPs) and immature rod precursors (iRPs), the latter also drawing cells from the low-resolution ER cluster (Figure 5A,C). Similar clusters were observed at reduced k.param values that defines nearest neighbors.

Two post-mitotic photoreceptor precursor populations expressing rod or cone markers.

A. UMAP plot colored by high resolution clusters. B. RNA velocity plots with cell clusters as in A. C. Enlarged view highlighting RPC and MG clusters (left), RPC-localized iCP and iRP clusters (middle), and RNA velocity (right). Black line: limit of RPC-localized region. Arrows depict inferred trajectories. D. Violin plots depict expression of selected genes in RPC, MG, and RPC-localized iCP and iRP cells. Colored asterisks compare cluster of same color to cluster at right of line. E. SCENIC regulon violin and box plots for RPC-localized cells in each cluster, selected from most specific regulons for MG, RPC, ER and LM clusters. ^, pAdj <0.1; *, <0.05; **, <0.005; ***, <0.0005 (post-hoc Dunn test).

The distinction between RPCs and MG was corroborated by the expression of known marker genes, with the RPC cluster having increased expression of cell cycle markers (CCNE2, CCNA2, CCNB2) and the cell cycle-associated PBK and E2F7 (Figure 5D, S9A,B, Table S7).

The MG cluster was spatially segregated, lacked expression of most cell cycle genes, and expressed genes known to be expressed in both populations (CCND1, SLC1A3, PAX6, LHX2, VSX2) as well as the MG marker RLBP140–42 (Figure 5C,D, S9A). Differential gene expression analyses revealed upregulation of cell cycle-related genes and G2/M checkpoint and E2F target ontologies in the RPC cluster but no significantly upregulated ontologies in the MG population (Figure S9B,C), consistent with early MGs resembling quiescent RPCs43. In contrast to other studies, we did not distinguish primary RPCs (PRPCs) from neurogenic RPCs (NRPCs)14,15,20, likely due to the underrepresentation of RPCs in our dataset.

The high resolution iCP and iRP clusters included ‘RPC-localized’ cells positioned adjacent to RPCs and ‘non-RPC-localized’ cells adjacent to the LM and ER clusters in UMAP space (Figure 5C). RNA velocity suggested that the RPC-localized cell groups flowed towards the larger early maturing L/M cones and rods, respectively (Figure 5B,C). Compared to RPCs and MG, RPC-localized iCP and iRP cells had minimal expression of cyclin RNAs or RPC markers PAX6, LHX2, and VSX2, and had increased expression of photoreceptor determinants OTX2 and CRX (Figure 5D), consistent with their being immediately post-mitotic photoreceptor precursors. iCP cells in this region upregulated the early cone cell fate determinant ONECUT13,44, the L/M cone determinant THRB, and the cone differentiation marker GNAT2, whereas iRP cells trended towards higher expression of the rod determinant NR2E3 (Figures 5D, S10A); however, such RNAs may be more highly expressed in cone- and rod-fated NRPCs than in the unspecified RPCs in our study. Notably, THRB and GNAT2 expression did not significantly change while ONECUT1 declined in the subsequent non-RPC-localized iCP and LM1 stages, whereas NR2E3 and NRL dramatically increased on transitioning to the ER state (Figure S10A).

In the RPC-localized region, iCPs had higher ONECUT1, THRB, and GNAT2, whereas iRPs trended towards higher NRL and NR2E3 (p= 0.19, p=0.054, respectively; Figure 5D). While we detected differential expression when selectively examining these genes of interest, no genes were differentially expressed (Padj < 0.05) in a transcriptome-wide comparison, likely due to the lack of statistical power given the small number of cells examined.

To our knowledge, past studies have not distinguished immature cone and rod precursors (i.e., iCPs and iRPs) from the subsequent maturing cone and rod precursor states. To explore whether early cone and rod precursors with similar properties are present in droplet-based scRNA-seq studies, we examined gene expression in spatiotemporally segregated cone and rod precursor populations as well as in cone and rod NRPCs in the Zuo et al. 3’ snRNA-seq dataset20. This revealed that cone NRPCs and the earliest (pcw 8-10) cone precursors had high ONECUT1 and THRB yet minimal NR2E3 (Figure S10B). Similarly, rod NRPCs had high NRL and NR2E3, yet it was not possible to examine the earliest (pcw 10-13) rod precursors, as many had cone-like gene expression profiles, suggesting they may have been mis-assigned (Figure S10B, grey boxes). Further evaluation suggested misassignment of a smaller proportion of rod and cone precursors at other ages (Figure S11). In particular, cells potentially mis-assigned as rod precursors were located in the cone precursor UMAP region and highly expressed THRB (Figure S11D, **), while cells potentially mis-assigned as cone precursors were located in the rod precursor UMAP region and highly expressed NR2E3 (Figure S11E, *). When considering only the main cone and rod precursor UMAP regions, early (pcw 8-13) cone precursors expressed THRB and lacked NR2E3 (Figure S11D,E, blue arrows), while early (pcw 10-15) rod precursors expressed NR2E3 and lacked THRB (Figure S11D,E, red arrows), similar to RPC-localized iCPs and iRPs in our study (Figure 5D). We conclude that Louvain clustering of deep full-length scRNA-seq distinguished immediately post-mitotic iCP and iRP populations with features similar to early cone and rod precursors in a large 3’ snRNA-seq dataset.

In a further comparison, we noted lower detection of GNAT2 in L/M cone precursors and lower detection CCNE2, CCNA2, and CCNB2 in RPCs in 3’ snRNA-seq versus full-length scRNA-seq datasets (Figure S10A,B). This suggests that deep full-length scRNA-seq enabled more sensitive detection of certain cell-type-specific RNAs along with improved discrimination of immature cone versus rod precursor states.

SCENIC regulons further clarified the identities of the four RPC-localized clusters. The MG cluster was best specified (i.e., had highest regulon specificity scores) by PAX6 and SOX9, both previously described in RPCs and MGs40,45,46, whereas RPCs were best specified by E2F2 and E2F3 (Figure 5E). These RPC and MG regulons were low or absent in the early post-mitotic iCPs and iRPs, suggesting they undergo an abrupt cell state change. Moreover, RPC-localized iCPs had greater THRB regulon signal than RPC-localized iRPs, while RPC-localized iRPs had higher NRL regulon signal (p<0.05 for both) (Figure 5E). Interestingly, RPC-localized and/or non-localized iCPs also had higher activity of the pan-photoreceptor regulons LHX3, OTX2, and NEUROD1 compared to their iRP counterparts (Figure S10C,D). Finally, activities of the cone-specific THRB and ISL2 regulons, the rod-specific NRL regulon, and the pan-photoreceptor LHX3, OTX2, CRX, and NEUROD1 regulons increased to varying extents on transitioning from immature iCP or iRP states to the early-maturing LM1 or ER states (Figure 10C).

Early cone and rod precursors with rod- and cone-related RNA co-expression

In addition to the immediately post-mitotic RPC-localized iCPs, the iCP cluster included cells bridging the UMAP region between early maturing cones and early maturing rods (Figure 6A). Many iCP and ER cells in this bridge region expressed RNAs encoding cone markers (GNAT2, THRB), rod markers (GNAT1, NR2E3), or, in many cases, both (Figure 6B), suggestive of a proposed hybrid cone/rod precursor state more extensive than implied by the co-expression of different THRB and NRL isoforms4.

An iCP sub-population with cone- and rod-related RNA co-expression.

A-C. UMAP ‘bridge’ region cells colored by ER and iCP cluster and RNA velocity (A), rod and cone marker gene expression (B), and NRL and THRB regulon activity (C, arrows indicate cells with both regulon signals). D. Box plot of Z-score-normalized NRL and THRB regulon AUCs for each cluster; Bridge ER and Bridge iCP represent cells present in the UMAP region in panels A-C.

To determine whether the intermixed cone and rod gene expression reflects a hybrid cone-rod precursor state, we examined bridge region NRL and THRB regulon activities, which embody the overall transcriptomic effects of these factors. Indeed, in the bridge region, cells with NRL and THRB regulon signals intermixed, and some cells showed signals for both (Figure 6C, arrows). However, iCP cells had low NRL regulon z-scores similar to L/M cones and most had THRB z-scores above that of adjacent ER cells (Figure 6D). The difference between regulon signals increased in rod and cone clusters outside the bridge region as NRL or THRB regulons became more dominant (Figure 6D).

To identify cells with early cone- and rod-related RNA co-expression in the developing retina, we performed multiplex RNA FISH for early cone marker GNAT2 and rod marker NR2E3 combined with immunofluorescence (IF) staining for RXRγ, which has high expression in outermost neuroblastic layer (NBL) cone precursors and low expression in middle NBL rod precursors (as earlier shown in Figure S8E), and for NR2E3, which is detected solely in rods. To infer the spatiotemporal pattern of such expression across human retinal development, we examined GNAT2 and NR2E3 RNA co-expression in RXRγ+ cone precursors in the outermost NBL and in RXRγ+ rod precursors in the middle NBL across 13 regions of a FW14 retina section (Figures 7; see Figure S12 for IF + FISH images). Limiting our analysis to the outer and middle NBL allowed us to disregard RXRγ+ retinal ganglion cells in the retinal ganglion cell layer or inner NBL.

Cone-related GNAT2 and rod-related NR2E3 RNA co-expression in human cone precursors.

A,B. Tiled composite fluorescence image of FW14 retinal section after immunofluorescence staining of RXRγ and NR2E3 and RNA FISH of GNAT2 and NR2E3 (A) and diagram of the same section indicating the most peripheral (i.e., youngest and least mature) and most central (i.e., oldest and most mature) rods and cones (B). Boxes indicate regions further evaluated in Supplementary Figure S12 and quantitated in panels C and D. C,D. Quantitation of (C) outermost and (D) middle (i.e., sub-outermost) NBL photoreceptor precursors expressing combinations of RXRγ and NR2E3 proteins and GNAT2 and NR2E3 RNAs (n.b., italics are used for RNAs, non-italics for proteins). E. Patterns of GNAT2 and NR2E3 RNA and RXRγ and NR2E3 protein expression inferred from in situ hybridization and immunofluorescence staining. RXRγ is expressed in the outermost NBL starting in the far periphery, consistent with cone precursors, and in middle NBL cells, starting more centrally, consistent with rod precursors. GNAT2 and NR2E3 RNA co-expression in outermost NBL cells lacked NR2E3 protein, representing putative cone precursors. RXRγ+ retinal ganglion cells in the inner NBL and ganglion cell layer not shown.

The analyses revealed that most of the far peripheral (hence, nascent) outer NBL RXRγ+ cone precursors (Figure 7A,B, regions 2 and 13) lacked detectable GNAT2 and NR2E3 RNA (green bars in Figure 7C), whereas those in the more mature central retina (regions 3-12) were uniformly GNAT2+ (yellow and red bars in Figure 7C). However, starting with regions 3 and 11, some outermost NBL GNAT2+ cones were also NR2E3+ (red bars in Figure 7C). The proportion of outermost NBL cells that co-expressed GNAT2 and NR2E3 RNA increased in the more central retina (regions 5-10), yet all of these cells had strong RXRγ staining and lacked NR2E3 protein, consistent with their having a cone identity (illustrated in Figure 7E). In contrast, most middle NBL RXRγ+ cells had either a low number of NR2E3 RNA puncta without NR2E3 protein (NR2E3+ / NR2E3−, magenta bars in Figure 7D), likely in nascent rods, or prominent NR2E3 RNA with NR2E3 protein (NR2E3+ / NR2E3+, purple bars in Figure 7D), in maturing rods (Figure 7D, E). However, we did not detect GNAT2 puncta in middle NBL RXRγ+, NR2E3+ cells, including the most peripheral RXRγ+ cells with low-level NR2E3 RNA (Figures 7E, S12). In summary, most photoreceptor precursors with GNAT2 and NR2E3 RNA co-expression had high RXRγ, no detectable NR2E3 protein, and outermost NBL positions expected of cone precursors. This supports the notion that, in our scRNA-seq analyses, bridge region iCP cells with combined GNAT2 and NR2E3 RNA expression are likely cone-directed.

Cone and rod precursor populations that co-express cone and rod marker genes were also evident in the Zuo et al. 3’ snRNA-seq dataset. While the earliest cone precursors were NR2E3-negative and the earliest rod precursors were THRB-negative (as described above), starting at pcw 15, later cone precursors expressed NR2E3 (Figure S11E, green arrows) and later rod precursors expressed THRB (Figure S11D, purple arrows). THRB was highest in the most distal rod precursors (yet still lower than in cone precursors), corroborating our findings in Figure 1D and Figure 4B-C. In contrast, in the 3’ snRNA-seq analysis, GNAT2 expression was sparse in rod precursors of all ages (Figure S11F), consistent with the lack of GNAT2 in rod precursors in our RNA FISH analyses and implying that different cone- and rod-related genes have different tendencies to be expressed in the other photoreceptor precursor type.

Thus, a 3’ snRNA-seq analysis confirmed the initial production of immature photoreceptor precursors with either L/M cone-precursor-specific THRB or rod-precursor-specific NR2E3 expression, followed by lower-level co-expression of their counterparts, NR2E3 in cone precursors and THRB in rod precursors. However, in the Zuo et al. analyses, the co-expression was first observed in well-separated UMAP regions, as opposed to a region that bridges the early cone and early rod populations in our UMAP plots. These findings are consistent with the notion that cone- and rod-related RNA co-expression begins in already fate-determined cone and rod precursors, and that such precursors aberrantly intermixed in our UMAP ‘bridge region’ due to their insufficient representation in our dataset.

Developmental expression of photoreceptor precursor markers and fate determinants

To further assess relationships between iCP subpopulations, we examined the expression and UMAP distributions of the four iCP marker genes identified in Figure S9A. Among these genes, CHRNA1 was mainly expressed in bridge region iCP and ER cells, whereas ONECUT1 was biased to cone-directed iCPs, lncRNA CTC-378H22.2 was expressed in a narrow zone of THRB+, GNAT2+ iCPs, and S100A6 was more widely expressed (Figure 8A). iCP cells also expressed ATOH7 and DLL3, which were previously proposed to define transitional photoreceptor precursors and promote cone fate determination14–16. ATOH7 was largely restricted to RPC-localized and bridge region iCPs whereas DLL3 was broadly expressed similar to S100A6 (Figure 8A).

Developmental expression of early cone and rod precursor markers.

A. UMAP plots of iCP marker genes CHRNA1, ONECUT1, CTC-378H22.2, and S100A (see Figure S9A) and previously identified photoreceptor precursor markers ATOH7 and DLL3. B-D. Combined RXRγ/NRL immunohistochemical staining and CHRNA1 RNA FISH of FW12 retina. B. Tiled images of retina section with nuclei stained with DAPI. White boxes: fields used for quantitative fluorescent imaging. Distances along apical edge of tissue marked in mm from midpoint of central image (0 mm, C). *: Imaged region shown in C. Scale bar = 500 µm. C. Top: Retinal nuclear and cellular segmentation and identification of cells as RXRγ+ (green outline) or NRL+ (red outline). Yellow box: Field shown below. Bottom: RXRγ or NRL immunofluorescence staining with CHRNA1 FISH. Arrows: RXRγ+,CHRNA1+ (green), NRL+,CHRNA1+ (yellow). Scale bars = 15 µm. D. Quantitation of fluorescent puncta in RXRγ+ and NRL+ cells by image field. X-axis: Distance from the midpoint of each image to retina center (0 mm, C). **, p <0.005 (Wald test, images from 0-6 mm).

A similar pattern was seen in the Zuo et al. 3’ snRNA-seq dataset20, where CHRNA1, ONECUT1, S100A, ATOH7, and DLL3 were most highly expressed in the youngest and developmentally earliest ML cone and rod precursors and persisted to various extents during cone and rod maturation, whereas CTC-378H22.2 (ENSG00000259436) had more restricted and cone-specific expression (Figure S13A). However, the early cone and rod precursor UMAP regions did not adjoin one another (in contrast to our UMAP bridge region), suggesting that the cone and rod precursors expressing the above markers had begun their distinct trajectories and that their juxtaposition in our UMAP analyses is spurious. Also, different iCP markers had different spatiotemporal expression: CHRNA1 and ATOH7 were most prominent in peripheral retina with ATOH7 strongest at pcw 10 and CHRNA1 strongest at pcw 13; CTC-378H22.2 was prominently expressed from pcw 10-13 in both the macula and the periphery; and DLL3 and ONECUT1 showed the earliest, strongest, and broadest expression (Figure S13B). The distinct patterns suggest spatiotemporally distinct roles for these factors in cone precursor differentiation.

As CHRNA1 appeared to be the most specific marker of both early cone and early rod precursors, we evaluated whether CHRNA1 RNA marked specific cone and rod populations in the developing retina using RNA-FISH combined with RXRγ and NRL immunofluorescent staining. In a FW12 retina, we detected highest CHRNA1 in the earliest, most peripheral NRL+ rods and RXRγ+ cones and fewer CHRNA1+ cells in more mature, central regions (p <0.005 in cones) (Figure 8B-D), consistent with its expression in early cone and rod precursors.

To identify factors that regulate transcriptomic states during cone cell fate determination, we examined the UMAP distributions of the most iCP-specific transcription factor regulons, OLIG2 and LHX9 (Figure S10D, Table S8). As Olig2 was previously detected in mouse RPCs in which Onecut1 enabled Thrb expression and cone fate determination3,47, the OLIG2 regulon was expected to be most active in RPCs and in early, immediately post-mitotic iCPs. While the OLIG2 and LHX9 regulons were indeed active in RPCs, both were also active in a narrow zone of ONECUT1+ iCPs positioned farthest from S cones and immediately preceding the upregulation of the THRB regulon, a location with sparse OLIG2 RNA expression (Figure S14). These data indicate that the OLIG2 and LHX9 regulons are active in and potentially relevant to human L/M cone fate determination in a post-mitotic iCP subpopulation.

An early L/M cone trajectory marked by successive lncRNA expression

After fate commitment, L/M cones initiate a maturation process with upregulation of RNAs and proteins related to phototransduction, axonogenesis, synaptogenesis, and outer segment morphogenesis5,6. To evaluate whether early L/M cone maturation is comprised of distinct transcriptomic cell states, we assessed marker gene and regulon differences between high resolution clusters LM1, LM2, LM3, and LM4. While clusters LM1-4 were distinguished by sequential increasing expression of ACOT7, RTN1, PDE6H, OLAH, and NPFF, they showed only subtle differences among the three regulons with highest LM1-4 specificity scores, THRB, ISL2, and LHX3 (Figure 9A,B). The lack of cluster-specific marker genes and regulons suggest that LM1-LM4 represent different stages of a graded maturation process.

L/M cone subcluster marker genes, regulons, and pseudotemporal trajectory with successive lncRNA gene expression.

A. Violin plots of high-resolution cone cluster marker genes with increasing maturation-associated expression. All significant differences between adjacent clusters are indicated. B. Violin plots of regulons with highest LM1-4 cone cluster specificity scores. *, p <0.05; **, <0.005; ***, <0.0005 (post-hoc Dunn test). C. Pseudotemporal trajectory through the L/M cone population derived with Monocle 3. *: Root cell used to define endpoint. The pseudotime trajectory may be related to LM1-LM4 subcluster distributions in Figure 5A. D. Trendlines of relative count expression (left) and UMAP plots for lncRNAs correlating with early or late-upregulating modules. Line color matched to labels. E, F. Combined RXRγ immunohistochemical staining and FISH of lncRNAs on FW16 retina. E. Tiled images of retina with nuclei stained with DAPI. White boxes: fields used for quantitative fluorescent imaging. Distances along apical edge of tissue marked in mm from fovea to ciliary margins. Scale bar = 500 µm. Asterisks identify fields shown in F. F. Combined RXRγ immunostaining and multiplex FISH for four lncRNAs, of which two are shown in peripheral and central retina regions. Arrows: White: RXRγ/CTC-378H22.2+. Magenta: RXRγ/HOTAIRM1+. Blue: RXRγ/CTD-2034I21.2+. Green: RXRγ/RP13-143G15.4+. Scale bar = 15 µm. G. Quantitation of lncRNA fluorescent puncta assigned to RXRγ+ cells after segmentation. Colored bars mark lncRNA expression regions as described in the text. Arrows indicate significant count differences between images (p <0.05, Wald test).

We further evaluated gene expression changes in maturing cones by pseudotemporally ordering iCP and L/M cone precursors (Figure 9C). This identified 967 pseudotime-correlated genes (q-value < 0.05, expression > 0.05 in >5% of cells; Table S9) in seven gene modules (Figure S15). Among the top 20 pseudotime-correlated genes in each module, we identified four lncRNAs that were sequentially expressed (Figure 9D). To determine if these lncRNAs distinguish developmentally distinct maturing cones in vivo, we probed their expression via multiplex RNA FISH in co-stained RXRγ+ cone precursors across a FW16 retina (Figure 9E,F). Quantitation of FISH puncta defined four cone maturation zones based on expression peaks and significant count differences for each lncRNA as color-coded in Figure 9G: the most peripheral cones with high CTC-378H22.2 and HOTAIRM1 (blue), peripheral cones with high HOTAIRM1 only (green), cones with low expression of all four lncRNAs (red), and parafoveal/foveal cones with upregulated RP13143G15.4, CTD-2034I21.2, and CTC-378H22.2 (purple) (Figure 9G). The detection of foveal CTC-378H22 by ISH but not in late LM transcriptomes may relate to a lack of the most mature cones in our scRNA-seq analyses. These data support the concept that maturing L/M cones sequentially express specific lncRNAs as they develop.

Cone-intrinsic SYK expression associated with the proliferative response to pRB loss

We next assessed whether early-maturing cone transcriptomic features are conducive to the proliferative response to pRB loss10,11. Given the high transcriptomic similarity across the L/M cone population, we compared gene expression between all early-maturing L/M cones and early-maturing rods. This identified 422 genes upregulated and 119 downregulated in cones (p<0.05, log2FC>|0.4|) (Figure S16A, Table S10). Among cone-enriched genes, the top three enriched ontologies related to translation initiation, protein localization to membrane, and MYC targets (Figure S16B). The upregulation of MYC target genes was of interest given that many MYC target genes are also targets of MYCN, that MYCN protein is highly expressed in maturing (ARR3+) cone precursors but not in NRL+ rods (Figure 10A), and that MYCN is critical to the cone precursor proliferative response to pRB loss8–10. Indeed, whereas MYC RNA was not detected, the LM cone cluster had increased MYCN RNA (log2FC = 0.54) and MYCN regulon activity, representing the seventh highest LM cluster regulon specificity score (Figure 10B,C, Table S6).

Cone intrinsic MYCN and SYK associated with proliferative response to pRB loss.

A. Immunofluorescent staining shows high MYCN in ARR3+ cones but not in NRL+ rods in FW18 retina. B-D. UMAP plots of MYCN expression (B), MYCN regulon activity (C), and SYK expression (D). E. SYK and MYCN gene expression violin plots by cluster. *, p <0.05; ns = not significant (t-test). F. Immunohistochemical staining of SYK and cone arrestin (ARR3) in FW18 and FW16 retinae. Green arrow: ARR3+, SYK+. White arrow: ARR3+, SYK–. Scale bar = 25 µm. G. Top: Effect of SYK inhibitor GS-9876 on Ki67 expression in RXRγ+ cones from FW16.5 retina co-transduced with YFP and shRB1- or control shSCR-shRNA. Values represent means of three analyses from two treatment replicates. Error bars: standard deviation. *, p <0.05; **, <0.005 (Student’s T-test with equal variance, 2-tailed). RXRγ+ cells: Experiment 1, n=1,340. Experiment 2, n=804. Range 107 – 366 cells per condition. Bottom: Example of Ki67, YFP, and RXRγ co-immunostaining with DAPI+ nuclei (yellow outlines). Arrows: Ki67+, YFP+, RXRγ+ nuclei. H. Model of SYK expression in cone maturation and retinoblastoma development.

Among other differentially expressed genes, we noted the L/M cone-specific upregulation of SYK (Figure 10D, Table S10), which encodes a non-receptor tyrosine kinase. Whereas SYK was previously implicated in retinoblastoma genesis and proposed to be induced in response to pRB loss48, its expression was not previously reported in developing fetal retina. Indeed, our scRNA-seq analyses revealed that SYK RNA expression increased from the iCP stage through cluster LM4, in contrast to its minimal expression in rods (Figure 10E). Moreover, SYK expression was abolished in the five-cell group with properties of late-maturing cones (characterized in Figure 1E), here displayed separately from the other LM4 cells and designated LM5 (Figure 10E). Similarly, immunohistochemical staining revealed high SYK protein expression in immature (ARR3-) and early-maturing (ARR3+) cones from the retinal periphery to the maturing foveal cones at FW16, while SYK was not detected in the most mature foveal cones at FW18 (Figure 10F). The loss of SYK protein and RNA expression with cone maturation is consistent with the lack of SYK in cones of normal retina adjacent to a retinoblastoma tumors48 and implies that SYK is a defining feature of the human early cone precursor state.

While MYCN RNA was also preferentially expressed in early cones, it did not increase as much relative to RPCs nor decline as much in late-maturing cone precursors when compared to SYK RNA dynamics (Figure 10E).

To determine if SYK might contribute to retinoblastoma initiation, dissociated fetal retinal cells were transduced with an RB1-directed shRNA (shRB1-733) known to induce cone precursor proliferation10, treated with the selective SYK inhibitor GS-987649 for 12 days, and examined for cone precursor cell cycle entry by co-staining for RXRγ and Ki67. GS-9876 treatment suppressed the proliferative response to pRB knockdown at all concentrations from 1.0 – 5.0 μM (Figure 10G), consistent with the notion that cone precursor intrinsic SYK activity contributes to the proliferative response to pRB loss and is retained in retinoblastoma cells (Figure 10H). However, given potential SYK inhibitor off-target effects, validation of the role of SYK in retinoblastoma initiation will require gene ablation studies.

Discussion

This study evaluated cell state changes associated with human cone and rod photoreceptor development using deep full-length scRNA-seq. Whereas prior scRNA-seq studies provided insights into RPC fate determination and trajectories using 3’ end counting14–18,20,26, deep full-length sequencing enabled more precise resolution of cell states and identification of cell type-specific regulon activities and transcript isoforms. Additionally, our cell enrichment strategy provided insight into the transcriptomic profiles and potential cancer-predisposing features of developing cones, a rare population with unique cancer cell-of-origin properties8,10,11.

One advantage of full-length scRNA-seq is that it enables the detection of differential transcript isoform expression. Here, we show that the rod determinant NRL, the L/M-cone determinant THRB, and the cone marker RXRG are all co-expressed in cone and rod precursors at the RNA level yet use distinct cell type-specific mechanisms to enable appropriate differential protein expression. Long-read cDNA sequencing revealed that L/M cones preferentially express novel NRL transcript isoforms predicted to encode a truncated NRL protein (Tr-NRL) lacking a transactivation domain as well as NRL isoforms with intra-exon-2 transcription initiation and alternative splicing. As with DD1035, Tr-NRL opposed transactivation by full-length NRL and thus may suppress rod-related transcription. However, our inability to detect intrinsic Tr-NRL or to overexpress Tr-NRL in cone precursors suggests there are additional layers of regulation, possible effects of untranslated Tr-NRL RNA, and alternative contexts in which these isoforms act. A similar assortment of NRL isoforms were expressed in rods, yet a far higher proportion of rod NRL transcripts encoded full-length (FL-) NRL protein. Thus, the L/M cone precursors’ greater use of the Tr-NRL first exon and NRL exon 2 disruptions reveal a multipronged approach to suppress FL-NRL function while downregulating overall NRL RNA by only ∼ 4-fold.

Similarly, we uncovered cell type-specific differences in THRB transcript isoforms, albeit generated through premature transcription termination (PTT) and 3’ UTR expression in late rod precursors. PTT is widely used to govern expression of transcription regulators50, and while PTT-shortened THRB transcripts had been identified51, their retinal cell specificity was not previously recognized. In contrast, the expression of THRB 3’ UTR sequences independent of THRB protein-coding sequences was not previously reported and may enable additional regulation52. As a general matter, the expression of 3’ UTR transcripts that are detached from protein-coding sequences may be inferred from full-length scRNA-seq but may be misinterpreted to represent protein-coding transcripts and thus confound the interpretation of scRNA-seq performed with 3’ end-counting. To address this issue, our publicly available Shiny app displays exon coverage for all genes expressed in the RPC and photoreceptor clusters in this study.

One limitation of these studies is that individual cells may express a small and varying spectrum of transcript isoforms. While this variability was mitigated by combining cDNAs of cells deemed to be in similar states based on short-read sequencing, analysis of more cells might better reveal the spectrum of isoforms expressed by specific cell populations

Understanding human photoreceptor development requires the identification of photoreceptor lineage developmental states and elucidation of mechanisms underlying their transitions. While cell states may be defined in various ways, at the transcriptome level they are perhaps best defined by unique combinations of transcription factor regulon activities30. Our deep sequencing approach uncovered discrete cell states distinguished by gene expression as well as by regulon activities. For example, early- and late-maturing rod populations were distinguished by the latter’s increased expression of phototransduction genes together with increased activity of the NRL, ATF4, CRX, and LHX3 regulons and decreased HMX1 and RAX regulons (Figure 2). Similarly, L/M cones formed an early-maturing cone cluster characterized by high THRB and ISL2 regulons, consistent with THRB and ISL2 driving L/M cone identity4,15,53, and a late-maturing cone group with decreased RAX activity. Downregulation of the RAX regulon in late-maturing rod and cone precursors is consistent with decreasing RAX protein during photoreceptor maturation and with RAX-mediated suppression of the cone opsin and rhodopsin promoters54,55.

Increased clustering resolution further divided the L/M cone cluster into four subgroups with subtle maturation-associated changes in marker gene expression and regulon activity, consistent with LM1-4 comprising different stages of a graded maturation process (Figure 9). However, trajectory analyses revealed successive expression of lncRNAs that was validated in developing retinal tissue. Three of these lncRNAs (HOTAIRM1, CTD-2034I21.1 (neighbor gene to CTD-2034I21.2), and the mouse ortholog of CTC-378H22.2 (D930028M14Rik)) were previously observed in cone scRNA-seq15,17,56. Their sequential expression in the absence of discrete regulon changes suggests that they demarcate early L/M cone precursor substates.

High resolution clustering also segregated immature cone precursor (iCP) and immature rod precursor (iRP) populations, which were further partitioned according to their UMAP positions. RPC-localized iCP and iRP cells (those positioned adjacent to RPCs and MG in UMAP plots) lacked cell cycle-related gene expression, showed higher L/M cone-like (THRB) or rod-like (NRL) gene expression and regulon activities, respectively, and had RNA velocities directed toward distinct cone and rod populations (Figure 5). iCPs also upregulated core photoreceptor regulons OTX2 and NEUROD1 to a greater extent than iRPs (Figure S10C), supporting their distinct trajectories. Prior droplet-based scRNA-seq analyses have not to our knowledge distinguished immediately post-mitotic immature cone and rod precursors from the subsequent early-maturing cone and rod precursor stages.

Whereas deep full-length scRNA-seq enabled discrimination of immature cone and rod precursors, 3’ scRNA-seq and snRNAs-seq enabled discrimination of the PRPC and cone- and rod-fated NRPCs 14,15,19 20 (Figure S17A). Combining these observations supports a model in which cone- and rod-fated NRPCs give rise to immediately post-mitotic (i.e., nascent) and subsequent immature cone and rod precursors (iCPs and iRPs), which transition to early-maturing and then late-maturing L/M cone and rod precursors, with each state having unique gene expression and regulon properties (Figure S17B,C). It will be important to corroborate this model with lineage tracing, to extend the model to S cones, and to determine whether late-maturing cone and rod states in fetal retina are distinct from mature post-natal cone and rod photoreceptors at the transcription factor regulon level.

Although not identified as a distinct cluster, our analyses also revealed cone and rod precursor subpopulations that co-express rod and cone genes and regulons (Figure 6, 7, S11, S12, S17C). In both the current full-length scRNA-seq and a recent 3’ snRNA-seq analysis20, the earliest cone and rod precursors lacked such co-expression, implying that this property is acquired after cone versus rod fate-determination. In support of this notion, NR2E3 RNA was detected in more mature central retina cone precursors but not in nascent peripheral cone precursors in fetal retina tissue (Figure 7). However, more information is needed to assess whether such co-expression serves a developmental purpose. For example, it is unknown if cone precursor NR2E3 RNA ever produces NR2E3 protein, as in zebrafish retina, which could suppress cone-related gene expression57 and delay terminal differentiation. Similarly, it is unclear if the THRB RNA expressed in rod precursors – which is largely truncated or comprised of non-coding 3’ UTR sequences (Figure 4) – has direct RNA effects52.

Our characterization of cone and rod-related RNA co-expression may help resolve questions about the retinoblastoma cell of origin. Past studies suggested that retinoblastoma cells co-express RNAs associated with rods, cones, or other retinal cells due to a loss of lineage fidelity12. However, the early L/M cone precursors’ expression of NR2E3 and NRL RNAs suggest that their presence in retinoblastomas12,13 reflects their normal expression in the L/M cone precursor cells of origin. This idea is further supported by the retinoblastoma cells’ preferential expression of cone-enriched NRL transcript isoforms (Figure S5B).

Our analyses also refine understanding of gene expression in the earliest cone and rod precursors. Early cone and rod precursors shared expression of neurogenic precursor markers ATOH7, DLL3, and the new marker CHRNA1, albeit with higher expression of each in the cone lineage and preferential expression of ATOH7 and CHRNA1 in the retinal periphery (Figures 8, S13). ATOH7, DLL3, and CHRNA1 RNAs persist in some maturing cone and rod precursors, whereas CTC-378H22.2 more precisely marks iCPs. iCPs with velocity directed towards early-maturing L/M cones expressed THRB and ONECUT1 RNAs and had OLIG2 regulon activity (Figure S14); as these elements were proposed to promote cone fate in lineage-restricted RPCs3, their expression in iCPs suggests a similar role, such as in L/M cone fate determination, in post-mitotic cells.

We also mined gene expression differences in early cone versus rod photoreceptors to identify factors that impact human cone sensitivity to RB1 loss10,11. We detected upregulation of genes targeted by MYC, some of which are similar to those targeted by MYCN58, along with upregulated MYCN RNA, MYCN protein, and MYCN regulon activity. As MYCN is required for the proliferative response to pRB loss10 and triggers retinoblastoma when amplified and overexpressed8, these findings suggest that increased MYCN RNA expression and regulon activity contribute to pRB-deficient cone precursor proliferation and retinoblastoma genesis. We further found that cone precursors highly express SYK, an oncoprotein previously detected in human retinoblastomas and RB1-null retinal organoids but not in healthy tumor-associated retina48,59. The high SYK expression preceding early cone precursor maturation and its loss during late maturation implies that high-level SYK expression is a retinoblastoma cell-of-origin-specific feature. Moreover, the pRB-depleted cone precursors’ sensitivity to a SYK inhibitor suggests that native SYK expression rather than de novo induction contributes to the cone precursors’ initial proliferation (Figure 10H), although genetic ablation of SYK is needed to confirm this notion.

In summary, through deep, full-length RNA sequencing, we identified photoreceptor cell type-specific differences in transcript isoform expression and post-mitotic photoreceptor precursor states with distinctive gene expression and regulon activities. The discrimination of these states enabled the identification of developmental stage-specific cone precursor features associated with the cone precursors’ predisposition to form retinoblastoma tumors.

Materials and Methods

Primary human tissue

Human fetal samples were provided by Advanced Bioscience Resources, Inc., Alameda, CA, or collected under Institutional Review Board approval at the University of Southern California (protocol HS-13-00399), and Children’s Hospital Los Angeles (protocol CHLA-14-00122). Following the patient decision for pregnancy termination, patients were offered the option of donation of the products of conception for research purposes, and those that agreed signed an informed consent. This did not alter the choice of termination procedure, and the products of conception from those who declined participation were disposed of in a standard fashion. Fetal age was determined according to the American College of Obstetrics and Gynecology guidelines60. Each retina subjected to sequencing was from a different donor.

Retinal dissociation and RNA-sequencing

Single cell isolation

Retinae were dissected while submerged in ice-cold phosphate-buffered saline (PBS) and dissociated as described10. Briefly, retina was removed and placed in 200 µl Earle’s balanced salt solution (EBSS) in a 6-well plate on ice. 2 ml of 10 u/ml papain (Worthington LK003176) solution was added and then incubated at 37°C for 10 min, followed by pipetting with a 1000 µl pipet tip to break up large pieces and additional 5 min incubations until cells were dissociated to single cell level. An additional 1-2 ml papain was added after 20 min if dissociation was insufficient (∼40 min total). Retina culture media [Iscove’s Modified Dulbecco’s Medium with glutamine (Corning, #12440046), 10% fetal bovine serum (FBS, FB-01, Omega Scientific), 0.28 U/mL insulin (Eli Lilly NDC 0002-0213-01), 55 µM β-mercaptoethanol (Fisher Scientific, # 21985023), penicillin and streptomycin] (Fisher Scientific, # MT30002CI) was used to stop enzyme activity and cells were centrifuged in 14 ml round bottom tubes at 300 x g, 4°C for 10 min. Supernatants were centrifuged at 1100 x g for an additional 3 min. After resuspension in Hank’s balanced salt solution (Fisher Scientific, 14025092), a 10 µl volume of cells was combined with 10 µl trypan blue and counted using a hemocytometer.

FACS isolation of single cells and cDNA synthesis

Single cells were FACS-isolated as described10 with differences as follows. Dissociated cells were centrifuged and resuspended in a 5% fetal bovine serum in PBS (FBS/PBS) with CD133-PE (Miltenyi Biotec 130-113-108), CD44-FITC (BD Pharmingen 555478) and CD49b-FITC antibodies (BD Pharmingen 555498) to a final concentration of 10,000 cells/µl. After incubation at room temperature (RT) for 1 h, samples were diluted in 5% FBS/PBS, centrifuged as above, and resuspended in 300-400 µl 1% FBS/PBS containing 10 µg/ml 4′,6-diamidino-2-phenylindole (DAPI). Single cells were sorted on a BD FACSAria I at 4°C using 100 µm nozzle in single-cell mode into each of eight 1.2 µl lysis buffer droplets on parafilm-covered glass slides, with droplets positioned over pre-defined marks (S. Lee et al., in preparation). Gating removed low forward- and side-scatter cells before collecting a CD133-high, CD44/49b-low population. Upon collection of eight cells per slide, droplets were transferred to individual low-retention PCR tubes (eight tubes per strip) (Bioplastics K69901, B57801) pre-cooled on ice to minimize evaporation. The process was repeated with a fresh piece of parafilm for up to 12 rounds to collect 96 cells. cDNA was prepared and amplified using the SMART-Seq V4 Ultra Low Input RNA Kit (Takara Bio 634891) in 10x reduced volume reactions using a Mantis liquid handling system (S. Lee et al., in preparation). Samples were stored at -20°C until quantitation and library preparation. All RNA/cDNA volumes during FACS and processing were handled using low retention siliconized pipette tips (VWR 53503-800, 535093-794).

C1 isolation of single cells and cDNA synthesis

Dissociated cells were FACS isolated as above and collected into a single 1.5 ml tube. Cells were centrifuged and resuspended at 400-800 cells/µl, combined with C1 suspension reagent, and loaded onto the C1 chip and imaged at each site to define cell number in each well. cDNA was synthesized using SMARTer chemistry (SMARTer Ultra Low RNA Kit for the Fluidigm C1 System, Clontech 634835/634935) (three retinae, Seq 1, FW17 and FW13-1), or SMART-Seq V4 (all others). Final cDNA was harvested into low-retention tube strips and stored at -20°C.

Quality control, library preparation and sequencing

DNA was quantitated using Quant-iT PicoGreen dsDNA assay (Life Technologies, P11496) and quality checked by Bioanalyzer. Libraries > 0.05 ng/µl were prepared using the Nextera XT DNA Library Preparation Kit (Illumina, FC-131-1096) and sequenced on the Illumina NextSeq 500 (2x75) (Seq 1) or on Illumina HiSeq 4000 (2x75) (all others).

5’ and 3’ RACE and nanopore sequencing

1 μl of amplified single cell cDNAs libraries produced as above (23 ER cells, 21 LM cells, and 5 LR cells) were re-amplified using the 5’ PCR Primer II A (described for Clontech SmartSeq V4 and purchased from Integrated DNA Technologies) and using CloneAmp HiFi PCR Premix (Clontech Laboratories, Inc. Takara Bio #639298) (Table S9) by heating to 98°C x 2 min followed by 20 cycles of 98°C x 10 s, 65 °C x 15 s, and 72 °C x 35 s, followed by 72 °C x 5 min. Re-amplified cDNAs were quantified using QubitTM Flex Fluorometer (InVitrogen) and 50 pg of each re-amplified library was used for separate RACE PCR reactions with gene-specific NRL, RXRγ, TRβ1 and TRβ2 primers and universal 5’ RACE and 3’RACE primers (Table S9) using CloneAmp HiFi PCR Premix by heating to 98°C x 2 min followed by 30 cycles of 98°C x 10 s, 60 °C x 15 s, and 72 °C x 35 s, followed by 72 °C x 5 min. The “5’ RACE primer for SmartSeq cDNA” sequence (5’–AAGCAGTGGTATCAACGCAGAGTACGGG–3’) was based on the SMART-Seq V4 SMARTer II A oligonucleotide (5’– AAGCAGTGGTATCAACGCAGAGTACXXXXX–3’) after library sequencing revealed that AAGCAGTGGTATCAACGCAGAGTAC was most often followed by GGGNN. The “3’ RACE primer for SmartSeq cDNA” sequence (5’- AAGCAGTGGTATCAACGCAGAGTACTTTT-3’ was based on the SMART-Seq V4 3’ SMART CDS Primer II A (5’– AAGCAGTGGTATCAACGCAGAGTACT(30)(N-1)N–3’) but with only four Ts at the 3’ end. Equal volumes of RACE PCR products were pooled according to cell type, and DNA concentrations of LM (32.8 ng/μl), ER (22.6 ng/μl), and LR (16.5ng/μl) samples determined using Qubit fluorometer. Long read sequencing libraries were prepared and samples barcoded using Oxford Nanopore Technology (ONT) Native Barcoding Kit 24 V14 as specified by the manufacturer for average DNA fragment lengths of ∼ 2kb. 200-300ng of each DNA pool was subjected to repair/dA-tailing (End-Prep) with NEB Ultra II End-Prep Enzyme Mix (New England Biolabs, Ipswich, MA) followed by Native Barcode Ligation as specified by ONT with NB01 assigned to the ER pool, NB02 assigned to the LM pool, and NB03 assigned to the LR pool. Ligation reactions were quenched by addition of 4 μl EDTA and each reaction mixture (24 μl) pooled for a 72 μl volume. The final pool was incubated with Native Adapter (NA) and NEB Quick T4 ligase. The adapter-ligated pool was recovered by AMPure XP magnetic beads and then loaded onto a PromethION flow cell (R10.4.1) as specified by the manufacturer. Sequencing was performed on a PromethION 24 system using MinKNOW software for 72 hours. Raw FAST5 data files were converted to FASTQ files using ONT MinKNOW base calling software.

Computational Methods

Software packages used in this study are described in Table S10.

Read processing and dimensionality reduction

Adapter sequences were removed using the trimgalore wrapper for Cutadapt61. Trimmed FASTQ files were used as input for HISAT262 with a non-canonical splice penalty of 20, maximum and minimum penalties for mismatch of 0 and 1, and maximum and minimum penalties for soft-clipping set to 3 and 1. Aligned bam files were quantified with StringTie63 and Tximport64, yielding cell-by-transcript and cell-by-gene count matrices annotated according to Ensembl build 87. All analyses can be reproduced using a custom snakemake65 workflow accessible at https://github.com/whtns/ARMOR.

For analysis of short-read sequencing of full-length scRNA-seq, dimensionality reduction and data visualization were performed using the Seurat (V3) toolset66,67. Expression counts from all sequencing datasets were integrated using Seurat’s standard integration workflow. Seven sample sequencing batches were integrated after default normalization and scaling using the top 2000 most variable features in each set to identify anchor features. Normalized read counts for gene expression are reported as raw feature read counts divided by total cell read counts, then multiplied by a scaling factor of 10,000 and natural log transformed. These features of the integrated dataset were used to calculate principal component analysis (PCA) and UMAP embeddings68,69. A nearest-neighbors graph was constructed from the PCA embedding and clusters were identified using a Louvain algorithm at low and high resolutions (0.4 and 1.6)70. Read coverage plots for genes of interest were generated using wiggleplotr71 to visualize BigWig files with ENSEMBL isoforms. Individual exon counts were identified using DEXseq72, which takes exons from available isoforms and bins them into intervals before assigning any read counts that overlap a bin to that same bin. These counts were normalized for length for calculation of fold change or relative difference of exon use. Reads crossing target splice donor sites were evaluated as spliced or unspliced from BAM files to determine intron readthrough.

For analysis of 3’ snRNA-seq, processed data provided with Zuo et al20 (CZ CELLxGENE Discover with accession code 5900dda8-2dc3-4770-b604084eac1c2c82) was downloaded as a Seurat Object. A pre-processed AnnData object containing inferred NRPC fate annotations was generously provided by Dr. Rui Chen, cone and rod fated NRPC annotations were transferred to the primary Seurat Object, and data visualization was performed using the Seurat (V5) toolset66,67.

Analysis of long-read sequencing data was performed with the nf-core nanoseq version 3.1.0 pipeline73 (https://nf-co.re/nanoseq/3.1.0). Quality control on raw reads was conducted using FastQC. Reads were aligned using minimap274. SAM files were converted to coordinate-sorted BAM files and mapping metrics were obtained using SAMtools. bigWig coverage tracks were created for visualization using BEDTools and bedGraphToBigWig. bigBed coverage tracks were created using BEDTools and bedToBigBed. Transcripts were reconstructed and quantified using bambu. Quality control results for raw reads and alignment results were presented using MultiQC. All reads whose ends overlapped gene-specific RACE primers were displayed and quantitated using the Integrated Genome Browser (IGV). For THRB 3’ RACE reactions, reads inferred to initiate through exonic internal oligo(dT) priming were removed based on the following features adjacent to the 3′ end of the alignments: six continuous adenines (As), more than seven As in a 10 nucleotide window, AG-runs of six or more nucleotides, eight or more As or Gs in a 10 nucleotide window, eight or more As or high A/T content (27 out of 30 bases), or 12 or more adenines present in an 18 nt window75. The proportion of alternative NRL exon 2 splicing was calculated based on the maximum read coverage between the exon 2 splice acceptor and the first splice donor and the minimum coverage after the internal splice donor as reported in IGV.

Differential expression and overrepresentation analyses

Marker features for each cluster were identified using the Wilcoxon-rank sum tests and specificity scores were computed using the genesorteR76 R package. Only genes with an adjusted p-value (pAdj) <0.05 were considered for further analyses. Differential expression analysis of full-length scRNA-seq was performed using the Seurat FindMarkers function with the Wilcoxon-rank sum test on log-normalized counts. Results were displayed as volcano plots made with EnhancedVolcano77. Differential expression analysis of a 3’ snRNA-seq was performed using the Seurat FindMarkers function with the Wilcoxon-rank sum test on log-normalized counts. Results were displayed as volcano plots made with EnhancedVolcano77.

Overrepresentation analyses were performed with WebGestalt78. Analyses were run with the default settings (5-2000 genes per category, Benjamini-Hochberg multiple-testing correction), gene list enrichment was compared to the genome reference set and ontologies were displayed with a false discovery rate (FDR) <0.05 and ‘weighted set cover’ redundancy reduction. All gene sets evaluated were provided within WebGestalt and the same three were used unless otherwise indicated: GO – Biological Process, KEGG, and Hallmark50.

Transcription factor regulons

Transcription factor regulons were identified using pySCENIC, the python-based implementation of Single-Cell Regulatory Network Inference and Clustering (SCENIC)30,79. The tool was run using the basic settings as shown (https://pyscenic.readthedocs.io/en/latest/tutorial.html). Initial filtering required a gene to have a minimum of 3 raw counts in 1% of cells to be considered for inclusion in a regulon. AUC scores were z-score normalized for comparison between regulons.

Trajectory analysis

RNA velocity analysis was performed with R package velocyto39. Spliced and unspliced count matrices were assembled using the run_smartseq2 subcommand with repeat-masked ENSEMBL build 87. RNA velocity was calculated using cell kNN pooling with a gamma fit based on a 95% quantile and velocities were overlaid on integrated UMAP visualizations using velocyto. Pseudotemporal trajectories were calculated using Monocle 380. The SeuratWrappers package provided integration between Seurat’s final integrated dimensional reduction and Monocle. The data was subset to include only cone-directed iCPs and L/M cones, then a principal graph was fit across the UMAP plot and a ‘root cell’ chosen to represent the latest maturation point in the OPN1LW+ late-maturing cone group. After assigning pseudotime values to each cell, genes that significantly changed as a function of pseudotime were identified and those with a correlation q-value of <0.05 with expression > 0.5 in at least 5% of the cells present in the pseudotime were grouped into modules of co-regulation by performing UMAP and Louvain community analysis.

Histology

Fixation and cryosectioning of fetal retina

Retinae were procured and dissected in cold 1x PBS. The cornea and lens were removed with a cut around the limbus of the iris leaving the front of the eye open. The tissue was submerged in ∼25 ml of 4% paraformaldehyde (PFA) in 1x TBS and placed at 4°C on a rocker at low speed for ∼16-18 hrs. The tissue was washed 3 times with 1x PBS before incubation in 30% sucrose [sucrose + 1x PBS] overnight at 4°C. A mold was made by cutting off the top 5 ml of a 50 ml conical tube and placed on dry ice. A solution of 1:2 OCT Compound:30% sucrose was mixed and centrifuged to remove bubbles. The mold was partially filled before adding the dissected eye and then covering fully. 10 µm frozen sections were cut at ∼-24°C with the front of the eye facing side on to the blade. Explanted retinae were fixed for 15 min in 4% PFA, washed with 1x PBS, incubated in 30% sucrose for 15 min, after which the membrane was cut from its frame and transferred to a mold containing 1:2 OCT:30% sucrose and frozen.

Immunohistochemical staining

For SYK and ARR3 co-staining, tissue sections were warmed at RT until dry and then washed in Coplin jars twice with 1x PBS for 5 min at RT. Samples were permeabilized in 1x PBS-T for 5 min, washed again in 1x PBS twice for 5 min, blocked for 1 hr in blocking solution [1x PBS, 0.1% Triton X-100, 5% normal donkey serum, 5% normal horse serum]. Primary mouse anti-SYK and rabbit anti-ARR3 (LUMIF-hCAR) antibodies were mixed in blocking solution and applied to samples overnight (16-18 h) at 4°C. Slides were washed three times with 1x PBS for 10 min each, probed with secondary antibodies (Key Resources Table) in blocking solution for 1 hr at RT, washed twice 1x PBS for 15 min each, incubated in1x PBS containing 10 µg/ml DAPI for 10 min, and mounted with Mowiol solution. Additional immunohistological staining of tissue sections and cells on coverslips was performed as described11 but without initial EDTA wash.

Combined in situ hybridization and immunohistological staining

Human retinal sections were prepared as above but using RNase-free reagents for all stages (PBS, TBS) and ultrapure sucrose (JT Baker, 4097-04) and cryostats cleaned with ELIMINase (Decon Labs, Inc) and 70% ethanol prior to use. RNA FISH hybridization chain reaction probes were designed by and obtained from Molecular Instruments, Inc, with 20 20-bp single strand DNA probes per target RNA sequence except Tr-NRL exon 1 (4 probes), FL-NRL exons 1 and (8 probes), and HOTAIRM1 (14 probes). The accession numbers for target sequences are shown in Table S11, except for RP13-143G15.4 which included regions of several isoforms for maximum gene coverage. The in situ protocol was performed with fluorescent hairpins and buffers from Molecular Instruments using the manufacturer’s instructions for mammalian cells on a slide (https://files.molecularinstruments.com/MI-Protocol-RNAFISH-MammalianCellSlide-Rev7.pdf), beginning with a 4% PFA treatment after thawing and drying slides at RT for 10 min, with the following changes: samples were incubated in 70% ethanol overnight in a Coplin jar, probes were added to the final hybridization volume at 4x the recommended concentration for tissue sections (2.4 pmol in 150 µl of hybridization buffer), tissues were pre-hybridized at 42°C and then lowered to 37°C once probes were added, and hairpins were used at a volume of 1 µl (6 pmol) in 50 µl amplification buffer per section. After removing amplification solution and washing, samples were directly used in the immunofluorescence protocol. Briefly, samples were blocked in a basic RNase-free solution (1% BSA, 0.1% Triton X-100, 0.05% Tween-20 in PBS) for 1 hr and primary antibodies mixed in blocking solution and applied to samples overnight at 4°C. Samples were washed three times for 10 min with 1x PBS, then incubated with secondary antibodies in blocking solution for 40 min at RT. Samples were washed as above with DAPI in the third and final wash, and mounted with Mowiol before imaging using a Leica STELLARIS 5 inverted confocal microscope. For GNAT2 and NR2E3 FISH, samples were washed with DAPI in the third and final wash, mounted and then imaged as above. After this first imaging, coverslips were removed, and slides were washed in PBS and then used in the immunofluorescence protocol with an alternate blocking solution) (2.5% Horse, donkey, human sera, 1% BSA, 0.1% Triton X-100, 0.05% Tween-20 in TBS).

Quantitation of FISH

For NRL, CHRNA1, and lncRNAs, QuPath81 was used to identify nuclei, predict cell expansion, classify cells, and count FISH puncta. Apical retina regions containing RXRγ+ and/or NRL+ cells were visually selected for evaluation. The StarDist extension82 was used to outline nuclei and cell expansion zones with deep learning model ‘dsb2018_heavy_augment.pb’ on the DAPI image channel (threshold = 0.6, pixelSize = 0.15). Model results were reviewed and cells with partial, multiple, or overlapping nuclei removed from consideration. FISH puncta were thresholded then output, using the subcellular detection function (expected spot size = 0.5 µm2, min spot size = 0.2 µm2, max spot size = 0.7 µm2 (NRL isoforms) or 1 µm2 (CHRNA1 and lncRNAs) to capture the most puncta above background. Image positions were determined by tracing the retina apical edge with line segments in FIJI, identifying the closest point on the line to the midpoint of each image, then measuring the distance from the line starting position. For GNAT2 and NR2E3 FISH, images were analyzed in FIJI. The outer layer of cells was delimited manually based on the RXRγ+ cells and the number of RXRγ+ and/or GNAT2+ and/or NR2E3+ cells were manually counted, examining each color channel separately. For the inner layer, the number of NR2E3 RNA positive cells and NR2E3 protein positive cells were counted.

NRL isoform analyses

Proteasome inhibition of CHLA-VC-RB-31

106 CHLA-VC-RB-31 cells37 were cultured in 2 ml retina culture medium in a 24-well plate with or without 10 µM retinoic acid (Cayman Chemical 11017) and incubated at 37°C in 5% CO2. MG132 (Sigma-Aldrich 474790) was added at 23 h to a final concentration of 10 µM and cells incubated for 4 h. Cells were collected and centrifuged at 300 x g for 4 min, resuspended in 60 µl cold 1x RIPA buffer (Millipore 20-188), 10% sodium dodecyl sulfate (SDS), protease and phosphatase inhibitors diluted per manufacturer instructions (Sigma-Aldrich 5892970001, 4906837001)], incubated on ice for 30 min, then centrifuged at 4°C for 20 min at 14,000 RPM. 15 µl of supernatant were mixed with 3 µl 6X sample buffer (300 mM Tris, 60% glycerol,12% SDS, 86 mM B-mercaptoethanol, 0.6 mg/ml bromophenol blue), heated to 95°C for 5 min, and separated on a 4-12% Bis-Tris gel (Invitrogen, NP0321) using 1xMOPS running buffer (Life Technologies NP0001). Protein was wet transferred to a PVDF membrane in Towbin transfer buffer (25 mM Tris, 192 mM glycine pH 8.3,10% methanol) at 20 V ON at RT. The membrane was blocked in 1xTBS-T containing 5% w/v nonfat dry milk for 1 hr, washed with TBS-T three times for 5 min, incubated overnight in 0.1% dry milk TBS-T solution containing primary antibody at 4°C with slow rocking, washed three times for 15 min in TBS-T, incubated with HRP-conjugated secondary antibody for 1 hr at RT in the 0.1% milk solution, washed six times for 15 min, incubated in chemiluminescent substrate (Thermo Fisher, 34094) and imaged.

NRL isoform and reporter constructs

The pcDNA4-His-Max-C-Nrl83 (gift of A. Swaroop) CMV promoter was replaced with EF1α promoter to produce pcDNA4-His-Max-C-EF1α-FL-NRL. Using In-Fusion (Takara Bio) the His-Max tag was removed to produce full-length NRL expression plasmid pCDNA4-C-EF1α-FL-NRL (primers #1,2). The His-Max tag plus first 417 bp of NRL N-terminal coding sequence were removed to produce truncated NRL expression plasmid pCDNA4-C-EF1α-Tr-NRL (primers #1,2,3,4). pcDNA4-C-EF1α-empty transfection carrier plasmid was made by removing NRL coding sequence from pCDNA4-C-EF1α-Fl-NRL (primers #5,6). pGL3-PDE6B-146-luc was produced by amplifying a 146 bp promoter region of PDE6B84 from H9 hESC genomic DNA and inserting into a pGL3-SV40 (Promega, primers #7,8), then removing the SV40 promoter (primers #9,10) to generate pGL3-PDE6B-146. The promoter-less pGL3-empty was produced by removing SV40 sequences from pGL3-SV40 (primers #9,11). pUltra-EGFP-P2A-Tr-NRL was produced by amplification of NRL coding sequence from pcDNA4-C-EF1α-FL-NRL and inserting into BamHI digested pUltra-EGFP (Addgene 24129) (primers #12, 13). All primers are listed in the Key Resources Table.

3NRL expression and analysis

pcDNA4-C-EF1α-NRL constructs were transfected into HEK293T cells. 4 µg plasmid in 1 ml DMEM (Life Technologies, 10313039) was combined with 1 ml DMEM + 12 µg polyethylenimine (PEI MAX) (MW 40,000, Polysciences 24765-1) and combined with 106 trypsinized HEK293T cells, plated into two wells of a 6-well plate with an additional 1 ml of DMEM + 10% FBS, then cultured overnight in a 37°C incubator. Media was replaced with fresh 293T media the next day and after two additional days cells were collected, lysed, and protein quantitated via BCA assay (Thermo Scientific PI23227). 30 µg of each well was used for western blot as above, with membranes probed with antibody raised against the NRL N-terminus (Santa Cruz SC-374277) or against full-length NRL (R&D AF2945).

Luciferase assays

575 ng of total DNA (50 ng reporter, 100ng FL-NRL and/or 100-300 ng Tr-NRL expression vector, and the remainder pcDNA4-C-EF1α empty vector) in 5 µl DMEM was mixed with 2.3 µg PEI in another 5 µl DMEM (per condition, in triplicate) and then plated into one well of a white bottom 96-well cell culture plate with 50,000 NIH3T3 cells in 100 µl 293T media. The luciferase assay was carried out 48 h after transfection with the Nano-Glo Dual-Luciferase Reporter Assay System (Promega N1610) per manufacturer instructions. Luminescence was recorded using the Promega Glomax Multi+ Detection System.

Truncated NRL overexpression in fetal retina

Concentrated pUltra-EGFP-NRL-205 and control pUltra-EGFP lentivirus were produced as described85. At 16-24 h after 293T transfection in 15 cm dishes, media was exchanged for 20 ml UltraCULTURE (Lonza Inc., BP12-725F) containing 1% HEPES (Sigma, 0887), 1% GlutaMax (Life Technologies 35050061), and 1% Penicillin-Streptomycin (Fisher Scientific, MT30002CI). After 60-64 h, media was harvested, centrifuged at 3000 x g for 10 min at 4°C, filtered through 0.45 µm PVDF flask filter, concentrated via tangential flow filtration using a MidiKros 20 cm 500 KD 0.5 mm column (D02-5500-05-S), with both input and final supernatant containers kept in ice. The concentrate was re-filtered using a 0.45 µm PVDF syringe filter (Millipore), stored in aliquots at -80°C, and titered using a p24 ELISA kit (ZeptoMetrix 801002).

As described11, fetal eyes were washed in 70% ethanol, washed three times in 1x sterile PBS, and dissected in cold 1x PBS. The cornea was removed and tissue cut ∼one-third of the distance towards the posterior pole along 4 lines approximately equidistant and between attached tendons. The retina was removed in this flattened state after cutting the optic nerve, transferred to a polytetrafluoroethylene culture insert (Millipore, PICM0RG50) with photoreceptor side down, and placed in a 6-well plate with 1,200 µl retinal culture medium. Retinae were infected as described, cultured for 7 days with half media changes every other day, then fixed in 4% PFA for 15 min, washed three times in 1x PBS, equilibrated in sucrose (30% in PBS) for 15 min, and embedded and sectioned as above. Immunostaining and imaging were performed as above.

RB1 knockdown and SYK inhibitor treatment

FW18.5 retina was partially dissociated with papain, cultured overnight in retinal culture media at 37°C in a 6-well plate, then frozen in 10% DMSO solution and stored in liquid nitrogen. Samples were thawed and revived in retinal culture media overnight and infected with lentivirus carrying shRNAs targeting RB1 (shRB1-733)10 or a scrambled control (shSCR) and diluted β-mercaptoethanol, insulin, glutamine, penicillin, and streptomycin to the final retina culture medium concentrations along with 5 µg/ml Polybrene, for 24 hrs. After replacing 2/3 of media in each well with fresh media, cells were treated with GS-9876 (MedChemExpress, HY-109091) in DMSO at or with the same volume of DMSO vehicle control. After 12 d, cells were attached to poly-L-Lysine-coated coverslips for 3 h, fixed in 4% PFA for 10 min, washed in 1x PBS three times, and stored at -20°C until immunocytological staining with antibodies to Ki67, RXRγ, and GFP (for YFP staining) as described10.

Statistical Analysis

Statistical methods and packages are cited in the Methods text and figure legends. Statistical comparison between more than two groups of cells for gene expression, regulon activity, or exon count proportions used the non-parametric Kruskal-Wallis test followed by pairwise post-hoc Dunn tests with Benjamini-Hochberg correction. Mean expression levels of NRL and RXRG isoforms across cell clusters were evaluated by Welch’s t-tests for groups with unequal variance; p-values for these tests were estimated using bootstrapping with 1,000,000 replications per comparison. Changes in FISH puncta counts were examined within each channel individually using zero-inflated negative binomial regression models with robust standard error terms. Distance across the sample tissue was used to model degenerate zero inflation. P values reported in the manuscript text and legends are Wald Chi-Squared tests that evaluate differences in count between inflection points. Analyses were performed using Microsoft Excel, the R statistical language in RStudio, or Stata SE v14.2.